---
title: "00.00-E-Peve-RNAseq-reads-FastQC-MultiQC.Rmd"
author: "Sam White"
date: "2024-02-20"
output: 
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

This Rmd file will download raw RNA-seq FastQs for *P.evermanni* and evaluate them using [FastQC](https://github.com/s-andrews/FastQC) and [MultiQC](https://multiqc.info/)[@ewels2016].

# Create a Bash variables file

This allows usage of Bash variables across R Markdown chunks.

```{r save-bash-variables-to-rvars-file, engine='bash', eval=TRUE}
{
echo "#### Assign Variables ####"
echo ""

echo "# Data directories"
echo 'export timeseries_dir=/home/shared/8TB_HDD_01/sam/gitrepos/urol-e5/timeseries_molecular'
echo 'export output_dir_top=${timeseries_dir}/E-Peve/output/00.00-E-Peve-RNAseq-reads-FastQC-MultiQC'
echo 'export raw_reads_dir=${timeseries_dir}/E-Peve/data/rnaseq-raw-fastqs'
echo 'export raw_reads_url="https://owl.fish.washington.edu/nightingales/E5-coral-time-series/30-1047560508/"'
echo ""

echo "# Input files"
echo 'export metadata_file="${timeseries_dir}/M-multi-species/data/rna_metadata.csv"'

echo "# Paths to programs"
echo 'export fastqc=/home/shared/FastQC-0.12.1/fastqc'
echo 'export multiqc=/home/sam/programs/mambaforge/bin/multiqc'
echo ""

echo "# Set FastQ filename patterns"
echo "export fastq_pattern='*.fastq.gz'"
echo "export R1_fastq_pattern='*_R1_*.fastq.gz'"
echo "export R2_fastq_pattern='*_R2_*.fastq.gz'"
echo ""

echo "# Set number of CPUs to use"
echo 'export threads=40'
echo ""


echo "## Inititalize arrays"
echo 'export fastq_array_R1=()'
echo 'export fastq_array_R2=()'
echo 'export raw_fastqs_array=()'
echo 'export R1_names_array=()'
echo 'export R2_names_array=()'
echo ""

echo "# Programs associative array"
echo "declare -A programs_array"
echo "programs_array=("
echo '[fastqc]="${fastqc}" \'
echo '[multiqc]="${multiqc}" \'
echo ")"
echo ""

echo "# Print formatting"
echo 'export line="--------------------------------------------------------"'
echo ""
} > .bashvars

cat .bashvars
```

# Download *P.evermanni* RNA-seq FastQs

## Inspect metadata file

```{r, inspect-metadata, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

head ${metadata_file} |
column -t -s","
```

## Download raw RNA-seq reads

Reads are downloaded from <https://owl.fish.washington.edu/nightingales/E5-coral-time-series/30-1047560508/>

Since sequencing included multiple species, the code will also parse only those that are *P.evermanni*.

The `--cut-dirs 3` command cuts the preceding directory structure (i.e. `nightingales/E5-coral-time-series/30-1047560508/`) so that we just end up with the reads.

```{bash download-raw-reads, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

# Make output directory if it doesn't exist
mkdir --parents ${raw_reads_dir}

# Create list of only P.evermanni sample names
# Extract the AzentaSampleName for P.evermanni samples
sample_list=$(awk -F"," '$9 == "Porites evermanni" { print $5 }' ${timeseries_dir}/M-multi-species/data/rna_metadata.csv | sort)

echo ""
echo "${line}"
echo ""
echo "Sample list:"
echo ""
echo "${sample_list}"
echo ""
echo "${line}"
echo ""

# Use printf to format each item for use in wget
# Add a wildcard to ignore the <number>-- prefix
formatted_list=$(printf "*--%s_*," ${sample_list})

# Remove the trailing comma
formatted_list="${formatted_list%,}"

# Output the final wget command
echo ""
echo "${line}"
echo ""
echo "Formatted wget accept list:"
echo ""
echo "wget --accept=\"$formatted_list\""
echo ""
echo "${line}"
echo ""

# Run wget to retrieve FastQs and MD5 files
# Note: the --no-clobber command will skip re-downloading any files that are already present in the output directory
wget \
--directory-prefix ${raw_reads_dir} \
--recursive \
--no-check-certificate \
--continue \
--cut-dirs 3 \
--no-host-directories \
--no-parent \
--quiet \
--no-clobber \
--accept=${formatted_list} ${raw_reads_url}

ls -lh "${raw_reads_dir}"
```


# Rename FastQs


```{bash rename-fastqs, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

cd "${raw_reads_dir}"

# Create an associative array to store the mapping
declare -A azenta_to_colony_timepoint

# Create a temporary file to store the mapping
temp_mapping_file=$(mktemp)

# Read the metadata file and populate the temporary file using awk
# Maps Azenta sample name to colony_id and timepoint.
awk -F, 'NR > 1 { print $5, $6 "-" $7 }' "$metadata_file" > "$temp_mapping_file"

# Read the temporary file and populate the associative array
while read -r azenta_sample_name colony_timepoint; do
    azenta_to_colony_timepoint["$azenta_sample_name"]="$colony_timepoint"
done < "$temp_mapping_file"

# Remove the temporary file
rm "$temp_mapping_file"

# Print the associative array for debugging
echo "Azenta sample name to colony_id-Timepoint mapping:"
for key in "${!azenta_to_colony_timepoint[@]}"; do
    echo ""
    echo "$key -> ${azenta_to_colony_timepoint[$key]}"
    echo ""
done

# Iterate over the FastQ files in the current directory
for fastq_file in *.fastq.gz *.md5; do
    # Extract the Azenta sample name from the filename
    azenta_sample_name=$(echo "$fastq_file" | sed -E 's/^[0-9]+--([A-Za-z0-9]+)_.*/\1/')
    
    # Check if the Azenta sample name exists in the associative array
    if [[ -n "${azenta_to_colony_timepoint[$azenta_sample_name]}" ]]; then
        new_sample_name="${azenta_to_colony_timepoint[$azenta_sample_name]}"
        new_filename=$(echo "$fastq_file" | sed -E "s/^[0-9]+--$azenta_sample_name/$new_sample_name/")
        if [[ "$fastq_file" != "$new_filename" ]]; then
            mv "$fastq_file" "$new_filename"
            echo "Renamed $fastq_file to $new_filename"
        else
            echo "No renaming needed for $fastq_file"
        fi

        # If the file is a .md5 file, update its contents
        if [[ "$fastq_file" == *.md5 ]]; then
            sed -i -E "s|(\./)[0-9]+--$azenta_sample_name|\1$new_sample_name|" "$new_filename"
            echo "Updated contents of $new_filename"
            echo ""
        fi
    else
        echo "Azenta sample name '$azenta_sample_name' not found in metadata."
        # Debugging: Print all keys in the associative array with quotes
        echo "Available keys in the associative array:"
        for key in "${!azenta_to_colony_timepoint[@]}"; do
            echo "'$key'"
        done
    fi
done
```

## Verify raw read checksums

```{bash verify-raw-read-checksums, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

cd "${raw_reads_dir}"

# Checksums file contains other files, so this just looks for the RNAseq files.
for file in *.md5
do
  md5sum --check "${file}"
done
```

# FastQC/MultiQC on raw reads

```{bash raw-fastqc-multiqc, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

# Make output directory if it doesn't exist
mkdir --parents "${raw_reads_dir}"

############ RUN FASTQC ############


# Create array of trimmed FastQs
raw_fastqs_array=(${raw_reads_dir}/${fastq_pattern})

# Pass array contents to new variable as space-delimited list
raw_fastqc_list=$(echo "${raw_fastqs_array[*]}")

echo "Beginning FastQC on raw reads..."
echo ""

# Run FastQC
### NOTE: Do NOT quote raw_fastqc_list
${programs_array[fastqc]} \
--threads ${threads} \
--outdir ${raw_reads_dir} \
--quiet \
${raw_fastqc_list}

echo "FastQC on raw reads complete!"
echo ""

############ END FASTQC ############

############ RUN MULTIQC ############
echo "Beginning MultiQC on raw FastQC..."
echo ""

${programs_array[multiqc]} ${raw_reads_dir} -o ${raw_reads_dir}

echo ""
echo "MultiQC on raw FastQs complete."
echo ""

############ END MULTIQC ############

echo "Removing FastQC zip files."
echo ""
rm ${raw_reads_dir}/*.zip
echo "FastQC zip files removed."
echo ""
```

```{bash list-fastqc, eval=TRUE}
# Load bash variables into memory
source .bashvars

# View directory contents
ls -lh ${raw_reads_dir}

```


