---
title: "13.00-multiomics-stdm-02"
author: "Sam White"
date: "2025-10-15"
output: 
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
citeproc: true
---
# BACKGROUND

This notebook performs tensor decomposition analysis on multi-species gene expression timeseries data using Steven's [`workflow-stdm-02`](https://github.com/sr320/workflow-stdm-02) (GitHub repo) tensor decomposition project.

# SETUP

## Libraries

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(reticulate)

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
    eval = FALSE,        # Evaluate code chunks
    results = "hold",   # Hold outputs and show them after the full code chunk
  warning = FALSE,     # Hide warnings
        collapse = FALSE,    # Keep code and output in separate blocks
        warning = FALSE,     # Hide warnings
        message = FALSE,     # Hide messages
        comment = "##"      # Prefix output lines with '##' so output is visually distinct
)
```

## Set R variables

```{r R-variables, eval=TRUE}
# OUTPUT DIRECTORY
output_dir <- "../output/13.00-multiomics-stdm"

#INPUT FILE(S)

# PYTHON ENVIRONMENT - Use the project's .venv environment
python_path <- "/home/shared/16TB_HDD_01/sam/gitrepos/urol-e5/timeseries_molecular/.venv/bin/python"
```

## Load project Python environment

The project uses a virtual environment (`.venv`) with all required packages. If this is successful, the Python path should show the `.venv` environment.

```{r load-python-env, eval=TRUE}
# Use the project's .venv Python environment directly
library(reticulate)

# Set Python path to the project's .venv
use_python(python_path, required = TRUE)

# Show final configuration
py_config()
```