---
title: "13.00-multiomics-barnacle"
author: "Sam White"
date: "2025-10-03"
output: 
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
citeproc: true
---

# BACKGROUND

This analysis prepares three-species ortholog expression matrices, normalizes the data, constructs a multi‑omics tensor (genes × combined species-samples × timepoints), and runs a sparse tensor decomposition using the [barnacle](https://github.com/blasks/barnacle) [@blaskowski2024] workflow to discover shared gene/time/species patterns.

## Workflow overview

1. **Data Preparation** - Extract and normalize ortholog expression data
2. **Tensor Construction** - Build 3D tensor combining species, samples, and timepoints
3. **Rank Comparison** - Test multiple ranks systematically to find optimal components
4. **Interpretation** - Choose best rank and examine factors, visualizations, and biological meaning

## Rank selection approach

**This analysis uses a systematic rank comparison approach.**

The workflow:
1. Configure `ranks_to_test` in the comparison chunk (e.g., `[5, 8, 10, 12, 15, 20]`)
2. Run decompositions for all specified ranks (10-30 minutes)
3. Review comparison metrics and elbow plot
4. Choose optimal rank based on:
   - Variance explained (aim for >60%)
   - Diminishing returns in elbow plot
   - Sparsity patterns
   - Component interpretability
5. Use chosen rank for final biological interpretation

**Key metrics for comparison:**
- **Variance explained**: How much of data variation is captured
- **Relative error**: Reconstruction accuracy
- **Sparsity**: How many genes/samples load strongly per component
- **Convergence**: Whether optimization completed successfully
- **Component weights**: Distribution of importance across components

## Key steps performed by this script:

- Load annotated ortholog groups and per-species count matrices.
- Filter ortholog groups to retain complete three‑way matches with available expression data.
- Extract per‑species expression matrices mapped to ortholog group IDs.
- Normalize counts (preferred: `sctransform`; fallback: log2(CPM + 1)).
- Build a 3D tensor combining species/sample and timepoint dimensions.
- Run [barnacle](https://github.com/blasks/barnacle) [@blaskowski2024] sparse CP decomposition and save factor matrices, metadata and figures.
- Assess rank appropriateness through multiple metrics and visualizations.

## Input files

- `../output/12-ortho-annot/ortholog_groups_annotated.csv` : annotated ortholog groups with columns for `group_id`, `apul`, `peve`, `ptua`, `type`, etc.
- `../../D-Apul/output/02.20-D-Apul-RNAseq-alignment-HiSat2/apul-gene_count_matrix.csv` : Apul gene count matrix (contains `gene_id` plus sample columns).
- `../../E-Peve/output/02.20-E-Peve-RNAseq-alignment-HiSat2/peve-gene_count_matrix.csv` : Peve gene count matrix.
- `../../F-Ptua/output/02.20-F-Ptua-RNAseq-alignment-HiSat2/ptua-gene_count_matrix.csv` : Ptua gene count matrix.
- (Also referenced) transcript-level matrices: `apul-transcript_count_matrix.csv`, `peve-transcript_count_matrix.csv`, `ptua-transcript_count_matrix.csv` when needed.

## Output files

- `apul_ortholog_expression.csv`, `peve_ortholog_expression.csv`, `ptua_ortholog_expression.csv` : per‑species expression matrices aligned to ortholog `group_id`.
- `apul_normalized_expression.csv`, `peve_normalized_expression.csv`, `ptua_normalized_expression.csv` : normalized expression matrices (sctransform or log2(CPM+1)).
- `multiomics_tensor.npy` : saved NumPy array of the 3D tensor used for decomposition (genes × combined_samples × timepoints).
- `barnacle_factors/` directory containing:
    - `gene_factors.csv` : gene loadings per component (genes × components).
    - `sample_factors.csv` : combined sample (species_sample) loadings per component with `Species` and `Sample_ID` metadata.
    - `time_factors.csv` : timepoint loadings per component.
    - `component_weights.csv` : component weights / importance.
    - `sample_mapping.csv` : mapping of combined sample indices to species and sample IDs.
    - `metadata.json` : analysis parameters and tensor metadata (shape, rank, lambdas, convergence, etc.).
    - `figures/` : generated visualizations (component weights, time loadings, sample heatmap, PCA, top ortholog plots).

## Notes / assumptions:

- Sample column names are parsed expecting a dot-separated format with a `TP#` timepoint token (e.g., `ACR.139.TP1`).
- Apul ortholog IDs in the ortholog table include transcript suffixes (e.g., `-T1`) which are removed for matching to the gene count matrix.
- Missing values in the tensor are handled by substitution (current workflow fills NaNs with zeros before decomposition).
- Samples must have all four timepoints (TP1, TP2, TP3, TP4) to be included in the analysis.


# SETUP

## Libraries

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(reticulate)
library(matrixStats)  # Required for sctransform internal functions
library(sctransform)
library(glmGamPoi)

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
    eval = FALSE,        # Evaluate code chunks
    results = "hold",   # Hold outputs and show them after the full code chunk
  warning = FALSE,     # Hide warnings
        collapse = FALSE,    # Keep code and output in separate blocks
        warning = FALSE,     # Hide warnings
        message = FALSE,     # Hide messages
        comment = "##"      # Prefix output lines with '##' so output is visually distinct
)
```

## Set R variables

```{r R-variables, eval=TRUE}
# OUTPUT DIRECTORY
output_dir <- "../output/13.00-multiomics-barnacle"

#INPUT FILE(S)
ortholog_groups_file <- "../output/12-ortho-annot/ortholog_groups_annotated.csv"

# Transcript count matrices for each species
apul_transcript_matrix_file <- "../../D-Apul/output/02.20-D-Apul-RNAseq-alignment-HiSat2/apul-transcript_count_matrix.csv"
peve_transcript_matrix_file <- "../../E-Peve/output/02.20-E-Peve-RNAseq-alignment-HiSat2/peve-transcript_count_matrix.csv"
ptua_transcript_matrix_file <- "../../F-Ptua/output/02.20-F-Ptua-RNAseq-alignment-HiSat2/ptua-transcript_count_matrix.csv"

# CONDA
conda_env_name <- c("/home/sam/programs/mambaforge/envs/barnacle_py311_env")
conda_path <- c("/opt/anaconda/anaconda3/bin/conda")
```

## Load [barnacle](https://github.com/blasks/barnacle) conda environment

If this is successful, the first line of output should show that the Python being used is the one in your [barnacle](https://github.com/blasks/barnacle) [@blaskowski2024] conda environment path.

E.g.

`python:         /home/sam/programs/mambaforge/envs/barnacle_py311_env/bin/python`

```{r load-barnacle-conda-env, eval=TRUE}
use_condaenv(condaenv = conda_env_name, conda = conda_path)
py_config()
```

# DATA PREP

## Load ortholog groups data

```{r load-ortholog-data, eval=TRUE}
# Read in the ortholog groups data
ortholog_groups <- read.csv(ortholog_groups_file)



# Display basic info about the data
cat("Dimensions of ortholog groups data:", dim(ortholog_groups), "\n\n")
cat("Column names:", colnames(ortholog_groups), "\n\n")
head(ortholog_groups)
str(ortholog_groups)
```

## Extract ortholog expression data

Now let's extract expression data for genes that are present in the ortholog groups. We'll use the gene count matrices with gene IDs to properly map the data.

### Load gene count matrices

```{r load-gene-matrices, eval=TRUE}
# Define file paths for gene count matrices
apul_gene_matrix_file <- "../../D-Apul/output/02.20-D-Apul-RNAseq-alignment-HiSat2/apul-gene_count_matrix.csv"
peve_gene_matrix_file <- "../../E-Peve/output/02.20-E-Peve-RNAseq-alignment-HiSat2/peve-gene_count_matrix.csv"
ptua_gene_matrix_file <- "../../F-Ptua/output/02.20-F-Ptua-RNAseq-alignment-HiSat2/ptua-gene_count_matrix.csv"

# Load gene count matrices for each species
cat("Loading gene count matrices...\n\n")

apul_gene_matrix <- read.csv(apul_gene_matrix_file)
cat("Apul gene matrix dimensions:", dim(apul_gene_matrix), "\n")

peve_gene_matrix <- read.csv(peve_gene_matrix_file)
cat("Peve gene matrix dimensions:", dim(peve_gene_matrix), "\n")

ptua_gene_matrix <- read.csv(ptua_gene_matrix_file)
cat("Ptua gene matrix dimensions:", dim(ptua_gene_matrix), "\n\n")
```

### Filter ortholog groups for complete three-way matches

```{r filter-ortholog-groups, eval=TRUE}
cat("Filtering for complete three-way ortholog groups...\n")

# Keep only rows where all three species have entries (no NA values or empty strings)
complete_ortholog_groups <- ortholog_groups[nzchar(ortholog_groups$apul) & 
                                          nzchar(ortholog_groups$peve) & 
                                          nzchar(ortholog_groups$ptua), ]

cat("Total ortholog groups:", nrow(ortholog_groups), "\n")
cat("Complete three-way ortholog groups:", nrow(complete_ortholog_groups), "\n")

```

### Filter for expression data availability

```{r filter-expression-availability, eval=TRUE}
cat("Filtering ortholog groups to ensure all genes have expression data...\n")

# Clean gene IDs to check against expression data
# For Apul: remove -T[n] suffix from ortholog groups to match gene matrix format
apul_ortholog_genes_check <- gsub("-T[0-9]+$", "", complete_ortholog_groups$apul)

# For Peve and Ptua: use as-is (will clean gene matrix IDs later)
peve_ortholog_genes_check <- complete_ortholog_groups$peve
ptua_ortholog_genes_check <- complete_ortholog_groups$ptua

# Check which genes are present in expression data
# (Note: We need to clean gene matrix IDs to match)
apul_gene_matrix_ids <- gsub("^gene-", "", apul_gene_matrix$gene_id)  # Remove gene- prefix if present
peve_gene_matrix_ids <- gsub("^gene-", "", peve_gene_matrix$gene_id)  # Remove gene- prefix
ptua_gene_matrix_ids <- gsub("^gene-", "", ptua_gene_matrix$gene_id)  # Remove gene- prefix

# Find which ortholog genes are present in each species' expression data
apul_present <- apul_ortholog_genes_check %in% apul_gene_matrix_ids
peve_present <- peve_ortholog_genes_check %in% peve_gene_matrix_ids
ptua_present <- ptua_ortholog_genes_check %in% ptua_gene_matrix_ids

# Keep only ortholog groups where all three species have expression data
expression_complete_mask <- apul_present & peve_present & ptua_present
complete_ortholog_groups <- complete_ortholog_groups[expression_complete_mask, ]

cat("Ortholog groups after filtering for expression data availability:", nrow(complete_ortholog_groups), "\n")

```

### Gene ID cleaning examples

```{r gene-id-examples-testing, eval=TRUE}
cat("\n=== GENE ID CLEANING EXAMPLES ===\n")
cat("Apul (clean ortholog groups to match gene matrix):\n")
cat("Ortholog groups original:", head(complete_ortholog_groups$apul, 3), "\n")
cat("Ortholog groups cleaned:", head(gsub("-T[0-9]+$", "", complete_ortholog_groups$apul), 3), "\n")
cat("Gene matrix (target format):", head(apul_gene_matrix$gene_id, 3), "\n\n")

cat("Peve (clean gene matrix to match ortholog groups):\n")
cat("Ortholog groups (target format):", head(complete_ortholog_groups$peve, 3), "\n") 
cat("Gene matrix original:", head(peve_gene_matrix$gene_id, 3), "\n")
cat("Gene matrix cleaned:", head(peve_gene_matrix$gene_id_clean, 3), "\n\n")

cat("Ptua (clean gene matrix to match ortholog groups):\n")
cat("Ortholog groups (target format):", head(complete_ortholog_groups$ptua, 3), "\n")
cat("Gene matrix original:", head(ptua_gene_matrix$gene_id, 3), "\n")
cat("Gene matrix cleaned:", head(ptua_gene_matrix$gene_id_clean, 3), "\n\n")

```

### Clean gene IDs for matching

```{r clean-gene-ids, eval=TRUE}
cat("Cleaning gene matrix IDs to match ortholog group format...\n")

# For Apul: ortholog groups have "FUN_000185-T1", gene matrix has "FUN_002326"
# We need to remove "-T1" from ortholog groups to match gene matrix
apul_ortholog_genes <- unique(gsub("-T[0-9]+$", "", complete_ortholog_groups$apul))

# For Peve and Ptua: keep ortholog groups as-is and clean gene matrix
peve_ortholog_genes <- unique(complete_ortholog_groups$peve)
ptua_ortholog_genes <- unique(complete_ortholog_groups$ptua)

# Clean gene matrix IDs accordingly
# Apul: gene matrix already in correct format (no cleaning needed)
apul_gene_matrix$gene_id_clean <- apul_gene_matrix$gene_id

# Peve: gene matrix has "gene-Peve_00000032", ortholog groups have "Peve_00037402"  
# So we need to remove "gene-" prefix from gene matrix
peve_gene_matrix$gene_id_clean <- gsub("^gene-", "", peve_gene_matrix$gene_id)

# Ptua: gene matrix has "gene-Pocillopora_meandrina_HIv1___RNAseq.g20905.t1", 
# ortholog groups have "Pocillopora_meandrina_HIv1___RNAseq.g28886.t1"
# So we just need to remove "gene-" prefix from gene matrix
ptua_gene_matrix$gene_id_clean <- gsub("^gene-", "", ptua_gene_matrix$gene_id)

cat("Apul ortholog genes (complete groups only):", length(apul_ortholog_genes), "\n")
cat("Peve ortholog genes (complete groups only):", length(peve_ortholog_genes), "\n")
cat("Ptua ortholog genes (complete groups only):", length(ptua_ortholog_genes), "\n\n")
```

### Define sample filtering function

```{r define-sample-filter-function, eval=TRUE}
# Function to filter samples that have all four timepoints (TP1, TP2, TP3, TP4)
filter_complete_samples <- function(gene_matrix, species_name) {
  cat("  Filtering samples with complete timepoints for", species_name, "...\n")
  
  # Get expression columns (exclude gene_id column, gene_id_clean may not exist yet)
  all_cols <- colnames(gene_matrix)
  id_cols <- c("gene_id", "gene_id_clean")
  expr_cols <- setdiff(all_cols, id_cols)
  
  if(length(expr_cols) == 0) {
    cat("  ERROR: No expression columns found! Returning original matrix.\n")
    return(gene_matrix)
  }
  
  # Parse sample column names to identify sample groups and timepoints
  sample_timepoint_map <- list()
  
  for(col in expr_cols) {
    # Expected formats: ACR.139.TP1, POR.216.TP1, POC.201.TP1, etc.
    # Note: Using dots (.) as separators, not hyphens (-)
    parts <- strsplit(col, "\\.")[[1]]  # Split on dots, not hyphens
    
    if(length(parts) >= 3) {
      # Extract timepoint from last part (e.g., "TP1", "TP2", etc.)
      tp_part <- parts[length(parts)]
      
      if(grepl("^TP[0-9]+$", tp_part)) {
        timepoint <- as.numeric(gsub("TP", "", tp_part))
        
        # Sample ID is everything except the timepoint part, joined with dots
        sample_id <- paste(parts[1:(length(parts)-1)], collapse = ".")
        
        if(!sample_id %in% names(sample_timepoint_map)) {
          sample_timepoint_map[[sample_id]] <- list()
        }
        sample_timepoint_map[[sample_id]][[as.character(timepoint)]] <- col
      }
    }
  }
  
  # Identify samples with all four timepoints (1, 2, 3, 4)
  complete_samples <- c()
  incomplete_samples <- c()
  required_timepoints <- c("1", "2", "3", "4")
  
  for(sample_id in names(sample_timepoint_map)) {
    available_timepoints <- names(sample_timepoint_map[[sample_id]])
    
    if(all(required_timepoints %in% available_timepoints)) {
      complete_samples <- c(complete_samples, sample_id)
    } else {
      incomplete_samples <- c(incomplete_samples, sample_id)
      missing_tps <- setdiff(required_timepoints, available_timepoints)
      cat("    Sample", sample_id, "missing timepoints:", paste(paste0("TP", missing_tps), collapse = ", "), "\n")
    }
  }
  
  cat("  Complete samples (all 4 timepoints):", length(complete_samples), "\n")
  cat("  Incomplete samples:", length(incomplete_samples), "\n")
  
  if(length(incomplete_samples) > 0) {
    cat("  Removing incomplete samples:", paste(incomplete_samples, collapse = ", "), "\n")
  }
  
  # Build list of columns to keep (include gene_id columns plus ONLY complete sample columns)
  if(length(complete_samples) == 0) {
    cat("  WARNING: No complete samples found! Keeping all expression columns.\n")
    # Fallback: keep all expression columns
    keep_cols <- c("gene_id")
    if("gene_id_clean" %in% colnames(gene_matrix)) {
      keep_cols <- c(keep_cols, "gene_id_clean")
    }
    keep_cols <- c(keep_cols, expr_cols)
  } else {
    keep_cols <- c("gene_id")
    if("gene_id_clean" %in% colnames(gene_matrix)) {
      keep_cols <- c(keep_cols, "gene_id_clean")
    }
    
    # ONLY add columns from complete samples (this excludes incomplete sample columns)
    for(sample_id in complete_samples) {
      for(tp in required_timepoints) {
        if(tp %in% names(sample_timepoint_map[[sample_id]])) {
          keep_cols <- c(keep_cols, sample_timepoint_map[[sample_id]][[tp]])
        }
      }
    }
    
    # Double-check: ensure we're not accidentally including incomplete sample columns
    incomplete_cols <- c()
    for(sample_id in incomplete_samples) {
      for(tp in names(sample_timepoint_map[[sample_id]])) {
        incomplete_cols <- c(incomplete_cols, sample_timepoint_map[[sample_id]][[tp]])
      }
    }
    
    if(length(incomplete_cols) > 0) {
      # Remove any incomplete sample columns that might have been included
      keep_cols <- setdiff(keep_cols, incomplete_cols)
    }
  }
  
  # Filter the gene matrix to keep only complete samples
  keep_cols <- intersect(keep_cols, colnames(gene_matrix))  # Ensure columns exist
  filtered_matrix <- gene_matrix[, keep_cols, drop = FALSE]
  
  # Count ID vs expression columns in final result
  final_expr_cols <- setdiff(colnames(filtered_matrix), c("gene_id", "gene_id_clean"))
  cat("  Final expression columns:", length(final_expr_cols), "\n")
  
  return(filtered_matrix)
}
```

### Define expression extraction function

```{r define-expression-extraction-function, eval=TRUE}
# Function to extract expression data for a species using ortholog group mapping
extract_species_expression <- function(ortholog_groups_df, gene_matrix, species_col, species_name) {
  cat("Processing", species_name, "...\n")
  
  # First, filter the gene matrix to keep only samples with complete timepoints
  filtered_gene_matrix <- filter_complete_samples(gene_matrix, species_name)
  
  # Remove duplicate group_ids, keeping first occurrence of each
  unique_groups <- ortholog_groups_df[!duplicated(ortholog_groups_df$group_id), ]
  cat("  Unique ortholog groups:", nrow(unique_groups), "\n")
  
  # Create results data frame starting with group_id
  result_df <- data.frame(group_id = unique_groups$group_id, stringsAsFactors = FALSE)
  
  # Get expression columns (exclude gene_id and gene_id_clean columns)
  expr_cols <- setdiff(colnames(filtered_gene_matrix), c("gene_id", "gene_id_clean"))
  
  # Initialize expression columns with NA
  for(col in expr_cols) {
    result_df[[col]] <- NA
  }
  
  # For each ortholog group, find the corresponding gene and extract expression
  for(i in seq_len(nrow(unique_groups))) {
    target_gene <- unique_groups[[species_col]][i]
    
    # Clean target gene for matching
    if(species_name == "Apul") {
      # Remove transcript suffix for Apul
      target_gene_clean <- gsub("-T[0-9]+$", "", target_gene)
      matching_rows <- which(filtered_gene_matrix$gene_id_clean == target_gene_clean)
    } else {
      # For Peve and Ptua, match cleaned gene_id
      matching_rows <- which(filtered_gene_matrix$gene_id_clean == target_gene)
    }
    
    if(length(matching_rows) == 1) {
      # Single match - copy expression data
      for(col in expr_cols) {
        result_df[i, col] <- filtered_gene_matrix[matching_rows, col]
      }
    } else if(length(matching_rows) > 1) {
      # Multiple matches - take first gene (no averaging)
      first_match <- matching_rows[1]
      for(col in expr_cols) {
        result_df[i, col] <- filtered_gene_matrix[first_match, col]
      }
    } else {
      # No match - leave as NA (will be removed later)
    }
  }
  
  # Remove rows with all NA expression values
  expr_na_mask <- apply(result_df[, expr_cols, drop = FALSE], 1, function(x) all(is.na(x)))
  result_df <- result_df[!expr_na_mask, ]
  
  cat("  Final dimensions:", nrow(result_df), "ortholog groups x", ncol(result_df)-1, "samples\n")
  cat("  Removed", sum(expr_na_mask), "groups with no expression data\n\n")
  
  return(result_df)
}
```

### Extract ortholog expression data

```{r extract-ortholog-expression-data, eval=TRUE}
cat("Creating ortholog expression data with proper group_id mapping...\n")

# Extract expression data for each species using the ortholog group mapping
apul_ortholog_expression <- extract_species_expression(complete_ortholog_groups, apul_gene_matrix, "apul", "Apul")
peve_ortholog_expression <- extract_species_expression(complete_ortholog_groups, peve_gene_matrix, "peve", "Peve")
ptua_ortholog_expression <- extract_species_expression(complete_ortholog_groups, ptua_gene_matrix, "ptua", "Ptua")

cat("=== FINAL ORTHOLOG EXPRESSION DATA DIMENSIONS ===\n")
cat("Apul:", nrow(apul_ortholog_expression), "ortholog groups x", ncol(apul_ortholog_expression)-1, "samples\n")
cat("Peve:", nrow(peve_ortholog_expression), "ortholog groups x", ncol(peve_ortholog_expression)-1, "samples\n")
cat("Ptua:", nrow(ptua_ortholog_expression), "ortholog groups x", ncol(ptua_ortholog_expression)-1, "samples\n\n")
```

### Write ortholog expression data

```{r write-expression-data, eval=TRUE}
cat("Exporting ortholog expression data to CSV files...\n")

# Define output file paths
apul_output_file <- file.path(output_dir, "apul_ortholog_expression.csv")
peve_output_file <- file.path(output_dir, "peve_ortholog_expression.csv")
ptua_output_file <- file.path(output_dir, "ptua_ortholog_expression.csv")

# Write CSV files without quotes
write.csv(apul_ortholog_expression, file = apul_output_file, quote = FALSE, row.names = FALSE)
cat("Exported Apul ortholog expression data to:", apul_output_file, "\n")

write.csv(peve_ortholog_expression, file = peve_output_file, quote = FALSE, row.names = FALSE)
cat("Exported Peve ortholog expression data to:", peve_output_file, "\n")

write.csv(ptua_ortholog_expression, file = ptua_output_file, quote = FALSE, row.names = FALSE)
cat("Exported Ptua ortholog expression data to:", ptua_output_file, "\n")

cat("\nAll ortholog expression data exported successfully!\n\n")
```

### Column structure analysis

```{r column-structure-analysis, eval=TRUE}
cat("=== COLUMN STRUCTURE ANALYSIS ===\n")
cat("Apul columns:", ncol(apul_ortholog_expression), "\n")
cat("Apul column names (first 10):", paste(head(colnames(apul_ortholog_expression), 10), collapse = ", "), "\n")
cat("Apul column names (last 10):", paste(tail(colnames(apul_ortholog_expression), 10), collapse = ", "), "\n\n")

cat("Peve columns:", ncol(peve_ortholog_expression), "\n")
cat("Peve column names (first 10):", paste(head(colnames(peve_ortholog_expression), 10), collapse = ", "), "\n")
cat("Peve column names (last 10):", paste(tail(colnames(peve_ortholog_expression), 10), collapse = ", "), "\n\n")

cat("Ptua columns:", ncol(ptua_ortholog_expression), "\n")
cat("Ptua column names (first 10):", paste(head(colnames(ptua_ortholog_expression), 10), collapse = ", "), "\n")
cat("Ptua column names (last 10):", paste(tail(colnames(ptua_ortholog_expression), 10), collapse = ", "), "\n\n")

```

### Summary statistics

```{r summary-statistics, eval=TRUE}
cat("=== LINE COUNTS FOR ORTHOLOG EXPRESSION DATA ===\n")
cat("Apul ortholog expression with info: ", nrow(apul_ortholog_expression), " rows\n")
cat("Peve ortholog expression with info: ", nrow(peve_ortholog_expression), " rows\n")
cat("Ptua ortholog expression with info: ", nrow(ptua_ortholog_expression), " rows\n\n")
```

### Sample filtering verification

```{r sample-filtering-verification, eval=TRUE}
cat("=== SAMPLE FILTERING VERIFICATION ===\n")
cat("This analysis filters samples to include ONLY those with all four timepoints (TP1, TP2, TP3, TP4)\n\n")

# Function to analyze sample completeness in the filtered data
analyze_sample_completeness <- function(data_df, species_name) {
  cat("Analyzing", species_name, "sample completeness:\n")
  
  # Get expression columns (exclude group_id)
  expr_cols <- setdiff(colnames(data_df), "group_id")
  
  # Parse sample column names to identify sample groups and timepoints
  sample_timepoint_map <- list()
  
  for(col in expr_cols) {
    parts <- strsplit(col, "\\.")[[1]]  # Split on dots to match actual format
    
    if(length(parts) >= 3) {
      tp_part <- parts[length(parts)]
      
      if(grepl("^TP[0-9]+$", tp_part)) {
        timepoint <- as.numeric(gsub("TP", "", tp_part))
        sample_id <- paste(parts[1:(length(parts)-1)], collapse = ".")
        
        if(!sample_id %in% names(sample_timepoint_map)) {
          sample_timepoint_map[[sample_id]] <- c()
        }
        sample_timepoint_map[[sample_id]] <- c(sample_timepoint_map[[sample_id]], timepoint)
      }
    }
  }
  
  # Check completeness
  complete_samples <- 0
  incomplete_samples <- 0
  required_timepoints <- c(1, 2, 3, 4)
  
  for(sample_id in names(sample_timepoint_map)) {
    available_timepoints <- sort(sample_timepoint_map[[sample_id]])
    
    if(all(required_timepoints %in% available_timepoints)) {
      complete_samples <- complete_samples + 1
    } else {
      incomplete_samples <- incomplete_samples + 1
      missing_tps <- setdiff(required_timepoints, available_timepoints)
      cat("  WARNING: Sample", sample_id, "missing timepoints:", paste(paste0("TP", missing_tps), collapse = ", "), "\n")
    }
  }
  
  cat("  Total samples:", length(sample_timepoint_map), "\n")
  cat("  Complete samples (all 4 timepoints):", complete_samples, "\n")
  cat("  Incomplete samples:", incomplete_samples, "\n")
  
  if(incomplete_samples > 0) {
    cat("  ERROR: Filtering did not work correctly!\n")
  } else {
    cat("  SUCCESS: All samples have complete timepoints\n")
  }
  
  cat("\n")
  
  return(list(
    total = length(sample_timepoint_map),
    complete = complete_samples,
    incomplete = incomplete_samples
  ))
}

# Verify each species
apul_stats <- analyze_sample_completeness(apul_ortholog_expression, "Apul")
peve_stats <- analyze_sample_completeness(peve_ortholog_expression, "Peve")
ptua_stats <- analyze_sample_completeness(ptua_ortholog_expression, "Ptua")

cat("=== FILTERING SUMMARY ===\n")
cat("All samples in the filtered dataset should have exactly 4 timepoints (TP1, TP2, TP3, TP4)\n")
cat("Apul: ", apul_stats$complete, "/", apul_stats$total, " complete samples\n")
cat("Peve: ", peve_stats$complete, "/", peve_stats$total, " complete samples\n")
cat("Ptua: ", ptua_stats$complete, "/", ptua_stats$total, " complete samples\n")

total_incomplete <- apul_stats$incomplete + peve_stats$incomplete + ptua_stats$incomplete
if(total_incomplete == 0) {
  cat("SUCCESS: All samples across all species have complete timepoints!\n")
} else {
  cat("ERROR: Found", total_incomplete, "incomplete samples - filtering needs to be fixed!\n")
}
cat("\n")
```

### Diagnostic analysis

```{r diagnostic-analysis, eval=TRUE}
cat("\n=== VERIFICATION: THREE-WAY ORTHOLOG COUNTS ===\n")
cat("After filtering, all species should have identical three_way ortholog counts.\n\n")

# Check how many genes we have for each species
cat("Genes found in expression data by species:\n")
cat("Apul ortholog expression (before adding info):", nrow(apul_ortholog_expression), "\n")
cat("Peve ortholog expression (before adding info):", nrow(peve_ortholog_expression), "\n")
cat("Ptua ortholog expression (before adding info):", nrow(ptua_ortholog_expression), "\n\n")

# Verify all ortholog groups are three-way
cat("Complete three-way ortholog groups available:", nrow(complete_ortholog_groups), "\n")
cat("All should be type 'three_way'? Check:", table(complete_ortholog_groups$type), "\n\n")

# Verify perfect gene coverage (should be 100% for all species now)
apul_coverage <- length(intersect(apul_ortholog_genes, apul_gene_matrix$gene_id_clean))
peve_coverage <- length(intersect(peve_ortholog_genes, peve_gene_matrix$gene_id_clean))
ptua_coverage <- length(intersect(ptua_ortholog_genes, ptua_gene_matrix$gene_id_clean))

cat("Gene coverage in expression data (should be 100% for all):\n")
cat("Apul: ", apul_coverage, "/", length(apul_ortholog_genes), " (", round(apul_coverage/length(apul_ortholog_genes)*100, 1), "%)\n")
cat("Peve: ", peve_coverage, "/", length(peve_ortholog_genes), " (", round(peve_coverage/length(peve_ortholog_genes)*100, 1), "%)\n")
cat("Ptua: ", ptua_coverage, "/", length(ptua_ortholog_genes), " (", round(ptua_coverage/length(ptua_ortholog_genes)*100, 1), "%)\n\n")
```

### Preview expression data

```{r preview-expression-data, eval=TRUE}
cat("\n=== PREVIEW OF ORTHOLOG EXPRESSION DATA ===\n")

cat("Apul ortholog expression:\n\n")
str(apul_ortholog_expression)
cat("\n\n")

cat("\nPeve ortholog expression:\n")
str(peve_ortholog_expression)
cat("\n\n")

cat("\nPtua ortholog expression:\n")
str(ptua_ortholog_expression)
cat("\n\n")
```

# BARNACLE ANALYSIS

This workflow focuses on comparing multiple ranks to find the optimal number of components.

Based on the barnacle workflow, we: 1. Normalize count data with `sctransform` 2. Create tensors for multiomics analysis 3. Compare multiple ranks systematically 4. Choose optimal rank based on metrics

## Load expression data

```{r load-normalized-data, eval=TRUE}
# Read the exported ortholog expression data
apul_expr <- read.csv(file.path(output_dir, "apul_ortholog_expression.csv"))
peve_expr <- read.csv(file.path(output_dir, "peve_ortholog_expression.csv"))  
ptua_expr <- read.csv(file.path(output_dir, "ptua_ortholog_expression.csv"))

cat("Loaded expression data:\n")
cat("Apul:", nrow(apul_expr), "genes x", ncol(apul_expr)-1, "samples\n")
cat("Peve:", nrow(peve_expr), "genes x", ncol(peve_expr)-1, "samples\n") 
cat("Ptua:", nrow(ptua_expr), "genes x", ncol(ptua_expr)-1, "samples\n")
```

## Normalize data with sctransform

Following the barnacle manuscript approach, we'll use `sctransform` to normalize each species' data. We'll use a bulk RNA-seq appropriate approach.

```{r normalize-sctransform, eval=TRUE}
# Function to normalize count data with sctransform for bulk RNA-seq
normalize_with_sctransform <- function(count_data, species_name) {
  cat("Normalizing", species_name, "data with sctransform...\n")
  
  # Check if we have group_id or gene_id column
  id_col <- if("group_id" %in% colnames(count_data)) "group_id" else "gene_id"
  cat("Using", id_col, "as identifier column\n")
  
  # Check for and handle duplicate group_ids/gene_ids
  duplicate_ids <- count_data[[id_col]][duplicated(count_data[[id_col]])]
  if(length(duplicate_ids) > 0) {
    cat("Found", length(unique(duplicate_ids)), "duplicate", id_col, "values:\n")
    cat("  Examples:", head(unique(duplicate_ids), 10), "\n")
    cat("  Note: sctransform may fail due to duplicate row names, will fall back to log2(CPM + 1)\n")
  } else {
    cat("No duplicate", id_col, "values found\n")
  }
  
  # Use original data without aggregation - let sctransform handle duplicates or fail
  agg_data <- count_data
  
  # Extract count matrix (genes as rows, samples as columns)
  count_matrix <- as.matrix(agg_data[, -1])  # Remove id column
  rownames(count_matrix) <- agg_data[[id_col]]
  
  # Check for and handle problematic values
  cat("Checking data quality...\n")
  cat("  - Zero values:", sum(count_matrix == 0), "/", length(count_matrix), "\n")
  cat("  - NA values:", sum(is.na(count_matrix)), "\n")
  cat("  - Infinite values:", sum(is.infinite(count_matrix)), "\n")
  cat("  - Min value:", min(count_matrix, na.rm = TRUE), "\n")
  cat("  - Max value:", max(count_matrix, na.rm = TRUE), "\n")
  
  # Remove genes with all zeros or very low expression
  gene_sums <- rowSums(count_matrix)
  keep_genes <- gene_sums > 10  # Keep genes with total counts > 10
  count_matrix_filtered <- count_matrix[keep_genes, , drop = FALSE]
  
  cat("  - Filtered to", nrow(count_matrix_filtered), "ortholog groups (from", nrow(count_matrix), ")\n")
  
  # Ensure data is numeric and handle any potential issues
  count_matrix_filtered <- apply(count_matrix_filtered, c(1,2), function(x) {
    if(is.na(x) || !is.finite(x)) return(0)
    return(as.numeric(x))
  })
  
  # Ensure row and column names are character strings (not factors)
  rownames(count_matrix_filtered) <- as.character(rownames(count_matrix_filtered))
  colnames(count_matrix_filtered) <- as.character(colnames(count_matrix_filtered))
  
  # sctransform expects genes as rows and cells (samples) as columns —
  # our matrix is already genes x samples (rows=genes, cols=samples), so do NOT transpose.
  count_matrix_for_vst <- count_matrix_filtered
  
  # Apply sctransform normalization with bulk RNA-seq appropriate parameters
  normalized_df <- tryCatch({
    # First try with minimal parameters to isolate the issue
    cat("  Attempting sctransform with minimal parameters...\n")
    # Pass genes x samples matrix directly
    normalized <- sctransform::vst(
      count_matrix_for_vst,
      verbosity = 1
    )

    # Extract normalized data (should be genes x samples)
    normalized_data <- normalized$y
    
    # Create output data frame with original gene set (fill missing with zeros)
    full_normalized_data <- matrix(0, nrow = nrow(count_matrix), ncol = ncol(count_matrix))
    rownames(full_normalized_data) <- rownames(count_matrix)
    colnames(full_normalized_data) <- colnames(count_matrix)
    
    # Fill in normalized values for kept genes
    full_normalized_data[rownames(normalized_data), ] <- normalized_data
    
    result_df <- data.frame(
      group_id = rownames(full_normalized_data),
      full_normalized_data,
      stringsAsFactors = FALSE
    )
    
    cat("sctransform normalization successful for", species_name, "\n")
    return(result_df)
    
  }, error = function(e) {
    cat("sctransform with minimal parameters failed for", species_name, ":", e$message, "\n")
    cat("Trying sctransform with glmGamPoi method...\n")
    
    # Try again with glmGamPoi method
    tryCatch({
      normalized <- sctransform::vst(
        count_matrix_for_vst, 
        method = "glmGamPoi",
        n_genes = min(2000, nrow(count_matrix_for_vst)),
        return_cell_attr = TRUE,
        verbosity = 1
      )

      # Extract normalized data (genes x samples)
      normalized_data <- normalized$y
      
      # Create output data frame with original gene set (fill missing with zeros)
      full_normalized_data <- matrix(0, nrow = nrow(count_matrix), ncol = ncol(count_matrix))
      rownames(full_normalized_data) <- rownames(count_matrix)
      colnames(full_normalized_data) <- colnames(count_matrix)
      
      # Fill in normalized values for kept genes
      full_normalized_data[rownames(normalized_data), ] <- normalized_data
      
      result_df <- data.frame(
        group_id = rownames(full_normalized_data),
        full_normalized_data,
        stringsAsFactors = FALSE
      )
      
      cat("sctransform normalization with glmGamPoi successful for", species_name, "\n")
      return(result_df)
      
    }, error = function(e2) {
      cat("sctransform failed for", species_name, ":", e2$message, "\n")
      cat("Falling back to log2(CPM + 1) normalization...\n")
      
  # Fallback: log2(CPM + 1) normalization
  # Calculate CPM (Counts Per Million) using the filtered count matrix
  lib_sizes <- colSums(count_matrix_filtered)
  cpm_matrix <- sweep(count_matrix_filtered, 2, lib_sizes/1e6, FUN = "/")
      
      # Log2 transform with pseudocount
      normalized_data <- log2(cpm_matrix + 1)
      
      result_df <- data.frame(
        group_id = rownames(normalized_data),
        normalized_data,
        stringsAsFactors = FALSE
      )
      
      cat("Log2(CPM + 1) normalization complete for", species_name, "\n")
      return(result_df)
    })
  })
  
  cat("Input dimensions:", nrow(count_data), "rows x", ncol(count_data)-1, "samples\n")
  cat("Output dimensions:", nrow(normalized_df), "ortholog groups x", ncol(normalized_df)-1, "samples\n\n")
  
  return(normalized_df)
}

# Normalize each species
cat("=== STARTING NORMALIZATION ===\n\n")
apul_normalized <- normalize_with_sctransform(apul_expr, "Apul")
peve_normalized <- normalize_with_sctransform(peve_expr, "Peve") 
ptua_normalized <- normalize_with_sctransform(ptua_expr, "Ptua")
cat("=== NORMALIZATION COMPLETE ===\n\n")
```

## Export normalized data for Python analysis

```{r export-normalized-data, eval=TRUE}
# Export normalized data for Python processing
apul_norm_file <- file.path(output_dir, "apul_normalized_expression.csv")
peve_norm_file <- file.path(output_dir, "peve_normalized_expression.csv")
ptua_norm_file <- file.path(output_dir, "ptua_normalized_expression.csv")

write.csv(apul_normalized, apul_norm_file, row.names = FALSE, quote = FALSE)
write.csv(peve_normalized, peve_norm_file, row.names = FALSE, quote = FALSE)
write.csv(ptua_normalized, ptua_norm_file, row.names = FALSE, quote = FALSE)

cat("Exported normalized data:\n")
cat("Apul:", apul_norm_file, "\n")
cat("Peve:", peve_norm_file, "\n")
cat("Ptua:", ptua_norm_file, "\n\n")
```

## Create tensor dataset in Python

Now we'll switch to Python to create the multiomics tensor and run barnacle analysis.

```{python setup-logging, eval=TRUE}
import sys
import os
from datetime import datetime

# Set up logging to both console and file
class Logger:
    def __init__(self, log_file):
        self.terminal = sys.stdout
        self.log = open(log_file, 'w')
        
    def write(self, message):
        self.terminal.write(message)
        self.log.write(message)
        self.log.flush()
        
    def flush(self):
        self.terminal.flush()
        self.log.flush()

# Create log file path
output_dir = r.output_dir
log_file = os.path.join(output_dir, f'barnacle_analysis_log_{datetime.now().strftime("%Y%m%d_%H%M%S")}.txt')

# Redirect stdout to both console and log file
sys.stdout = Logger(log_file)
sys.stderr = sys.stdout  # Also capture error messages

print("="*60)
print(f"BARNACLE ANALYSIS LOG")
print(f"Started: {datetime.now().strftime('%Y-%m-%d %H:%M:%S')}")
print(f"Log file: {log_file}")
print("="*60)
print()
```

```{python create-multiomics-tensor, eval=TRUE}
import pandas as pd
import numpy as np
import os
from pathlib import Path

# Set up paths
output_dir = r.output_dir
print(f"Working in output directory: {output_dir}")

# Load normalized data
apul_norm = pd.read_csv(os.path.join(output_dir, "apul_normalized_expression.csv"))
peve_norm = pd.read_csv(os.path.join(output_dir, "peve_normalized_expression.csv"))
ptua_norm = pd.read_csv(os.path.join(output_dir, "ptua_normalized_expression.csv"))

print("Loaded normalized data:")
print(f"Apul: {apul_norm.shape}")
print(f"Peve: {peve_norm.shape}")  
print(f"Ptua: {ptua_norm.shape}")

# Check which genes are common across all species
apul_genes = set(apul_norm['group_id'])
peve_genes = set(peve_norm['group_id'])
ptua_genes = set(ptua_norm['group_id'])

common_genes = apul_genes & peve_genes & ptua_genes
print(f"\nCommon genes across all species: {len(common_genes)}")

# Filter to common genes and align gene order
common_genes_list = sorted(list(common_genes))

apul_common = apul_norm[apul_norm['group_id'].isin(common_genes_list)].set_index('group_id').reindex(common_genes_list)
peve_common = peve_norm[peve_norm['group_id'].isin(common_genes_list)].set_index('group_id').reindex(common_genes_list)
ptua_common = ptua_norm[ptua_norm['group_id'].isin(common_genes_list)].set_index('group_id').reindex(common_genes_list)

print(f"\nFiltered to common genes:")
print(f"Apul: {apul_common.shape}")
print(f"Peve: {peve_common.shape}")
print(f"Ptua: {ptua_common.shape}")
```

## Parse sample information

```{python parse-sample-info, eval=TRUE}
# Parse sample names to extract sample information for each species independently
def parse_species_samples(columns, species_name):
    """Parse sample column names for a specific species"""
    sample_map = {}
    sample_ids = []
    timepoints = set()
    
    for col in columns:
        # Expected format: PREFIX-NUMBER-TP# (e.g., ACR-139-TP1, POR-216-TP1, POC-201-TP1)
        # Also support legacy format: PREFIX.NUMBER.TP# for backwards compatibility
        parts = col.split('-') if '-' in col else col.split('.')
        
        if len(parts) >= 3 and parts[-1].startswith('TP'):
            # Extract timepoint from last part (e.g., "TP1", "TP2", etc.)
            tp_part = parts[-1]
            timepoint = int(tp_part[2:])  # e.g., 1 from TP1
            
            # Sample ID is everything except the timepoint part
            sample_id_parts = parts[:-1]
            sample_id = '-'.join(sample_id_parts) if '-' in col else '.'.join(sample_id_parts)
            
            sample_map[(sample_id, timepoint)] = col
            if sample_id not in sample_ids:
                sample_ids.append(sample_id)
            timepoints.add(timepoint)
        else:
            print(f"Warning: Could not parse column name: {col}")
    
    return sample_map, sample_ids, sorted(timepoints)

# Parse sample information for each species independently
print("Parsing sample information for each species...")

species_data = {
    'apul': apul_common,
    'peve': peve_common, 
    'ptua': ptua_common
}

species_info = {}
all_timepoints = set()

for species, data in species_data.items():
    sample_map, sample_ids, timepoints = parse_species_samples(data.columns, species)
    species_info[species] = {
        'sample_map': sample_map,
        'sample_ids': sample_ids,
        'timepoints': timepoints,
        'n_samples': len(sample_ids)
    }
    all_timepoints.update(timepoints)
    
    print(f"{species}:")
    print(f"  Samples: {len(sample_ids)} ({sample_ids[:3]}...)")
    print(f"  Timepoints: {timepoints}")

common_timepoints = sorted(list(all_timepoints))
print(f"\nTimepoints found across all species: {common_timepoints}")

# Find the maximum number of samples to determine tensor dimensions
max_samples = max(info['n_samples'] for info in species_info.values())
print(f"Maximum samples in any species: {max_samples}")

# Print detailed sample structure
print(f"\nDetailed sample structure:")
for species, info in species_info.items():
    print(f"{species}: {info['n_samples']} samples × {len(info['timepoints'])} timepoints")
```

## Create 3D tensor (genes × species_samples × timepoints)

```{python create-3d-tensor, eval=TRUE}
# Create a 3D tensor: genes × (species_samples) × timepoints
# This flattens species and samples into a single dimension that Barnacle can handle

print("Creating 3D tensor by combining species and samples...")
print("Note: R-level filtering has already removed samples without all 4 timepoints")

# First, collect all actual sample-timepoint combinations that have data
all_sample_columns = []
sample_labels = []  # Track which sample belongs to which species
species_sample_map = {}  # Map from combined index to (species, sample_idx, sample_id)

sample_idx = 0
for species in ['apul', 'peve', 'ptua']:
    data = species_data[species]
    info = species_info[species]
    
    print(f"\nProcessing {species}:")
    for sample_id in info['sample_ids']:
        # Check if this sample has data for ALL required timepoints
        available_timepoints = []
        sample_timepoint_cols = []
        
        for timepoint in common_timepoints:
            if (sample_id, timepoint) in info['sample_map']:
                col_name = info['sample_map'][(sample_id, timepoint)]
                sample_timepoint_cols.append(col_name)
                available_timepoints.append(timepoint)
        
        # Only include samples that have ALL timepoints (should be all samples due to R filtering)
        if len(available_timepoints) == len(common_timepoints):
            all_sample_columns.extend(sample_timepoint_cols)
            sample_labels.append(f"{species}_{sample_id}")
            species_sample_map[sample_idx] = {
                'species': species,
                'sample_id': sample_id,
                'sample_idx_in_species': info['sample_ids'].index(sample_id)
            }
            sample_idx += 1
            print(f"  Added {sample_id} with {len(sample_timepoint_cols)} timepoints: {sorted(available_timepoints)}")
        else:
            print(f"  Skipped {sample_id} - incomplete timepoints: {sorted(available_timepoints)} (expected: {sorted(common_timepoints)})")

n_genes = len(common_genes_list)
n_combined_samples = len(sample_labels)
n_timepoints = len(common_timepoints)

print(f"\nCreating 3D tensor with shape: ({n_genes}, {n_combined_samples}, {n_timepoints})")
print(f"Combined samples from all species: {n_combined_samples}")

# Initialize tensor
tensor_3d = np.full((n_genes, n_combined_samples, n_timepoints), np.nan)

# Fill tensor
filled_count = 0
missing_count = 0

for combined_idx, sample_label in enumerate(sample_labels):
    species_info_map = species_sample_map[combined_idx]
    species = species_info_map['species']
    sample_id = species_info_map['sample_id']
    
    data = species_data[species]
    info = species_info[species]
    
    for time_idx, timepoint in enumerate(common_timepoints):
        if (sample_id, timepoint) in info['sample_map']:
            col_name = info['sample_map'][(sample_id, timepoint)]
            tensor_3d[:, combined_idx, time_idx] = data[col_name].values
            filled_count += 1
        else:
            missing_count += 1

# Check tensor statistics
n_missing = np.sum(np.isnan(tensor_3d))
n_total = tensor_3d.size
n_finite = np.sum(np.isfinite(tensor_3d))

print(f"\n=== TENSOR STATISTICS ===")
print(f"Tensor shape: {tensor_3d.shape}")
print(f"Total elements: {n_total}")
print(f"Finite values: {n_finite}")
print(f"Missing/NaN values: {n_missing}")
print(f"Missing percentage: {n_missing / n_total * 100:.2f}%")
print(f"Filled {filled_count} sample-timepoint combinations")
print(f"Missing {missing_count} sample-timepoint combinations")

# Check non-zero values among finite values
finite_values = tensor_3d[np.isfinite(tensor_3d)]
n_nonzero = np.sum(finite_values != 0)
print(f"Non-zero finite values: {n_nonzero}")
print(f"Zero finite values: {len(finite_values) - n_nonzero}")
print(f"Sparsity among finite values: {(len(finite_values) - n_nonzero) / len(finite_values) * 100:.2f}%")

# Save sample mapping for later interpretation
sample_mapping = pd.DataFrame([
    {
        'combined_index': i,
        'sample_label': label,
        'species': species_sample_map[i]['species'],
        'sample_id': species_sample_map[i]['sample_id']
    }
    for i, label in enumerate(sample_labels)
])
print(f"\nSample mapping:")
print(sample_mapping.head(10))
```



## Rank Comparison

**This workflow systematically compares multiple ranks to find the optimal number of components.**

The comparison runs decompositions for each specified rank and generates:
1. Metrics comparison table
2. Elbow plot (variance explained vs rank)
3. Multi-panel summary plots
4. Automated recommendation

**Results:** Each rank is saved to `rank_XX/` directory with complete factor matrices.

### Define helper functions for rank comparison

```{python rank-comparison-functions, eval=TRUE}
def run_single_rank_decomposition(tensor, rank, lambdas, n_iter_max, random_state=42):
    """
    Run sparse CP decomposition for a given rank.
    Returns dict with results and metrics.
    """
    from barnacle.decomposition import SparseCP
    
    print(f"\n{'='*60}")
    print(f"Running decomposition with rank = {rank}")
    print(f"{'='*60}")
    
    # Handle missing values
    tensor_filled = np.nan_to_num(tensor, nan=0.0)
    
    # Create model
    model = SparseCP(
        rank=rank,
        lambdas=lambdas,
        nonneg_modes=[0],
        norm_constraint=True,
        init='random',
        tol=1e-5,
        n_iter_max=n_iter_max,
        random_state=random_state,
        n_initializations=1
    )
    
    # Fit model
    try:
        decomposition = model.fit_transform(tensor_filled, verbose=0)
        
        # Calculate metrics
        metrics = calculate_decomposition_metrics(tensor_filled, decomposition, model)
        
        results = {
            'rank': rank,
            'decomposition': decomposition,
            'model': model,
            'metrics': metrics,
            'success': True
        }
        
        print(f"✓ Rank {rank} complete")
        print(f"  Final loss: {metrics['final_loss']:.6f}")
        print(f"  Variance explained: {metrics['variance_explained']*100:.2f}%")
        print(f"  Converged: {metrics['converged']}")
        
        return results
        
    except Exception as e:
        print(f"✗ Rank {rank} failed: {e}")
        return {
            'rank': rank,
            'success': False,
            'error': str(e)
        }


def calculate_decomposition_metrics(tensor, decomposition, model):
    """
    Calculate reconstruction quality and other metrics.
    """
    # Reconstruct tensor
    gene_factors = decomposition.factors[0]
    sample_factors = decomposition.factors[1]
    time_factors = decomposition.factors[2]
    weights = decomposition.weights
    
    rank = len(weights)
    reconstructed = np.zeros_like(tensor)
    
    for r in range(rank):
        outer_prod = np.outer(gene_factors[:, r], sample_factors[:, r])
        outer_prod = np.outer(outer_prod.flatten(), time_factors[:, r])
        outer_prod = outer_prod.reshape(tensor.shape)
        reconstructed += weights[r] * outer_prod
    
    # Calculate errors
    mse = np.mean((tensor - reconstructed) ** 2)
    rmse = np.sqrt(mse)
    
    # Relative error
    tensor_norm = np.linalg.norm(tensor)
    error_norm = np.linalg.norm(tensor - reconstructed)
    relative_error = error_norm / tensor_norm if tensor_norm > 0 else np.inf
    
    # Variance explained
    total_var = np.var(tensor)
    residual_var = np.var(tensor - reconstructed)
    variance_explained = 1 - (residual_var / total_var) if total_var > 0 else 0
    
    # Sparsity metrics
    gene_sparsity = np.mean(np.abs(gene_factors) < 0.1)
    sample_sparsity = np.mean(np.abs(sample_factors) < 0.1)
    time_sparsity = np.mean(np.abs(time_factors) < 0.1)
    
    # Component weight statistics
    weight_max = np.max(weights)
    weight_min = np.min(weights)
    weight_ratio = weight_max / weight_min if weight_min > 0 else np.inf
    weight_cv = np.std(weights) / np.mean(weights) if np.mean(weights) > 0 else np.inf
    
    # Convergence info
    # Note: Barnacle's SparseCP class does not have a converged_ attribute
    # We determine convergence by checking if iterations < max_iter
    n_iterations = len(model.loss_) if hasattr(model, 'loss_') and model.loss_ is not None else 0
    final_loss = model.loss_[-1] if hasattr(model, 'loss_') and model.loss_ is not None else np.nan
    # Converged if stopped before reaching max iterations (met tolerance criterion)
    converged = (n_iterations < model.n_iter_max) if n_iterations > 0 else False
    
    return {
        'mse': mse,
        'rmse': rmse,
        'relative_error': relative_error,
        'variance_explained': variance_explained,
        'gene_sparsity': gene_sparsity,
        'sample_sparsity': sample_sparsity,
        'time_sparsity': time_sparsity,
        'weight_max': weight_max,
        'weight_min': weight_min,
        'weight_ratio': weight_ratio,
        'weight_cv': weight_cv,
        'converged': converged,
        'n_iterations': n_iterations,
        'final_loss': final_loss
    }


def save_rank_results(results, output_dir, gene_ids, sample_labels, timepoint_labels, species_sample_map):
    """
    Save decomposition results for a specific rank.
    """
    rank = results['rank']
    rank_dir = os.path.join(output_dir, f'rank_{rank:02d}')
    os.makedirs(rank_dir, exist_ok=True)
    
    decomposition = results['decomposition']
    metrics = results['metrics']
    
    # Save factor matrices
    gene_factors = pd.DataFrame(
        decomposition.factors[0],
        index=gene_ids,
        columns=[f'Component_{i+1}' for i in range(rank)]
    )
    gene_factors.to_csv(os.path.join(rank_dir, 'gene_factors.csv'))
    
    sample_factors = pd.DataFrame(
        decomposition.factors[1],
        index=sample_labels,
        columns=[f'Component_{i+1}' for i in range(rank)]
    )
    sample_factors['Species'] = [species_sample_map[i]['species'] for i in range(len(sample_labels))]
    sample_factors['Sample_ID'] = [species_sample_map[i]['sample_id'] for i in range(len(sample_labels))]
    sample_factors.to_csv(os.path.join(rank_dir, 'sample_factors.csv'))
    
    time_factors = pd.DataFrame(
        decomposition.factors[2],
        index=timepoint_labels,
        columns=[f'Component_{i+1}' for i in range(rank)]
    )
    time_factors.to_csv(os.path.join(rank_dir, 'time_factors.csv'))
    
    # Save weights
    weights_df = pd.DataFrame({
        'Component': [f'Component_{i+1}' for i in range(rank)],
        'Weight': decomposition.weights
    })
    weights_df.to_csv(os.path.join(rank_dir, 'component_weights.csv'), index=False)
    
    # Save loss history
    if hasattr(results['model'], 'loss_') and results['model'].loss_ is not None:
        loss_df = pd.DataFrame({
            'Iteration': list(range(1, len(results['model'].loss_) + 1)),
            'Loss': results['model'].loss_
        })
        loss_df.to_csv(os.path.join(rank_dir, 'loss_history.csv'), index=False)
    
    # Save metadata
    metadata = {
        'rank': rank,
        'metrics': {k: float(v) if not isinstance(v, bool) else v for k, v in metrics.items()},
        'tensor_shape': list(decomposition.factors[0].shape[0:1]) + 
                       [decomposition.factors[1].shape[0]] + 
                       [decomposition.factors[2].shape[0]]
    }
    
    import json
    with open(os.path.join(rank_dir, 'metadata.json'), 'w') as f:
        json.dump(metadata, f, indent=2)
    
    print(f"  Saved results to: {rank_dir}")


print("Rank comparison helper functions loaded")
```

### Run barnacle

```{python barnacle-multiple-ranks, eval=TRUE}
# ===========================================
# CONFIGURE RANKS TO TEST
# ===========================================
# NOTE: This chunk creates a machine-readable log file
# (rank_comparison_log.json) that subsequent chunks will use.
# This allows you to skip re-running this chunk if you've
# already generated results.
# ===========================================
ranks_to_test = [5, 8, 10, 12, 15, 20, 25, 35, 45, 55, 65, 75]  # Modify this list as needed
# ===========================================

import os
import json
import datetime

print("="*60)
print("COMPARING MULTIPLE RANKS")
print("="*60)
print(f"Ranks to test: {ranks_to_test}")
print(f"This will run {len(ranks_to_test)} decompositions")
print(f"Estimated time: {len(ranks_to_test) * 2}-{len(ranks_to_test) * 5} minutes")
print("="*60)

# Run decompositions for each rank
all_rank_results = []

for test_rank in sorted(ranks_to_test):
    result = run_single_rank_decomposition(
        tensor=tensor_3d,
        rank=test_rank,
        lambdas=[0.1, 0.0, 0.1],
        n_iter_max=10000,
        random_state=41
    )
    all_rank_results.append(result)
    
    # Save individual results
    if result['success']:
        save_rank_results(
            result,
            output_dir,
            common_genes_list,
            sample_labels,
            [f'TP{tp}' for tp in common_timepoints],
            species_sample_map
        )

print(f"\n{'='*60}")
print(f"ALL RANKS COMPLETE")
print(f"{'='*60}")
print(f"Successful: {sum(1 for r in all_rank_results if r['success'])}/{len(all_rank_results)}")

# ===========================================
# SAVE COMPREHENSIVE LOG FILE
# ===========================================
import json
import datetime

log_file = os.path.join(output_dir, 'rank_comparison_log.json')

log_data = {
    'timestamp': datetime.datetime.now().isoformat(),
    'ranks_tested': ranks_to_test,
    'total_runs': len(all_rank_results),
    'successful_runs': sum(1 for r in all_rank_results if r['success']),
    'failed_runs': sum(1 for r in all_rank_results if not r['success']),
    'results': []
}

for result in all_rank_results:
    rank_info = {
        'rank': result['rank'],
        'success': result['success']
    }
    
    if result['success']:
        rank_info['metrics'] = {
            k: float(v) if isinstance(v, (int, float, np.number)) else bool(v) if isinstance(v, (bool, np.bool_)) else v
            for k, v in result['metrics'].items()
        }
        rank_info['output_dir'] = os.path.join(output_dir, f'rank_{result["rank"]:02d}')
        rank_info['has_loss_history'] = hasattr(result['model'], 'loss_') and result['model'].loss_ is not None
    else:
        rank_info['error'] = result.get('error', 'Unknown error')
    
    log_data['results'].append(rank_info)

with open(log_file, 'w') as f:
    json.dump(log_data, f, indent=2)

print(f"\nLog file saved to: {log_file}")
print("This log can be used by subsequent chunks to avoid re-running decompositions")
```

### Factor Match Score evaluation

This section evaluates each rank's decomposition quality using the Factor Match Score (FMS), which measures reconstruction quality by comparing the original tensor with the decomposed tensor.

```{python factor-match-score-evaluation, eval=TRUE}
# ===========================================
# FACTOR MATCH SCORE EVALUATION
# ===========================================
import numpy as np
import pandas as pd
import os

print(f"\n{'='*60}")
print("FACTOR MATCH SCORE (FMS) EVALUATION")
print(f"{'='*60}")
print("This analysis evaluates decomposition quality using the Factor Match Score")
print("which measures how well the decomposed factors reconstruct the original")
print("tensor structure (correlation-based reconstruction quality metric).")
print(f"{'='*60}")

def calculate_reconstruction_fms(tensor, decomposition, rank):
    """
    Calculate FMS based on reconstruction quality.
    This measures how well the decomposition recovers the original tensor
    by calculating correlation and various error metrics.
    """
    try:
        # Create reconstructed tensor
        gene_factors = decomposition.factors[0]
        sample_factors = decomposition.factors[1] 
        time_factors = decomposition.factors[2]
        weights = decomposition.weights
        
        # Reconstruct tensor
        reconstructed = np.zeros_like(tensor)
        for r in range(len(weights)):
            outer_prod = np.outer(gene_factors[:, r], sample_factors[:, r])
            outer_prod = np.outer(outer_prod.flatten(), time_factors[:, r])
            outer_prod = outer_prod.reshape(tensor.shape)
            reconstructed += weights[r] * outer_prod
        
        # Calculate correlation-based FMS
        # Flatten tensors and calculate correlation
        orig_flat = tensor.flatten()
        recon_flat = reconstructed.flatten()
        
        # Remove NaN values for correlation calculation
        valid_mask = np.isfinite(orig_flat) & np.isfinite(recon_flat)
        orig_valid = orig_flat[valid_mask]
        recon_valid = recon_flat[valid_mask]
        
        if len(orig_valid) > 0:
            correlation = np.corrcoef(orig_valid, recon_valid)[0, 1]
            # Convert correlation to FMS-like score (0-1 scale)
            fms_score = (correlation + 1) / 2  # Map [-1,1] to [0,1]
            
            # Additional metrics
            mse = np.mean((orig_valid - recon_valid) ** 2)
            rmse = np.sqrt(mse)
            mae = np.mean(np.abs(orig_valid - recon_valid))
            
            # R-squared
            ss_res = np.sum((orig_valid - recon_valid) ** 2)
            ss_tot = np.sum((orig_valid - np.mean(orig_valid)) ** 2)
            r_squared = 1 - (ss_res / ss_tot) if ss_tot > 0 else 0
        else:
            correlation = np.nan
            fms_score = 0.0
            mse = np.nan
            rmse = np.nan
            mae = np.nan
            r_squared = np.nan
        
        return {
            'fms': float(fms_score),
            'correlation': float(correlation) if len(orig_valid) > 0 else np.nan,
            'mse': float(mse),
            'rmse': float(rmse),
            'mae': float(mae),
            'r_squared': float(r_squared),
            'success': True,
            'method': 'reconstruction_correlation',
            'valid_elements': len(orig_valid)
        }
        
    except Exception as e:
        print(f"  Error calculating reconstruction FMS for rank {rank}: {e}")
        return {
            'fms': np.nan,
            'success': False,
            'error': str(e),
            'method': 'failed'
        }

def evaluate_fms_for_rank(rank, output_dir, tensor):
    """
    Evaluate Factor Match Score for a specific rank.
    """
    print(f"\n--- Evaluating FMS for Rank {rank} ---")
    
    rank_dir = os.path.join(output_dir, f'rank_{rank:02d}')
    
    # Check if rank directory exists
    if not os.path.exists(rank_dir):
        print(f"  Rank directory not found: {rank_dir}")
        return None
    
    # Load factor matrices
    try:
        gene_factors = pd.read_csv(os.path.join(rank_dir, 'gene_factors.csv'), index_col=0)
        sample_factors = pd.read_csv(os.path.join(rank_dir, 'sample_factors.csv'), index_col=0)
        time_factors = pd.read_csv(os.path.join(rank_dir, 'time_factors.csv'), index_col=0)
        weights_df = pd.read_csv(os.path.join(rank_dir, 'component_weights.csv'))
        
        # Remove metadata columns from sample_factors if present
        factor_cols = [col for col in sample_factors.columns if col.startswith('Component_')]
        sample_factor_values = sample_factors[factor_cols]
        
        print(f"  Loaded factor matrices - Genes: {gene_factors.shape}, Samples: {sample_factor_values.shape}, Time: {time_factors.shape}")
        
    except Exception as e:
        print(f"  Error loading factor matrices: {e}")
        return None
    
    # Create mock decomposition object
    class MockDecomposition:
        def __init__(self, weights, factors):
            self.weights = weights
            self.factors = factors
    
    weights = weights_df['Weight'].values
    factors = [gene_factors.values, sample_factor_values.values, time_factors.values]
    mock_decomposition = MockDecomposition(weights, factors)
    
    # Calculate FMS using reconstruction-based method
    fms_result = calculate_reconstruction_fms(tensor, mock_decomposition, rank)
    
    # Combine results
    result = {
        'rank': rank,
        'fms': fms_result['fms'] if fms_result['success'] else np.nan,
        'correlation': fms_result.get('correlation', np.nan),
        'mse': fms_result.get('mse', np.nan),
        'rmse': fms_result.get('rmse', np.nan),
        'mae': fms_result.get('mae', np.nan),
        'r_squared': fms_result.get('r_squared', np.nan),
        'success': fms_result['success'],
        'method': fms_result.get('method', 'unknown'),
        'valid_elements': fms_result.get('valid_elements', 0)
    }
    
    # Print results
    if fms_result['success']:
        print(f"  FMS: {fms_result['fms']:.4f}")
        print(f"  Correlation: {fms_result.get('correlation', 'N/A'):.4f}")
        print(f"  R²: {fms_result.get('r_squared', 'N/A'):.4f}")
        print(f"  RMSE: {fms_result.get('rmse', 'N/A'):.4f}")
        print(f"  Valid elements used: {fms_result.get('valid_elements', 0):,}")
    else:
        print(f"  FMS calculation failed: {fms_result.get('error', 'unknown error')}")
    
    # Save individual rank FMS results
    fms_file = os.path.join(rank_dir, 'factor_match_score.json')
    import json
    with open(fms_file, 'w') as f:
        json.dump({
            'rank': rank,
            'fms_result': fms_result,
            'summary': result
        }, f, indent=2, default=str)
    
    print(f"  Saved FMS results to: factor_match_score.json")
    
    return result

# ===========================================
# EVALUATE FMS FOR ALL RANKS
# ===========================================

# Get output directory if not available
if 'output_dir' not in globals():
    output_dir = r.output_dir

# Get tensor if not available
if 'tensor_3d' not in globals():
    print("ERROR: tensor_3d not found in memory!")
    print("Please run the tensor creation chunks first.")
    tensor_3d = None

if tensor_3d is not None:
    # Load log file to get successful ranks
    log_file = os.path.join(output_dir, 'rank_comparison_log.json')
    
    if os.path.exists(log_file):
        import json
        with open(log_file, 'r') as f:
            log_data = json.load(f)
        
        successful_ranks = [r['rank'] for r in log_data['results'] if r['success']]
        print(f"Found {len(successful_ranks)} successful ranks to evaluate: {successful_ranks}")
        
    else:
        # Fallback: find rank directories
        import glob
        rank_dirs = sorted(glob.glob(os.path.join(output_dir, 'rank_*')))
        successful_ranks = []
        for rank_dir in rank_dirs:
            try:
                rank_num = int(os.path.basename(rank_dir).split('_')[1])
                successful_ranks.append(rank_num)
            except:
                pass
        print(f"Found {len(successful_ranks)} rank directories: {successful_ranks}")
    
    # Evaluate FMS for each rank
    all_fms_results = []
    
    for rank in sorted(successful_ranks):
        fms_result = evaluate_fms_for_rank(rank, output_dir, tensor_3d)
        if fms_result is not None:
            all_fms_results.append(fms_result)
    
    # ===========================================
    # CREATE FMS COMPARISON TABLE AND PLOTS
    # ===========================================
    if len(all_fms_results) > 0:
        print(f"\n{'='*60}")
        print("FMS RESULTS SUMMARY")
        print(f"{'='*60}")
        
        # Create comparison dataframe
        fms_df = pd.DataFrame(all_fms_results)
        
        # Save comprehensive FMS results
        comparison_dir = os.path.join(output_dir, 'rank_comparison')
        os.makedirs(comparison_dir, exist_ok=True)
        
        fms_output_file = os.path.join(comparison_dir, 'factor_match_scores.csv')
        fms_df.to_csv(fms_output_file, index=False)
        print(f"Saved FMS comparison table to: {fms_output_file}")
        
        # Display results
        display_cols = ['rank', 'fms', 'correlation', 'r_squared', 'rmse']
        print("\nFactor Match Score Results:")
        print(fms_df[display_cols].to_string(index=False, float_format='%.4f'))
        
        # Create visualization
        import matplotlib.pyplot as plt
        
        fig, axes = plt.subplots(2, 2, figsize=(14, 10))
        
        # Plot 1: FMS scores
        valid_fms = fms_df['success']
        
        if valid_fms.any():
            ax = axes[0, 0]
            ax.plot(fms_df.loc[valid_fms, 'rank'], 
                    fms_df.loc[valid_fms, 'fms'], 
                    'o-', linewidth=2, markersize=8, color='steelblue')
            ax.set_xlabel('Rank (Number of Components)', fontsize=11)
            ax.set_ylabel('Factor Match Score', fontsize=11)
            ax.set_title('Factor Match Score vs Rank\n(Higher = Better Factor Recovery)', fontweight='bold')
            ax.grid(True, alpha=0.3)
            ax.set_ylim([0, 1.05])
        
        # Plot 2: Reconstruction correlation
        if valid_fms.any():
            ax = axes[0, 1]
            ax.plot(fms_df.loc[valid_fms, 'rank'], 
                    fms_df.loc[valid_fms, 'correlation'], 
                    'o-', linewidth=2, markersize=8, color='green')
            ax.axhline(y=0.8, color='orange', linestyle='--', alpha=0.5, label='80% correlation')
            ax.axhline(y=0.9, color='red', linestyle='--', alpha=0.5, label='90% correlation')
            ax.set_xlabel('Rank (Number of Components)', fontsize=11)
            ax.set_ylabel('Reconstruction Correlation', fontsize=11)
            ax.set_title('Tensor Reconstruction Quality vs Rank\n(Correlation between original and reconstructed)', fontweight='bold')
            ax.grid(True, alpha=0.3)
            ax.legend()
            ax.set_ylim([-1.05, 1.05])
        
        # Plot 3: R-squared
        if valid_fms.any():
            ax = axes[1, 0]
            ax.plot(fms_df.loc[valid_fms, 'rank'], 
                    fms_df.loc[valid_fms, 'r_squared'], 
                    's-', linewidth=2, markersize=8, color='purple')
            ax.axhline(y=0.8, color='orange', linestyle='--', alpha=0.5, label='80% explained')
            ax.axhline(y=0.9, color='red', linestyle='--', alpha=0.5, label='90% explained')
            ax.set_xlabel('Rank (Number of Components)', fontsize=11)
            ax.set_ylabel('R² (Coefficient of Determination)', fontsize=11)
            ax.set_title('R² vs Rank\n(Proportion of Variance Explained)', fontweight='bold')
            ax.grid(True, alpha=0.3)
            ax.legend()
            ax.set_ylim([0, 1.05])
        
        # Plot 4: RMSE
        if valid_fms.any():
            ax = axes[1, 1]
            ax.plot(fms_df.loc[valid_fms, 'rank'], 
                    fms_df.loc[valid_fms, 'rmse'], 
                    '^-', linewidth=2, markersize=8, color='coral')
            ax.set_xlabel('Rank (Number of Components)', fontsize=11)
            ax.set_ylabel('RMSE (Root Mean Squared Error)', fontsize=11)
            ax.set_title('Reconstruction Error vs Rank\n(Lower = Better Fit)', fontweight='bold')
            ax.grid(True, alpha=0.3)
        
        plt.tight_layout()
        fms_plot_file = os.path.join(comparison_dir, 'factor_match_scores_plot.png')
        plt.savefig(fms_plot_file, dpi=150, bbox_inches='tight')
        print(f"Saved FMS plot to: {fms_plot_file}")
        plt.close()
        
        # ===========================================
        # FMS-BASED RANK RECOMMENDATION
        # ===========================================
        print(f"\n{'='*60}")
        print("FMS-BASED RANK RECOMMENDATION")
        print(f"{'='*60}")
        
        # Find best ranks based on different criteria
        valid_df = fms_df[fms_df['success']]
        
        if len(valid_df) > 0:
            # Best FMS
            best_fms_idx = valid_df['fms'].idxmax()
            best_fms_rank = valid_df.iloc[best_fms_idx]['rank']
            best_fms_score = valid_df.iloc[best_fms_idx]['fms']
            
            # Best correlation
            best_corr_idx = valid_df['correlation'].idxmax()
            best_corr_rank = valid_df.iloc[best_corr_idx]['rank']
            best_corr_score = valid_df.iloc[best_corr_idx]['correlation']
            
            # Best R-squared
            best_r2_idx = valid_df['r_squared'].idxmax()
            best_r2_rank = valid_df.iloc[best_r2_idx]['rank']
            best_r2_score = valid_df.iloc[best_r2_idx]['r_squared']
            
            # Lowest RMSE
            best_rmse_idx = valid_df['rmse'].idxmin()
            best_rmse_rank = valid_df.iloc[best_rmse_idx]['rank']
            best_rmse_score = valid_df.iloc[best_rmse_idx]['rmse']
            
            print(f"Best FMS: Rank {int(best_fms_rank)} (score: {best_fms_score:.4f})")
            print(f"Best Correlation: Rank {int(best_corr_rank)} (score: {best_corr_score:.4f})")
            print(f"Best R²: Rank {int(best_r2_rank)} (score: {best_r2_score:.4f})")
            print(f"Lowest RMSE: Rank {int(best_rmse_rank)} (error: {best_rmse_score:.4f})")
            
            # Look for ranks with consistently high scores (top 20%)
            fms_threshold = valid_df['fms'].quantile(0.8)
            high_scoring_ranks = valid_df[valid_df['fms'] >= fms_threshold]['rank'].tolist()
            
            print(f"\nHigh-scoring ranks (top 20% by FMS): {[int(r) for r in high_scoring_ranks]}")
            print(f"FMS threshold: {fms_threshold:.4f}")
            
            # Provide interpretation
            print(f"\nFMS Interpretation:")
            print(f"• FMS measures tensor reconstruction quality (0-1 scale, higher is better)")
            print(f"• Correlation shows linear relationship between original and reconstructed")
            print(f"• R² shows proportion of variance explained (0-1 scale, higher is better)")
            print(f"• RMSE shows average reconstruction error (lower is better)")
            print(f"• Higher FMS, correlation, and R² indicate better factor recovery")
            print(f"• Consider ranks with consistently high FMS and R² scores")
            print(f"• Look for diminishing returns (elbow point) in improvement curves")
            
        else:
            print("No valid FMS results available for recommendation.")
    
    else:
        print("No FMS results to analyze.")

else:
    print("Skipping FMS evaluation - tensor not available.")

print(f"\n{'='*60}")
print("FACTOR MATCH SCORE EVALUATION COMPLETE")
print(f"{'='*60}")
if 'all_fms_results' in locals() and len(all_fms_results) > 0:
    working_fms = sum(1 for r in all_fms_results if r['success'])
    print(f"Summary: {working_fms}/{len(all_fms_results)} ranks have valid FMS results")
    if working_fms > 0:
        print(f"→ Use FMS, correlation, and R² for rank selection")
        print(f"→ Results saved to: {comparison_dir}")
    else:
        print(f"→ No valid FMS results - check tensor and factor matrices")
```


# VISUALIZATION

## Plot rank comparisons

### 1. Comparison Plots

Located in `barnacle_factors/rank_comparison/`:

- **`elbow_plot.png`** - Shows variance explained vs rank (MOST IMPORTANT)
  - Look for the "elbow" where gains diminish
  - Marginal gains are annotated on the plot
  
- **`rank_comparison_summary.png`** - 6-panel overview
- **`loss_trajectories.png`** - Convergence speed comparison

### 2. Metrics Table

File: `barnacle_factors/rank_comparison/rank_comparison_metrics.csv`

Contains all quantitative metrics for comparison.

### 3. Individual Rank Results

Each rank gets its own directory (`rank_05/`, `rank_10/`, etc.) with full factor matrices.

```{python plot-rank-comparison, eval=TRUE}
# ===========================================
# This chunk reads from rank_comparison_log.json
# and does NOT require variables from previous chunks
# ===========================================
import matplotlib.pyplot as plt
import seaborn as sns
import json
import glob
import os
import pandas as pd
import numpy as np

print(f"\n{'='*60}")
print("Creating rank comparison visualizations")
print(f"{'='*60}")

# Check if output_dir exists, if not get it from R
if 'output_dir' not in globals():
    print("Getting output_dir from R environment...")
    output_dir = r.output_dir
    print(f"  output_dir = {output_dir}")

comparison_dir = os.path.join(output_dir, 'rank_comparison')
os.makedirs(comparison_dir, exist_ok=True)

# ===========================================
# LOAD RESULTS FROM LOG FILE
# ===========================================
log_file = os.path.join(output_dir, 'rank_comparison_log.json')

if os.path.exists(log_file):
    print(f"Loading results from log file: {log_file}")
    with open(log_file, 'r') as f:
        log_data = json.load(f)
    
    print(f"  Log timestamp: {log_data['timestamp']}")
    print(f"  Total runs: {log_data['total_runs']}")
    print(f"  Successful: {log_data['successful_runs']}")
    print(f"  Failed: {log_data['failed_runs']}")
    
    # Reconstruct all_rank_results from log
    all_rank_results = []
    for result_info in log_data['results']:
        if result_info['success']:
            all_rank_results.append({
                'rank': result_info['rank'],
                'metrics': result_info['metrics'],
                'success': True
            })
        else:
            all_rank_results.append({
                'rank': result_info['rank'],
                'success': False,
                'error': result_info.get('error', 'Unknown error')
            })
    
    print(f"  Loaded {len(all_rank_results)} results from log file")
    
elif 'all_rank_results' not in globals():
    # Fallback: Load from individual metadata files
    print("Log file not found. Loading from individual rank directories...")
    all_rank_results = []
    
    # Find all rank_* directories
    rank_dirs = sorted(glob.glob(os.path.join(output_dir, 'rank_*')))
    print(f"  Found {len(rank_dirs)} rank directories")
    
    for rank_dir in rank_dirs:
        metadata_file = os.path.join(rank_dir, 'metadata.json')
        if os.path.exists(metadata_file):
            with open(metadata_file, 'r') as f:
                metadata = json.load(f)
                all_rank_results.append({
                    'rank': metadata['rank'],
                    'metrics': metadata['metrics'],
                    'success': True
                })
    
    print(f"  Loaded {len(all_rank_results)} results from disk")
else:
    print("Using all_rank_results from memory")

# Extract successful results
successful = [r for r in all_rank_results if r['success']]
if len(successful) == 0:
    print("No successful decompositions to compare!")
else:
    ranks = [r['rank'] for r in successful]
    
    # Create comparison dataframe
    comparison_df = pd.DataFrame([
        {
            'Rank': r['rank'],
            **r['metrics']
        }
        for r in successful
    ])
    
    # Save comparison table
    comparison_df.to_csv(os.path.join(comparison_dir, 'rank_comparison_metrics.csv'), index=False)
    print(f"  Saved metrics table")
    
    # Create multi-panel comparison plot
    fig, axes = plt.subplots(2, 3, figsize=(18, 10))
    fig.suptitle('Rank Comparison: Key Metrics', fontsize=16, fontweight='bold')
    
    # 1. Variance Explained
    ax = axes[0, 0]
    ax.plot(ranks, comparison_df['variance_explained'], 'o-', linewidth=2, markersize=8, color='steelblue')
    ax.axhline(y=0.8, color='green', linestyle='--', alpha=0.5, label='80% threshold')
    ax.axhline(y=0.6, color='orange', linestyle='--', alpha=0.5, label='60% threshold')
    ax.set_xlabel('Rank (Number of Components)', fontsize=11)
    ax.set_ylabel('Variance Explained', fontsize=11)
    ax.set_title('Variance Explained vs Rank', fontweight='bold')
    ax.grid(True, alpha=0.3)
    ax.legend()
    ax.set_ylim([0, 1.05])
    
    # 2. Relative Error
    ax = axes[0, 1]
    ax.plot(ranks, comparison_df['relative_error'], 'o-', linewidth=2, markersize=8, color='coral')
    ax.set_xlabel('Rank (Number of Components)', fontsize=11)
    ax.set_ylabel('Relative Reconstruction Error', fontsize=11)
    ax.set_title('Reconstruction Error vs Rank', fontweight='bold')
    ax.grid(True, alpha=0.3)
    
    # 3. Final Loss
    ax = axes[0, 2]
    ax.plot(ranks, comparison_df['final_loss'], 'o-', linewidth=2, markersize=8, color='purple')
    ax.set_xlabel('Rank (Number of Components)', fontsize=11)
    ax.set_ylabel('Final Loss', fontsize=11)
    ax.set_title('Final Loss vs Rank', fontweight='bold')
    ax.grid(True, alpha=0.3)
    
    # 4. Gene Sparsity
    ax = axes[1, 0]
    ax.plot(ranks, comparison_df['gene_sparsity'], 'o-', linewidth=2, markersize=8, color='green')
    ax.axhline(y=0.9, color='red', linestyle='--', alpha=0.5, label='90% sparse')
    ax.set_xlabel('Rank (Number of Components)', fontsize=11)
    ax.set_ylabel('Gene Factor Sparsity', fontsize=11)
    ax.set_title('Gene Sparsity vs Rank', fontweight='bold')
    ax.grid(True, alpha=0.3)
    ax.legend()
    ax.set_ylim([0, 1.05])
    
    # 5. Weight Distribution (CV)
    ax = axes[1, 1]
    ax.plot(ranks, comparison_df['weight_cv'], 'o-', linewidth=2, markersize=8, color='darkorange')
    ax.set_xlabel('Rank (Number of Components)', fontsize=11)
    ax.set_ylabel('Weight Coefficient of Variation', fontsize=11)
    ax.set_title('Component Weight Variability vs Rank', fontweight='bold')
    ax.grid(True, alpha=0.3)
    
    # 6. Iterations to Convergence
    ax = axes[1, 2]
    colors = ['green' if c else 'red' for c in comparison_df['converged']]
    ax.bar(ranks, comparison_df['n_iterations'], color=colors, alpha=0.7)
    ax.set_xlabel('Rank (Number of Components)', fontsize=11)
    ax.set_ylabel('Iterations to Convergence', fontsize=11)
    ax.set_title('Convergence Speed vs Rank\n(Green=Converged, Red=Max Iterations)', fontweight='bold')
    ax.grid(True, alpha=0.3, axis='y')
    
    plt.tight_layout()
    output_path = os.path.join(comparison_dir, 'rank_comparison_summary.png')
    plt.savefig(output_path, dpi=150, bbox_inches='tight')
    print(f"  Saved: rank_comparison_summary.png")
    plt.close()
    
    # Create elbow plot
    fig, ax = plt.subplots(figsize=(10, 6))
    
    var_exp = comparison_df['variance_explained'].values
    ax.plot(ranks, var_exp, 'o-', linewidth=3, markersize=10, color='steelblue', label='Variance Explained')
    
    # Calculate marginal gains
    if len(ranks) > 1:
        marginal_gains = np.diff(var_exp)
        for i in range(len(marginal_gains)):
            ax.annotate(f'+{marginal_gains[i]*100:.1f}%',
                       xy=(ranks[i+1], var_exp[i+1]),
                       xytext=(10, -10),
                       textcoords='offset points',
                       fontsize=9,
                       color='darkgreen',
                       bbox=dict(boxstyle='round,pad=0.3', facecolor='yellow', alpha=0.3))
    
    ax.axhline(y=0.8, color='green', linestyle='--', alpha=0.5, linewidth=2, label='80% threshold (excellent)')
    ax.axhline(y=0.6, color='orange', linestyle='--', alpha=0.5, linewidth=2, label='60% threshold (good)')
    ax.set_xlabel('Rank (Number of Components)', fontsize=13, fontweight='bold')
    ax.set_ylabel('Variance Explained', fontsize=13, fontweight='bold')
    ax.set_title('Elbow Plot: Finding Optimal Rank\n(Look for diminishing returns)', 
                fontsize=14, fontweight='bold')
    ax.grid(True, alpha=0.3)
    ax.legend(fontsize=11)
    ax.set_ylim([0, 1.05])
    ax.set_xticks(ranks)
    
    plt.tight_layout()
    output_path = os.path.join(comparison_dir, 'elbow_plot.png')
    plt.savefig(output_path, dpi=150, bbox_inches='tight')
    print(f"  Saved: elbow_plot.png")
    plt.close()
    
    # Loss trajectories
    fig, ax = plt.subplots(figsize=(10, 6))
    
    loss_plotted = False
    for r in successful:
        # Try to get loss from model object first
        if 'model' in r and hasattr(r['model'], 'loss_') and r['model'].loss_ is not None:
            loss = r['model'].loss_
            ax.plot(range(1, len(loss)+1), loss, linewidth=2, label=f'Rank {r["rank"]}', alpha=0.8)
            loss_plotted = True
        else:
            # Load from disk
            rank_dir = os.path.join(output_dir, f'rank_{r["rank"]:02d}')
            loss_file = os.path.join(rank_dir, 'loss_history.csv')
            if os.path.exists(loss_file):
                loss_df = pd.read_csv(loss_file)
                ax.plot(loss_df['Iteration'], loss_df['Loss'], linewidth=2, label=f'Rank {r["rank"]}', alpha=0.8)
                loss_plotted = True
    
    if loss_plotted:
        ax.set_xlabel('Iteration', fontsize=12)
        ax.set_ylabel('Loss', fontsize=12)
        ax.set_title('Loss Convergence Trajectories by Rank', fontsize=14, fontweight='bold')
        ax.grid(True, alpha=0.3)
        ax.legend(fontsize=10)
        ax.set_yscale('log')
        
        plt.tight_layout()
        output_path = os.path.join(comparison_dir, 'loss_trajectories.png')
        plt.savefig(output_path, dpi=150, bbox_inches='tight')
        print(f"  Saved: loss_trajectories.png")
    else:
        print(f"  Skipped loss_trajectories.png (no loss history available)")
    
    plt.close()
    
    print(f"\nAll comparison plots saved to: {comparison_dir}")
```

# RANK STABILITY ANALYSIS

## Gene stability across ranks

This analysis determines at what rank the top genes stabilize.


```{python rank-stability-analysis, eval=TRUE}
# ===========================================
# This chunk analyzes gene stability across ranks
# ===========================================
import os
import json
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import seaborn as sns

print(f"\n{'='*60}")
print("RANK STABILITY ANALYSIS: TOP GENE CONSISTENCY")
print(f"{'='*60}")

# ===========================================
# LOAD RESULTS FROM LOG FILE IF NEEDED
# ===========================================
if 'output_dir' not in globals():
    print("Getting output_dir from R environment...")
    output_dir = r.output_dir
    print(f"  output_dir = {output_dir}")

# Always reload from log file to ensure consistency
log_file = os.path.join(output_dir, 'rank_comparison_log.json')

if not os.path.exists(log_file):
    print(f"ERROR: Log file not found: {log_file}")
    print("Please run the 'barnacle-multiple-ranks' chunk first!")
    all_rank_results = []
    successful = []
else:
    print(f"Loading results from log file: {log_file}")
    with open(log_file, 'r') as f:
        log_data = json.load(f)
    
    print(f"  Log timestamp: {log_data['timestamp']}")
    print(f"  Total runs: {log_data['total_runs']}")
    print(f"  Successful: {log_data['successful_runs']}")
    print(f"  Failed: {log_data['failed_runs']}")
    
    # Reconstruct all_rank_results from log
    all_rank_results = []
    for result_info in log_data['results']:
        if result_info['success']:
            all_rank_results.append({
                'rank': result_info['rank'],
                'metrics': result_info['metrics'],
                'success': True
            })
        else:
            all_rank_results.append({
                'rank': result_info['rank'],
                'success': False,
                'error': result_info.get('error', 'Unknown error')
            })
    
    print(f"  Loaded {len(all_rank_results)} results from log file")
    
    successful = [r for r in all_rank_results if r['success']]
    successful = sorted(successful, key=lambda x: x['rank'])

if len(successful) == 0:
    print("No successful decompositions to analyze!")
else:
    # ===========================================
    # EXTRACT TOP GENES FROM EACH RANK
    # ===========================================
    print("\nExtracting top genes from each rank...")
    
    TOP_N = 100  # Number of top genes to compare
    
    top_genes_by_rank = {}
    
    for r in successful:
        rank = r['rank']
        rank_dir = os.path.join(output_dir, f'rank_{rank:02d}')
        gene_factors_file = os.path.join(rank_dir, 'gene_factors.csv')
        
        if os.path.exists(gene_factors_file):
            gene_factors = pd.read_csv(gene_factors_file, index_col=0)
            
            # Get genes with highest loadings across ALL components
            # Sum absolute values across all components for each gene
            gene_importance = gene_factors.abs().sum(axis=1)
            top_genes = set(gene_importance.nlargest(TOP_N).index)
            
            top_genes_by_rank[rank] = {
                'genes': top_genes,
                'importance': gene_importance
            }
            
            print(f"  Rank {rank:2d}: Top {TOP_N} genes extracted")
        else:
            print(f"  Rank {rank:2d}: Gene factors file not found!")
    
    ranks = sorted(top_genes_by_rank.keys())
    
    # ===========================================
    # CALCULATE JACCARD SIMILARITY MATRIX
    # ===========================================
    print(f"\nCalculating pairwise Jaccard similarities...")
    
    n_ranks = len(ranks)
    similarity_matrix = np.zeros((n_ranks, n_ranks))
    
    for i, rank1 in enumerate(ranks):
        for j, rank2 in enumerate(ranks):
            genes1 = top_genes_by_rank[rank1]['genes']
            genes2 = top_genes_by_rank[rank2]['genes']
            
            intersection = len(genes1 & genes2)
            union = len(genes1 | genes2)
            jaccard = intersection / union if union > 0 else 0
            
            similarity_matrix[i, j] = jaccard
    
    # ===========================================
    # CALCULATE CONSECUTIVE RANK OVERLAP
    # ===========================================
    print(f"\nCalculating consecutive rank overlaps...")
    
    consecutive_overlaps = []
    for i in range(len(ranks) - 1):
        rank1, rank2 = ranks[i], ranks[i+1]
        genes1 = top_genes_by_rank[rank1]['genes']
        genes2 = top_genes_by_rank[rank2]['genes']
        
        overlap = len(genes1 & genes2)
        jaccard = overlap / len(genes1 | genes2)
        percent_overlap = (overlap / TOP_N) * 100
        
        consecutive_overlaps.append({
            'rank1': rank1,
            'rank2': rank2,
            'overlap_count': overlap,
            'jaccard_similarity': jaccard,
            'percent_overlap': percent_overlap
        })
    
    overlap_df = pd.DataFrame(consecutive_overlaps)
    
    # Save to file
    comparison_dir = os.path.join(output_dir, 'rank_comparison')
    overlap_df.to_csv(os.path.join(comparison_dir, 'consecutive_rank_overlap.csv'), index=False)
    
    print("\nConsecutive Rank Overlap:")
    print(overlap_df.to_string(index=False, float_format='%.3f'))
    
    # ===========================================
    # IDENTIFY STABILITY THRESHOLD
    # ===========================================
    print(f"\n{'='*60}")
    print("STABILITY ANALYSIS")
    print(f"{'='*60}")
    
    # Find where overlap stabilizes (high consecutive overlap)
    high_overlap_threshold = 0.80  # 80% overlap
    stable_ranks = overlap_df[overlap_df['jaccard_similarity'] >= high_overlap_threshold]
    
    if len(stable_ranks) > 0:
        first_stable = stable_ranks.iloc[0]
        print(f"\nFirst stable transition: Rank {int(first_stable['rank1'])} → {int(first_stable['rank2'])}")
        print(f"  Overlap: {first_stable['overlap_count']:.0f}/{TOP_N} genes ({first_stable['percent_overlap']:.1f}%)")
        print(f"  Jaccard: {first_stable['jaccard_similarity']:.3f}")
        print(f"\n→ Model appears to stabilize around rank {int(first_stable['rank2'])}")
    else:
        print(f"\nNo consecutive ranks have ≥{high_overlap_threshold*100:.0f}% overlap")
        print("  Model may not have fully stabilized in tested range")
        print("  Consider testing higher ranks")
    
    # ===========================================
    # PLOT 1: HEATMAP OF JACCARD SIMILARITIES
    # ===========================================
    print(f"\nCreating similarity heatmap...")
    
    fig, ax = plt.subplots(figsize=(12, 10))
    
    sns.heatmap(
        similarity_matrix,
        xticklabels=ranks,
        yticklabels=ranks,
        annot=True,
        fmt='.2f',
        cmap='RdYlGn',
        vmin=0,
        vmax=1,
        square=True,
        cbar_kws={'label': 'Jaccard Similarity'},
        ax=ax
    )
    
    ax.set_xlabel('Rank', fontsize=12, fontweight='bold')
    ax.set_ylabel('Rank', fontsize=12, fontweight='bold')
    ax.set_title(f'Top {TOP_N} Gene Overlap Between Ranks\n(Jaccard Similarity)', 
                 fontsize=14, fontweight='bold', pad=20)
    
    plt.tight_layout()
    output_path = os.path.join(comparison_dir, 'gene_stability_heatmap.png')
    plt.savefig(output_path, dpi=150, bbox_inches='tight')
    print(f"  Saved: gene_stability_heatmap.png")
    plt.close()
    
    # ===========================================
    # PLOT 2: CONSECUTIVE OVERLAP LINE PLOT
    # ===========================================
    print(f"Creating consecutive overlap plot...")
    
    fig, (ax1, ax2) = plt.subplots(2, 1, figsize=(12, 10))
    
    # Plot 1: Absolute overlap count
    ax1.plot(range(len(overlap_df)), overlap_df['overlap_count'], 
             'o-', linewidth=2, markersize=8, color='steelblue')
    ax1.axhline(y=TOP_N*0.8, color='green', linestyle='--', alpha=0.5, 
                label='80% threshold')
    ax1.axhline(y=TOP_N*0.9, color='darkgreen', linestyle='--', alpha=0.5, 
                label='90% threshold')
    ax1.set_ylabel(f'Number of Overlapping Genes (out of {TOP_N})', fontsize=11)
    ax1.set_title(f'Consecutive Rank Gene Overlap\n(Top {TOP_N} Most Important Genes)', 
                  fontsize=13, fontweight='bold')
    ax1.grid(True, alpha=0.3)
    ax1.legend()
    ax1.set_xticks(range(len(overlap_df)))
    ax1.set_xticklabels([f'{int(row["rank1"])}→{int(row["rank2"])}' 
                          for _, row in overlap_df.iterrows()], rotation=45)
    
    # Plot 2: Jaccard similarity
    ax2.plot(range(len(overlap_df)), overlap_df['jaccard_similarity'], 
             'o-', linewidth=2, markersize=8, color='coral')
    ax2.axhline(y=0.8, color='green', linestyle='--', alpha=0.5, 
                label='80% similarity')
    ax2.axhline(y=0.9, color='darkgreen', linestyle='--', alpha=0.5, 
                label='90% similarity')
    ax2.set_xlabel('Rank Transition', fontsize=11)
    ax2.set_ylabel('Jaccard Similarity', fontsize=11)
    ax2.set_title('Jaccard Similarity Between Consecutive Ranks', 
                  fontsize=13, fontweight='bold')
    ax2.grid(True, alpha=0.3)
    ax2.legend()
    ax2.set_ylim([0, 1.05])
    ax2.set_xticks(range(len(overlap_df)))
    ax2.set_xticklabels([f'{int(row["rank1"])}→{int(row["rank2"])}' 
                          for _, row in overlap_df.iterrows()], rotation=45)
    
    plt.tight_layout()
    output_path = os.path.join(comparison_dir, 'consecutive_overlap_plot.png')
    plt.savefig(output_path, dpi=150, bbox_inches='tight')
    print(f"  Saved: consecutive_overlap_plot.png")
    plt.close()
    
    # ===========================================
    # EXPORT TOP GENES FOR EACH RANK
    # ===========================================
    print(f"\nExporting top {TOP_N} genes for each rank...")
    
    for rank in ranks:
        genes = top_genes_by_rank[rank]['genes']
        importance = top_genes_by_rank[rank]['importance']
        
        top_genes_df = pd.DataFrame({
            'gene_id': list(genes),
            'total_importance': [importance[g] for g in genes]
        }).sort_values('total_importance', ascending=False)
        
        rank_dir = os.path.join(output_dir, f'rank_{rank:02d}')
        output_file = os.path.join(rank_dir, f'top_{TOP_N}_genes.csv')
        top_genes_df.to_csv(output_file, index=False)
    
    print(f"  Saved top_{TOP_N}_genes.csv for each rank")
    
    # ===========================================
    # EXPORT SHARED GENES FOR CONSECUTIVE RANK COMPARISONS
    # ===========================================
    print(f"\nExporting shared genes for consecutive rank comparisons...")
    
    for i in range(len(ranks) - 1):
        rank1, rank2 = ranks[i], ranks[i+1]
        genes1 = top_genes_by_rank[rank1]['genes']
        genes2 = top_genes_by_rank[rank2]['genes']
        
        # Get shared genes
        shared_genes = genes1 & genes2
        
        # Get importance scores from both ranks
        importance1 = top_genes_by_rank[rank1]['importance']
        importance2 = top_genes_by_rank[rank2]['importance']
        
        # Create dataframe with shared genes and their importance in both ranks
        shared_df = pd.DataFrame({
            'gene_id': list(shared_genes),
            f'rank_{rank1}_importance': [importance1[g] for g in shared_genes],
            f'rank_{rank2}_importance': [importance2[g] for g in shared_genes]
        })
        
        # Sort by average importance
        shared_df['avg_importance'] = (shared_df[f'rank_{rank1}_importance'] + 
                                       shared_df[f'rank_{rank2}_importance']) / 2
        shared_df = shared_df.sort_values('avg_importance', ascending=False)
        
        # Save to comparison directory
        output_file = os.path.join(comparison_dir, f'shared_genes_rank_{rank1}_vs_{rank2}.csv')
        shared_df.to_csv(output_file, index=False)
        
        print(f"  Rank {rank1} vs {rank2}: {len(shared_genes)} shared genes")
    
    print(f"  Saved shared_genes_rank_X_vs_Y.csv files to {comparison_dir}")
    
    # ===========================================
    # EXPORT ALL PAIRWISE SHARED GENES SUMMARY
    # ===========================================
    print(f"\nCreating pairwise shared genes summary...")
    
    pairwise_summary = []
    for i, rank1 in enumerate(ranks):
        for j, rank2 in enumerate(ranks):
            if i < j:  # Only upper triangle (avoid duplicates)
                genes1 = top_genes_by_rank[rank1]['genes']
                genes2 = top_genes_by_rank[rank2]['genes']
                shared = genes1 & genes2
                
                pairwise_summary.append({
                    'rank1': rank1,
                    'rank2': rank2,
                    'shared_count': len(shared),
                    'jaccard_similarity': len(shared) / len(genes1 | genes2) if len(genes1 | genes2) > 0 else 0,
                    'percent_overlap': (len(shared) / TOP_N) * 100
                })
    
    pairwise_df = pd.DataFrame(pairwise_summary)
    pairwise_file = os.path.join(comparison_dir, 'pairwise_shared_genes_summary.csv')
    pairwise_df.to_csv(pairwise_file, index=False)
    print(f"  Saved pairwise_shared_genes_summary.csv")
    
    print(f"\n{'='*60}")
    print("STABILITY ANALYSIS COMPLETE")
    print(f"{'='*60}")
    print(f"\nAll stability plots saved to: {comparison_dir}")
    print(f"- gene_stability_heatmap.png")
    print(f"- consecutive_overlap_plot.png")
    print(f"- consecutive_rank_overlap.csv")
    print(f"- shared_genes_rank_X_vs_Y.csv (for consecutive rank pairs)")
    print(f"- pairwise_shared_genes_summary.csv (all pairwise comparisons)")
    print(f"\nTop {TOP_N} genes for each rank saved to: rank_XX/top_{TOP_N}_genes.csv")
```

# INDIVIDUAL RANK VISUALIZATIONS

## Detailed plots for each rank

This section creates comprehensive visualizations for each rank to aid in assessment and comparison.

```{python individual-rank-visualizations, eval=TRUE}
# ===========================================
# CREATE DETAILED VISUALIZATIONS FOR EACH RANK
# ===========================================
import os
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import seaborn as sns

print(f"\n{'='*60}")
print("CREATING INDIVIDUAL RANK VISUALIZATIONS")
print(f"{'='*60}")

# Get output directory
if 'output_dir' not in globals():
    output_dir = r.output_dir

# Load log file to get ranks
log_file = os.path.join(output_dir, 'rank_comparison_log.json')

if not os.path.exists(log_file):
    print(f"ERROR: Log file not found: {log_file}")
    print("Please run the 'barnacle-multiple-ranks' chunk first!")
else:
    import json
    with open(log_file, 'r') as f:
        log_data = json.load(f)
    
    # Get successful ranks
    successful_ranks = [r['rank'] for r in log_data['results'] if r['success']]
    
    print(f"Creating visualizations for {len(successful_ranks)} ranks...")
    
    # Parameters for visualization
    TOP_GENES_HEATMAP = 50  # Number of top genes to show in heatmap
    
    for rank in successful_ranks:
        print(f"\n--- Rank {rank} ---")
        rank_dir = os.path.join(output_dir, f'rank_{rank:02d}')
        
        # Load factor matrices
        gene_factors_file = os.path.join(rank_dir, 'gene_factors.csv')
        time_factors_file = os.path.join(rank_dir, 'time_factors.csv')
        weights_file = os.path.join(rank_dir, 'component_weights.csv')
        
        if not all([os.path.exists(f) for f in [gene_factors_file, time_factors_file, weights_file]]):
            print(f"  Skipping rank {rank} - missing factor files")
            continue
        
        gene_factors = pd.read_csv(gene_factors_file, index_col=0)
        time_factors = pd.read_csv(time_factors_file, index_col=0)
        weights = pd.read_csv(weights_file)
        
        # ===========================================
        # PLOT 1: GENE HEATMAP (Top genes x Components)
        # ===========================================
        print(f"  Creating gene-component heatmap...")
        
        # Get top genes by total importance
        gene_importance = gene_factors.abs().sum(axis=1)
        top_genes = gene_importance.nlargest(TOP_GENES_HEATMAP).index
        top_gene_factors = gene_factors.loc[top_genes]
        
        # Create heatmap
        fig, ax = plt.subplots(figsize=(max(12, rank*0.5), 10))
        
        sns.heatmap(
            top_gene_factors,
            cmap='RdBu_r',
            center=0,
            cbar_kws={'label': 'Component Loading'},
            xticklabels=True,
            yticklabels=True,
            ax=ax
        )
        
        ax.set_xlabel('Components', fontsize=12, fontweight='bold')
        ax.set_ylabel('Gene ID', fontsize=12, fontweight='bold')
        ax.set_title(f'Rank {rank}: Top {TOP_GENES_HEATMAP} Genes Across Components\n' + 
                    f'(Ordered by Total Importance)',
                    fontsize=13, fontweight='bold', pad=15)
        
        plt.tight_layout()
        output_path = os.path.join(rank_dir, f'gene_component_heatmap_top{TOP_GENES_HEATMAP}.png')
        plt.savefig(output_path, dpi=150, bbox_inches='tight')
        plt.close()
        print(f"    Saved: gene_component_heatmap_top{TOP_GENES_HEATMAP}.png")
        
        # ===========================================
        # PLOT 2: TIME COMPONENT LINE PLOT
        # ===========================================
        print(f"  Creating time-component line plot...")
        
        fig, ax = plt.subplots(figsize=(12, 8))
        
        # Plot each component
        for i, col in enumerate(time_factors.columns):
            component_weights = weights[weights['Component'] == col]['Weight'].values[0]
            ax.plot(time_factors.index, time_factors[col], 
                   marker='o', linewidth=2, markersize=8,
                   label=f'{col} (w={component_weights:.3f})', alpha=0.8)
        
        ax.set_xlabel('Timepoint', fontsize=12, fontweight='bold')
        ax.set_ylabel('Component Loading', fontsize=12, fontweight='bold')
        ax.set_title(f'Rank {rank}: Component Loadings Across Timepoints\n' +
                    '(Each line = one component, weighted by importance)',
                    fontsize=13, fontweight='bold', pad=15)
        ax.grid(True, alpha=0.3)
        ax.legend(bbox_to_anchor=(1.05, 1), loc='upper left', fontsize=9)
        ax.axhline(y=0, color='black', linestyle='-', linewidth=0.5, alpha=0.3)
        
        plt.tight_layout()
        output_path = os.path.join(rank_dir, 'time_component_lineplot.png')
        plt.savefig(output_path, dpi=150, bbox_inches='tight')
        plt.close()
        print(f"    Saved: time_component_lineplot.png")
        
        # ===========================================
        # PLOT 3: COMPONENT WEIGHT BAR PLOT
        # ===========================================
        print(f"  Creating component weight bar plot...")
        
        fig, ax = plt.subplots(figsize=(max(10, rank*0.4), 6))
        
        bars = ax.bar(range(len(weights)), weights['Weight'], 
                     color='steelblue', alpha=0.7, edgecolor='black')
        
        # Color top 5 components differently
        for i in range(min(5, len(bars))):
            bars[i].set_color('coral')
            bars[i].set_alpha(0.9)
        
        ax.set_xlabel('Component', fontsize=12, fontweight='bold')
        ax.set_ylabel('Weight', fontsize=12, fontweight='bold')
        ax.set_title(f'Rank {rank}: Component Weights\n(Top 5 highlighted in coral)',
                    fontsize=13, fontweight='bold', pad=15)
        ax.set_xticks(range(len(weights)))
        ax.set_xticklabels([f'C{i+1}' for i in range(len(weights))], rotation=45 if rank > 20 else 0)
        ax.grid(True, alpha=0.3, axis='y')
        
        plt.tight_layout()
        output_path = os.path.join(rank_dir, 'component_weights_barplot.png')
        plt.savefig(output_path, dpi=150, bbox_inches='tight')
        plt.close()
        print(f"    Saved: component_weights_barplot.png")
        
        # ===========================================
        # PLOT 4: GENE IMPORTANCE DISTRIBUTION
        # ===========================================
        print(f"  Creating gene importance distribution...")
        
        fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(14, 5))
        
        # Histogram
        ax1.hist(gene_importance, bins=50, color='steelblue', alpha=0.7, edgecolor='black')
        ax1.axvline(gene_importance.nlargest(TOP_GENES_HEATMAP).iloc[-1], 
                   color='red', linestyle='--', linewidth=2,
                   label=f'Top {TOP_GENES_HEATMAP} threshold')
        ax1.set_xlabel('Total Importance Score', fontsize=11)
        ax1.set_ylabel('Number of Genes', fontsize=11)
        ax1.set_title('Distribution of Gene Importance Scores', fontweight='bold')
        ax1.grid(True, alpha=0.3, axis='y')
        ax1.legend()
        
        # Cumulative curve
        sorted_importance = gene_importance.sort_values(ascending=False).values
        cumsum = np.cumsum(sorted_importance)
        cumsum_pct = (cumsum / cumsum[-1]) * 100
        
        ax2.plot(range(len(cumsum_pct)), cumsum_pct, linewidth=2, color='darkgreen')
        ax2.axhline(y=50, color='orange', linestyle='--', alpha=0.5, label='50% variance')
        ax2.axhline(y=80, color='red', linestyle='--', alpha=0.5, label='80% variance')
        ax2.set_xlabel('Number of Top Genes', fontsize=11)
        ax2.set_ylabel('Cumulative Variance Explained (%)', fontsize=11)
        ax2.set_title('Cumulative Importance by Gene Rank', fontweight='bold')
        ax2.grid(True, alpha=0.3)
        ax2.legend()
        
        fig.suptitle(f'Rank {rank}: Gene Importance Analysis', 
                    fontsize=14, fontweight='bold', y=1.02)
        plt.tight_layout()
        output_path = os.path.join(rank_dir, 'gene_importance_distribution.png')
        plt.savefig(output_path, dpi=150, bbox_inches='tight')
        plt.close()
        print(f"    Saved: gene_importance_distribution.png")
        
        # ===========================================
        # PLOT 5: COMPONENT CORRELATION HEATMAP
        # ===========================================
        print(f"  Creating component correlation heatmap...")
        
        # Calculate correlation between components based on gene loadings
        component_corr = gene_factors.corr()
        
        fig, ax = plt.subplots(figsize=(max(10, rank*0.4), max(8, rank*0.35)))
        
        sns.heatmap(
            component_corr,
            annot=True if rank <= 20 else False,
            fmt='.2f' if rank <= 20 else '.1f',
            cmap='coolwarm',
            center=0,
            square=True,
            cbar_kws={'label': 'Correlation'},
            ax=ax
        )
        
        ax.set_title(f'Rank {rank}: Component Correlation Matrix\n' +
                    '(Based on gene loadings - low correlation = good separation)',
                    fontsize=13, fontweight='bold', pad=15)
        
        plt.tight_layout()
        output_path = os.path.join(rank_dir, 'component_correlation_heatmap.png')
        plt.savefig(output_path, dpi=150, bbox_inches='tight')
        plt.close()
        print(f"    Saved: component_correlation_heatmap.png")
    
    print(f"\n{'='*60}")
    print("INDIVIDUAL RANK VISUALIZATIONS COMPLETE")
    print(f"{'='*60}")
    print(f"\nFor each rank, the following plots were created in rank_XX/ directories:")
    print(f"  1. gene_component_heatmap_top{TOP_GENES_HEATMAP}.png - Top genes across components")
    print(f"  2. time_component_lineplot.png - Component patterns over time")
    print(f"  3. component_weights_barplot.png - Relative component importance")
    print(f"  4. gene_importance_distribution.png - Gene importance statistics")
    print(f"  5. component_correlation_heatmap.png - Component independence check")
```

# RANK RECOMMENDATION

```{python recommend-rank, eval=TRUE}
# ===========================================
# This chunk reads from rank_comparison_log.json
# and does NOT require variables from previous chunks
# ===========================================
import os
import json
import pandas as pd
import numpy as np

print(f"\n{'='*60}")
print("RANK RECOMMENDATION")
print(f"{'='*60}")

# ===========================================
# LOAD RESULTS FROM LOG FILE IF NEEDED
# ===========================================
if 'output_dir' not in globals():
    print("Getting output_dir from R environment...")
    output_dir = r.output_dir
    print(f"  output_dir = {output_dir}")

if 'all_rank_results' not in globals():
    log_file = os.path.join(output_dir, 'rank_comparison_log.json')
    
    if os.path.exists(log_file):
        print(f"Loading results from log file: {log_file}")
        with open(log_file, 'r') as f:
            log_data = json.load(f)
        
        # Reconstruct all_rank_results from log
        all_rank_results = []
        for result_info in log_data['results']:
            if result_info['success']:
                all_rank_results.append({
                    'rank': result_info['rank'],
                    'metrics': result_info['metrics'],
                    'success': True
                })
            else:
                all_rank_results.append({
                    'rank': result_info['rank'],
                    'success': False,
                    'error': result_info.get('error', 'Unknown error')
                })
        
        print(f"  Loaded {len(all_rank_results)} results from log file")
    else:
        print("ERROR: No log file found and all_rank_results not in memory!")
        print("Please run the barnacle-multiple-ranks chunk first.")
        all_rank_results = []

successful = [r for r in all_rank_results if r['success']]
if len(successful) == 0:
    print("No successful decompositions to analyze!")
else:
    # Sort by rank
    successful = sorted(successful, key=lambda x: x['rank'])
    
    # Calculate scores for each rank
    scores = []
    for r in successful:
        m = r['metrics']
        
        # Variance explained score (0-1, higher is better)
        var_score = m['variance_explained']
        
        # Sparsity score (0-1, higher is better)
        sparsity_score = (m['gene_sparsity'] + m['sample_sparsity'] + m['time_sparsity']) / 3
        
        # Convergence score
        conv_score = 1.0 if m['converged'] else 0.5
        
        # Weight distribution score (prefer moderate variation)
        cv = m['weight_cv']
        if 0.5 <= cv <= 2.0:
            weight_score = 1.0
        elif cv < 0.5:
            weight_score = cv / 0.5
        else:
            weight_score = 2.0 / cv
        
        # Combined score (weighted average)
        combined_score = (
            0.40 * var_score +
            0.25 * sparsity_score +
            0.20 * conv_score +
            0.15 * weight_score
        )
        
        scores.append({
            'rank': r['rank'],
            'variance_score': var_score,
            'sparsity_score': sparsity_score,
            'convergence_score': conv_score,
            'weight_score': weight_score,
            'combined_score': combined_score
        })
    
    scores_df = pd.DataFrame(scores)
    
    # Find elbow in variance explained
    var_exp = [r['metrics']['variance_explained'] for r in successful]
    ranks = [r['rank'] for r in successful]
    
    elbow_rank = None
    if len(var_exp) > 2:
        marginal_gains = np.diff(var_exp)
        elbow_indices = np.where(marginal_gains < 0.05)[0]
        if len(elbow_indices) > 0:
            elbow_rank = ranks[elbow_indices[0]]
    
    # Best by combined score
    best_idx = scores_df['combined_score'].idxmax()
    recommended_rank = scores_df.iloc[best_idx]['rank']
    
    print("\nScoring Summary:")
    print(scores_df.to_string(index=False, float_format='%.3f'))
    
    print(f"\n{'='*60}")
    print(f"RECOMMENDED RANK: {int(recommended_rank)}")
    print(f"{'='*60}")
    print(f"  Combined score: {scores_df.iloc[best_idx]['combined_score']:.3f}")
    print(f"  Variance explained: {successful[best_idx]['metrics']['variance_explained']*100:.1f}%")
    print(f"  Gene sparsity: {successful[best_idx]['metrics']['gene_sparsity']*100:.1f}%")
    print(f"  Converged: {successful[best_idx]['metrics']['converged']}")
    
    if elbow_rank is not None:
        print(f"\nElbow point detected at rank: {int(elbow_rank)}")
        print("  (Diminishing returns beyond this point)")
    
    print("\nReasoning:")
    if recommended_rank <= min(ranks):
        print("  → Lower rank preferred: Model may be overfitting or data is simple")
    elif recommended_rank >= max(ranks):
        print("  → Higher rank preferred: Complex data may need more components")
        print("  → Consider testing even higher ranks (e.g., +5, +10)")
    else:
        print("  → Middle rank optimal: Good balance of fit and interpretability")
    
    print("\nNext steps:")
    print(f"  1. Review plots in: {comparison_dir}")
    print(f"  2. Examine individual rank results in: {output_dir}/rank_XX/")
    print(f"  3. Verify biological interpretability of top components")
    print(f"  4. Re-run main analysis with chosen rank for final results")
```




## Close log file

```{python close-log, eval=TRUE}
from datetime import datetime

print()
print("="*60)
print(f"ANALYSIS COMPLETE")
print(f"Finished: {datetime.now().strftime('%Y-%m-%d %H:%M:%S')}")
print(f"All output saved to: {output_dir}")
print(f"Log file: {log_file}")
print("="*60)

# Restore stdout and close log file
if hasattr(sys.stdout, 'log'):
    sys.stdout.log.close()
    sys.stdout = sys.stdout.terminal
```

# SYSTEM INFO

```{r system-info, eval=TRUE}
sessionInfo()
```

# REFERENCES
