---
title: "35.1-physiology-gene-sets-counts"
author: "Kathleen Durkin"
date: "2026-06-16"
output: 
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
---

Now that I have domain-specific gene sets for each physiological domain, across the three species, I need to produce some filtered count matrices for those genes' expression

```{r}
library(dplyr)
library(tidyr)
```


Read in full count matrices
```{r}
apul_counts <- read.csv("../../D-Apul/output/02.20-D-Apul-RNAseq-alignment-HiSat2/apul-gene_count_matrix.csv")
peve_counts <- read.csv("../../E-Peve/output/02.20-E-Peve-RNAseq-alignment-HiSat2/peve-gene_count_matrix.csv")
ptua_counts <- read.csv("../../F-Ptua/output/02.20-F-Ptua-RNAseq-alignment-HiSat2/ptua-gene_count_matrix.csv")
```

Read in gene sets
```{r}
ant_genes <- read.csv("../output/35-physiology-gene-sets/antioxidant_geneset_expanded.csv")
grow_genes <- read.csv("../output/35-physiology-gene-sets/growth_geneset_expanded.csv")
resp_genes <- read.csv("../output/35-physiology-gene-sets/respiration_geneset_expanded.csv")
sym_genes <- read.csv("../output/35-physiology-gene-sets/symbiosis_geneset_expanded.csv")

# Remove -TXX suffix on Apul gene names
ant_genes$apul <- sub("-.*", "", ant_genes$apul)
grow_genes$apul <- sub("-.*", "", grow_genes$apul)
resp_genes$apul <- sub("-.*", "", resp_genes$apul)
sym_genes$apul <- sub("-.*", "", sym_genes$apul)

# Add gene- prefix on Peve gene names
ant_genes$peve <- paste0("gene-", ant_genes$peve)
grow_genes$peve <- paste0("gene-", grow_genes$peve)
resp_genes$peve <- paste0("gene-", resp_genes$peve)
sym_genes$peve <- paste0("gene-", sym_genes$peve)

# Add gene- prefix on Peve gene names
ant_genes$ptua <- paste0("gene-", ant_genes$ptua)
grow_genes$ptua <- paste0("gene-", grow_genes$ptua)
resp_genes$ptua <- paste0("gene-", resp_genes$ptua)
sym_genes$ptua <- paste0("gene-", sym_genes$ptua)
```

Join full counts with the domain set (want to retain both the ortholog group ID and the species gene ID)
```{r}
# Apul
apul_counts_ant <- left_join(ant_genes, apul_counts, by = c("apul" = "gene_id")) %>% rename(gene_id = apul) %>% select(group_id, gene_id, starts_with("ACR")) %>% drop_na()
apul_counts_grow <- left_join(grow_genes, apul_counts, by = c("apul" = "gene_id")) %>% rename(gene_id = apul) %>% select(group_id, gene_id, starts_with("ACR")) %>% drop_na()
apul_counts_resp <- left_join(resp_genes, apul_counts, by = c("apul" = "gene_id")) %>% rename(gene_id = apul) %>% select(group_id, gene_id, starts_with("ACR")) %>% drop_na()
apul_counts_sym <- left_join(sym_genes, apul_counts, by = c("apul" = "gene_id")) %>% rename(gene_id = apul) %>% select(group_id, gene_id, starts_with("ACR")) %>% drop_na()

# Peve
peve_counts_ant <- left_join(ant_genes, peve_counts, by = c("peve" = "gene_id")) %>% rename(gene_id = peve) %>% select(group_id, gene_id, starts_with("POR")) %>% drop_na()
peve_counts_grow <- left_join(grow_genes, peve_counts, by = c("peve" = "gene_id")) %>% rename(gene_id = peve) %>% select(group_id, gene_id, starts_with("POR")) %>% drop_na()
peve_counts_resp <- left_join(resp_genes, peve_counts, by = c("peve" = "gene_id")) %>% rename(gene_id = peve) %>% select(group_id, gene_id, starts_with("POR")) %>% drop_na()
peve_counts_sym <- left_join(sym_genes, peve_counts, by = c("peve" = "gene_id")) %>% rename(gene_id = peve) %>% select(group_id, gene_id, starts_with("POR")) %>% drop_na()

# Ptua
ptua_counts_ant <- left_join(ant_genes, ptua_counts, by = c("ptua" = "gene_id")) %>% rename(gene_id = ptua) %>% select(group_id, gene_id, starts_with("POC")) %>% drop_na()
ptua_counts_grow <- left_join(grow_genes, ptua_counts, by = c("ptua" = "gene_id")) %>% rename(gene_id = ptua) %>% select(group_id, gene_id, starts_with("POC")) %>% drop_na()
ptua_counts_resp <- left_join(resp_genes, ptua_counts, by = c("ptua" = "gene_id")) %>% rename(gene_id = ptua) %>% select(group_id, gene_id, starts_with("POC")) %>% drop_na()
ptua_counts_sym <- left_join(sym_genes, ptua_counts, by = c("ptua" = "gene_id")) %>% rename(gene_id = ptua) %>% select(group_id, gene_id, starts_with("POC")) %>% drop_na()
```


Save
```{r}
# Apul
write.csv(apul_counts_ant, "../output/35.1-physiology-gene-sets-counts/apul-antioxidant-counts.csv", row.names = FALSE, quote = FALSE)
write.csv(apul_counts_grow, "../output/35.1-physiology-gene-sets-counts/apul-growth-counts.csv", row.names = FALSE, quote = FALSE)
write.csv(apul_counts_resp, "../output/35.1-physiology-gene-sets-counts/apul-respiration-counts.csv", row.names = FALSE, quote = FALSE)
write.csv(apul_counts_sym, "../output/35.1-physiology-gene-sets-counts/apul-symbiosis-counts.csv", row.names = FALSE, quote = FALSE)

# Peve
write.csv(peve_counts_ant, "../output/35.1-physiology-gene-sets-counts/peve-antioxidant-counts.csv", row.names = FALSE, quote = FALSE)
write.csv(peve_counts_grow, "../output/35.1-physiology-gene-sets-counts/peve-growth-counts.csv", row.names = FALSE, quote = FALSE)
write.csv(peve_counts_resp, "../output/35.1-physiology-gene-sets-counts/peve-respiration-counts.csv", row.names = FALSE, quote = FALSE)
write.csv(peve_counts_sym, "../output/35.1-physiology-gene-sets-counts/peve-symbiosis-counts.csv", row.names = FALSE, quote = FALSE)

# Ptua
write.csv(ptua_counts_ant, "../output/35.1-physiology-gene-sets-counts/ptua-antioxidant-counts.csv", row.names = FALSE, quote = FALSE)
write.csv(ptua_counts_grow, "../output/35.1-physiology-gene-sets-counts/ptua-growth-counts.csv", row.names = FALSE, quote = FALSE)
write.csv(ptua_counts_resp, "../output/35.1-physiology-gene-sets-counts/ptua-respiration-counts.csv", row.names = FALSE, quote = FALSE)
write.csv(ptua_counts_sym, "../output/35.1-physiology-gene-sets-counts/ptua-symbiosis-counts.csv", row.names = FALSE, quote = FALSE)
```

Note that there may be a slight difference in # of rows, since some of the identified domain-specific genes may be very lowly or un-expressed
