---
title: "01.00-D-Apul-RNAseq-trimming-fastp-FastQC-MultiQC"
author: "Sam White"
date: "2024-10-05"
output: 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

------------------------------------------------------------------------

This Rmd file trims *A.pulchra* RNA-seq files using [fastp](https://github.com/OpenGene/fastp) [@chen2023], followed by quality checks with [FastQC](https://github.com/s-andrews/FastQC) and [MultiQC](https://multiqc.info/)[@ewels2016].

Expects input files formatted like so: `<number>--<sample_name>_R[12]_001.fastq.gz`

# Create a Bash variables file

This allows usage of Bash variables across R Markdown chunks.

```{r save-bash-variables-to-rvars-file, engine='bash', eval=TRUE}
{
echo "#### Assign Variables ####"
echo ""

echo "# Data directories"
echo 'export timeseries_dir=/home/shared/8TB_HDD_01/sam/gitrepos/urol-e5/timeseries_molecular'
echo 'export output_dir_top=${timeseries_dir}/D-Apul/output/01.00-D-Apul-RNAseq-trimming-fastp-FastQC-MultiQC'
echo 'export raw_reads_dir=${timeseries_dir}/D-Apul/data/raw-fastqs'
echo 'export raw_reads_url="https://owl.fish.washington.edu/nightingales/E5-coral-time-series/30-1047560508/"'
echo 'export trimmed_fastqs_dir=${output_dir_top}/trimmed-fastqs'
echo 'export trimmed_fastqc_dir=${output_dir_top}/trimmed-fastqc'
echo ""

echo "# Paths to programs"
echo 'export programs_dir="/home/shared"'
echo 'export fastp="${programs_dir}/fastp"'
echo 'export fastqc=${programs_dir}/FastQC-0.12.1/fastqc'
echo 'export multiqc=/home/sam/programs/mambaforge/bin/multiqc'
echo ""

echo "# Set FastQ filename patterns"
echo "export fastq_pattern='*.fastq.gz'"
echo "export R1_fastq_pattern='*_R1_*.fastq.gz'"
echo "export R2_fastq_pattern='*_R2_*.fastq.gz'"
echo ""

echo "# Set number of CPUs to use"
echo 'export threads=40'
echo ""


echo "## Inititalize arrays"
echo 'export fastq_array_R1=()'
echo 'export fastq_array_R2=()'
echo 'export raw_fastqs_array=()'
echo 'export R1_names_array=()'
echo 'export R2_names_array=()'
echo ""

echo "# Programs associative array"
echo "declare -A programs_array"
echo "programs_array=("
echo '[fastp]="${fastp}" \'
echo '[fastqc]="${fastqc}" \'
echo '[multiqc]="${multiqc}" \'
echo ")"
echo ""

echo "# Print formatting"
echo 'export line="--------------------------------------------------------"'
echo ""
} > .bashvars

cat .bashvars
```

If needed, download raw RNA-seq.

Change `eval=FALSE` to `eval=TRUE` to execute the next two chunks to download RNA-seq and then verify MD5 checksums.

```{bash download-raw-reads, engine='bash', eval=FALSE}
# Load bash variables into memory
source .bashvars

# Make output directory if it doesn't exist
mkdir --parents ${raw_reads_dir}

# Create list of only A.pulchra sample names
sample_list=$(awk -F "," 'NR > 2 {print $5"\t"$6}' ${timeseries_dir}/data/rna_metadata.csv | awk -F"[\t-]" '$2 == "ACR" {print $1}')

echo ""
echo "${line}"
echo ""
echo "Sample list:"
echo ""
echo "${sample_list}"
echo ""
echo "${line}"
echo ""


# Use printf to format each item for use in wget
formatted_list=$(printf "*%s*," ${sample_list})

# Remove the trailing comma and append *.md5
formatted_list="${formatted_list%,},*.md5"

# Output the final wget command
echo ""
echo "${line}"
echo ""
echo "Formatted wget accept list:"
echo ""
echo "wget --accept=\"$formatted_list\""
echo ""
echo "${line}"
echo ""

# Run wget to retrieve FastQs and MD5 files
wget \
--directory-prefix ${raw_reads_dir} \
--recursive \
--no-check-certificate \
--continue \
--cut-dirs 3 \
--no-host-directories \
--no-parent \
--quiet \
--accept=\"$formatted_list\" ${raw_reads_url}

ls -lh "${raw_reads_dir}"
```

Verify raw read checksums
```{bash verify-raw-read-checksums, engine='bash', eval=FALSE}
# Load bash variables into memory
source .bashvars

cd "${raw_reads_dir}"

# Checksums file contains other files, so this just looks for the sRNAseq files.
for file in *.md5
do
  md5sum --check "${file}"
done
```

# Fastp Trimming

```{bash fastp-trimming, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

# Make output directories, if it doesn't exist
mkdir --parents "${trimmed_fastqs_dir}"

# Change to raw reads directory
cd "${raw_reads_dir}"

# Create arrays of fastq R1 files and sample names
for fastq in ${R1_fastq_pattern}
do
  fastq_array_R1+=("${fastq}")
  R1_names_array+=("$(echo "${fastq}" | awk -F"_" '{print $1}')")
done

# Create array of fastq R2 files
for fastq in ${R2_fastq_pattern}
do
  fastq_array_R2+=("${fastq}")
  R2_names_array+=("$(echo "${fastq}" | awk -F"_" '{print $1}')")
done

# Create list of fastq files used in analysis
# Create MD5 checksum for reference
if [ ! -f "${output_dir_top}"/raw-fastq-checksums.md5 ]; then
for fastq in *.gz
  do
    md5sum ${fastq} >> "${output_dir_top}"/raw-fastq-checksums.md5
  done
fi

# Run fastp on files
# Adds JSON report output for downstream usage by MultiQC
for index in "${!fastq_array_R1[@]}"
do
  R1_sample_name=$(echo "${R1_names_array[index]}")
  R2_sample_name=$(echo "${R2_names_array[index]}")
  ${fastp} \
  --in1 ${fastq_array_R1[index]} \
  --in2 ${fastq_array_R2[index]} \
  --detect_adapter_for_pe \
  --thread ${threads} \
  --html "${trimmed_fastqs_dir}"/"${R1_sample_name}".fastp-trim.report.html \
  --json "${trimmed_fastqs_dir}"/"${R1_sample_name}".fastp-trim.report.json \
  --out1 "${trimmed_fastqs_dir}"/"${R1_sample_name}"_R1_001.fastp-trim.fq.gz \
  --out2 "${trimmed_fastqs_dir}"/"${R2_sample_name}"_R2_001.fastp-trim.fq.gz \
  2>> "${trimmed_fastqs_dir}"/fastp.stderr

  # Generate md5 checksums for newly trimmed files
  md5sum "${R1_sample_name}"_R1_001.fastp-trim.fq.gz > "${R1_sample_name}"_R1_001.fastp-trim.fq.gz.md5
  md5sum "${R2_sample_name}"_R2_001.fastp-trim.fq.gz > "${R2_sample_name}"_R2_001.fastp-trim.fq.gz.md5
done
```