---
title: "03.00-D-Apul-RNAseq-gene-expression-DESeq2"
author: "Sam White"
date: "2024-10-15"
output: 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
---

```{r setup, include=FALSE}
library(DESeq2)
library(ggplot2)
library(knitr)
library(pheatmap)
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Set Bash variables
```{r save-bash-variables-to-rvars-file, engine='bash', eval=TRUE}
{
echo "#### Assign Variables ####"
echo ""

echo "# Data directories"
echo 'export timeseries_dir=/home/shared/8TB_HDD_01/sam/gitrepos/urol-e5/timeseries_molecular'
echo 'export output_dir_top="${timeseries_dir}/D-Apul/output/03.00-D-Apul-RNAseq-gene-expression-DESeq2"'
echo ""



echo "# Output files"
echo 'export coldata="${output_dir_top}/DESeq2-coldata.tab"'
echo ""

echo "# Paths to programs"
echo 'export programs_dir="/home/shared"'


echo "# Set number of CPUs to use"
echo 'export threads=40'
echo ""


echo "# Print formatting"
echo 'export line="--------------------------------------------------------"'
echo ""
} > .bashvars

cat .bashvars
```
# Set R variables
```{r set-variables}
# Define the output directory path
output_dir <- "../output/03.00-D-Apul-RNAseq-gene-expression-DESeq2/"

# Set desired false discovery rate threshold (i.e. adjusted p-value, padj)
fdr <- 0.05

# Set log2 fold change threshold (a value of '1' is equal to a fold change of '2')
log2fc <- 1
```


# Create coldat file
```{r create-coldata-file, engine='bash', eval=FALSE}
# Load bash variables into memory
source .bashvars

# Create output directory, if it doesn't exist
mkdir --parents "${output_dir_top}"

# Create associative array with sample and timepoint
metadata="../../M-multi-species/data/rna_metadata.csv"

# Declare the array
declare -A sample_timepoint_map

# Read the metadata file line by line
while IFS=',' read -r sample_number sample_name plate well_number azenta_sample_name colony_id timepoint sample_type species_strain SampleBuffer; do
    # Check if the species is "Acropora pulchra"
    if [[ "${species_strain}" == "Acropora pulchra" ]]; then
        # Add the Azenta sample name as the key and Timepoint as the value in the associative array
        sample_timepoint_map["${azenta_sample_name}"]="${timepoint}"
    fi
done < <(tail -n +2 "${metadata}") # Skip the header

# Create DESeq2-formatted coldata file

## Create header
printf "\t%s\n" "condition" > "${coldata}"

## Tab-delimited output of sample and timepoint
for sample in "${!sample_timepoint_map[@]}"
do
  printf "%s\t%s\n" "$sample" "${sample_timepoint_map[$sample]}"
done | sort -k1,1 \
>> "${coldata}"

# Peek at output
head "${coldata}"

echo ""
echo "${line}"
echo ""

wc -l "${coldata}"
```

# Read in gene counts and coldata files
```{r read-gene-counts}
cts <- as.matrix(read.csv(file = "../output/02.20-D-Apul-RNAseq-alignment-HiSat2/apul-gene_count_matrix.csv", row.names="gene_id", check.names = FALSE))

coldata <- read.csv(file = "../output/03.00-D-Apul-RNAseq-gene-expression-DESeq2/DESeq2-coldata.tab", row.names=1, sep = "\t")
coldata$condition <- factor(coldata$condition)

head(cts)

head(coldata)
```

## Verify rownames match
```{r check-rownames}
all(rownames(coldata) == colnames(cts))
```

# Create DESeq2 data set
```{r create-deseq2-data-set}
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ condition)
dds

```

## Add gene columun feature
```{r add-gene-feature}
featureData <- data.frame(gene=rownames(cts))
mcols(dds) <- DataFrame(mcols(dds), featureData)
mcols(dds)
```

```{r level}
dds$condition <- factor(dds$condition, levels = c("TP1","TP2", "TP3", "TP4"))
```

# DESeq analysis
```{r deseq}
dds <- DESeq(dds)

```

## Pairwise results tables

### Full 
```{r deseq2-pairwise-results-tables}
tp1.v.tp2.results <- results(dds, contrast=c("condition","TP1","TP2"), alpha = fdr, lfcThreshold = log2fc)
tp1.v.tp3.results <- results(dds, contrast=c("condition","TP1","TP3"), alpha = fdr, lfcThreshold = log2fc)
tp1.v.tp4.results <- results(dds, contrast=c("condition","TP1","TP4"), alpha = fdr, lfcThreshold = log2fc)
tp2.v.tp3.results <- results(dds, contrast=c("condition","TP2","TP3"), alpha = fdr, lfcThreshold = log2fc)
tp2.v.tp4.results <- results(dds, contrast=c("condition","TP2","TP4"), alpha = fdr, lfcThreshold = log2fc)
tp3.v.tp4.results <- results(dds, contrast=c("condition","TP3","TP4"), alpha = fdr, lfcThreshold = log2fc)

tp1.v.tp2.results

summary(tp1.v.tp2.results)

table(tp1.v.tp2.results$padj < 0.05)
```

# Write DDS results tables to CSVs
```{r write-dds-results-csv}
# Create a named list of the data frames
results_list <- list(
  tp1.v.tp2.results = tp1.v.tp2.results,
  tp1.v.tp3.results = tp1.v.tp3.results,
  tp1.v.tp4.results = tp1.v.tp4.results,
  tp2.v.tp3.results = tp2.v.tp3.results,
  tp2.v.tp4.results = tp2.v.tp4.results,
  tp3.v.tp4.results = tp3.v.tp4.results
)

# Loop through the list and write each data frame to a CSV file in the specified directory
for (df_name in names(results_list)) {
  write.csv(results_list[[df_name]], file = paste0(output_dir, df_name, ".table.csv"), row.names = FALSE, quote = FALSE)
}
```

# Variance stabilizing transformations (VST)
- Here we transform counts using a variance stabilizing transformation (VST), since we have many samples. 
```{r VST}
vsd <- varianceStabilizingTransformation(dds, blind=FALSE)
```


NOTE: Hover over points to see the sample numbers

# Plotting

## PCA - All time points
Visualize sample clustering via PCA (after transformation)
```{r pca-all-timepoints}
# PCA with points color coded by time point 
plotPCA(vsd)
```

## Heatmap - All time points
```{r heatmap-all-timepoints-top20}
top_20_counts_all <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]

timepoint_annotation = colData(dds) %>% as.data.frame() %>% select(condition)




pheatmap(assay(vsd)[top_20_counts_all,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col = timepoint_annotation)

```
