---
title: "07-Apul-miRNA-mRNA-miRanda"
author: "Kathleen Durkin"
date: "2024-12-19"
always_allow_html: true
output: 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true 
---

miRanda is a target prediction software, used to identify likely miRNA-mRNA interactions.

Inputs:

  - FASTA of A.pulchra 3'UTRs `Apul_3UTR_1kb.fasta`, generated in `05-Apul-annotate-UTRs`
  
  - FASTA of A.pulchra mature miRNAs `miRNA_mature-Apul.fasta`. miRNAs identified in `04-Apul-sRNA-discovery-ShortStack`, matures isolated for use in `06-Apul-miRNA-mRNA-RNAhybrid`

Outputs:






```{r, engine='bash', eval=FALSE}

# score cutoff >100
# energy cutoff <-10
# strict binding

/home/shared/miRanda-3.3a/src/miranda \
../data/06-Apul-miRNA-mRNA-RNAhybrid/miRNA_mature-Apul.fasta \
../output/05-Apul-annotate-UTRs/Apul_3UTR_1kb.fasta \
-sc 100 \
-en -10 \
-strict \
-out ../output/07-Apul-miRNA-mRNA-miRanda/Apul-miRanda-3UTR-strict_all.tab
```


Let's look at the output

```{r, engine='bash'}

echo "miranda run finished!"
echo "counting number of putative interactions predicted"

zgrep -c "Performing Scan" ../output/07-Apul-miRNA-mRNA-miRanda/Apul-miRanda-3UTR-strict_all.tab

echo "Parsing output"
grep -A 1 "Scores for this hit:" ../output/07-Apul-miRNA-mRNA-miRanda/Apul-miRanda-3UTR-strict_all.tab | sort | grep '>' > ../output/07-Apul-miRNA-mRNA-miRanda/Apul-miRanda-3UTR-strict-parsed.txt

echo "counting number of putative interactions predicted"
wc -l ../output/07-Apul-miRNA-mRNA-miRanda/Apul-miRanda-3UTR-strict_all.tab

```

