---
title: "Apul WGCNA"
output: html_document
date: "2024-12-23"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

To do: 

- Troubleshoot correlations with Timepoint and site as categories 
- See if we can run correlations in which some samples have NA with traits (e.g., cre)
- Look at physiology outliers 

# Install and load packages

```{r load_libraries, inlcude = TRUE}
library(tidyverse)
library(ggplot2)
library(WGCNA)
library(magrittr)
library(genefilter)
```

# Load data

Load in count matrices for RNAseq. 

```{r load-RNA-counts}
# RNA variance stabilized counts data
genes <- read_csv("D-Apul/output/02.20-D-Apul-RNAseq-alignment-HiSat2/apul-gene_count_matrix.csv")

genes<-as.data.frame(genes)

rownames(genes)<-genes$gene_id

genes<-genes%>%select(!gene_id)

metadata<-read_csv("M-multi-species/data/rna_metadata.csv")%>%select(AzentaSampleName, ColonyID, Timepoint)%>%
  filter(grepl("ACR", ColonyID))

colonies<-unique(metadata$ColonyID)

phys<-read_csv("https://github.com/urol-e5/timeseries/raw/refs/heads/master/time_series_analysis/Output/master_timeseries.csv")%>%filter(colony_id_corr %in% colonies)%>%
  select(colony_id_corr, species, timepoint, site, Host_AFDW.mg.cm2, Sym_AFDW.mg.cm2, Am, AQY, Rd, Ik, Ic, calc.umol.cm2.hr, cells.mgAFDW, prot_mg.mgafdw, Ratio_AFDW.mg.cm2, Total_Chl, Total_Chl_cell, cre.umol.mgafdw)
```

Check that there are no genes with 0 counts across all samples.

```{r}
nrow(genes)

genes<-genes %>%
     mutate(Total = rowSums(.[, 1:40]))%>%
    filter(!Total==0)%>%
    dplyr::select(!Total)

nrow(genes)
```

# pOverA filtering

*pOverA*: Specifying the minimum count for a proportion of samples for each gene. Here, we are using a pOverA of 0.25. This is because we have 40 samples with a minimum of n=10 samples per timepoint Therefore, we will accept genes that are present in 10/40 = 0.25 of the samples because we expect different expression by life stage. We are further setting the minimum count of genes to 10, such that 25% of the samples must have a gene count of >10 in order for the gene to remain in the data set.  

Filter in the package "genefilter". Pre-filtering our dataset to reduce the memory size dataframe, increase the speed of the transformation and testing functions, and improve quality of statistical analysis by removing low-coverage counts. Removed counts could represent outliers in the data and removing these improves sensitivity of statistical tests.   

```{r}
filt <- filterfun(pOverA(0.25,10))

#create filter for the counts data
gfilt <- genefilter(genes, filt)

#identify genes to keep by count filter
gkeep <- genes[gfilt,]

#identify genes to keep by count filter
gkeep <- genes[gfilt,]

#identify gene lists
gn.keep <- rownames(gkeep)

#gene count data filtered in PoverA, P percent of the samples have counts over A
genes_filt <- as.data.frame(genes[which(rownames(genes) %in% gn.keep),])

#How many rows do we have before and after filtering?
nrow(genes) #Before
nrow(genes_filt) #After
```

# Assign metadata and arrange order of columns 

Display current order of metadata and gene count matrix.  
```{r}
metadata$AzentaSampleName
colnames(genes_filt)
```

Order metadata the same as the column order in the gene matrix.  
```{r}
list<-colnames(genes_filt)
list<-as.factor(list)

metadata$AzentaSampleName<-as.factor(metadata$AzentaSampleName)

# Re-order the levels
metadata$AzentaSampleName <- factor(as.character(metadata$AzentaSampleName), levels=list)
# Re-order the data.frame
metadata_ordered <- metadata[order(metadata$AzentaSampleName),]
metadata_ordered$AzentaSampleName
```

Metadata and gene count matrix are now ordered the same. 

# Conduct variance stabilized transformation 

```{r}
library(DESeq2)
#Set DESeq2 design
gdds <- DESeqDataSetFromMatrix(countData = genes_filt,
                              colData = metadata_ordered,
                              design = ~Timepoint)
```

Check size factors. 
```{r}
SF.gdds <- estimateSizeFactors(gdds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.gdds)) #View size factors

all(sizeFactors(SF.gdds)) < 4
```

All size factors are less than 4, so we can use VST transformation.  

```{r}
gvst <- vst(gdds, blind=FALSE) #apply a variance stabilizing transformation to minimize effects of small counts and normalize wrt library size
head(assay(gvst), 3) #view transformed gene count data for the first three genes in the dataset.  
```

# Conduct permanova and pca 

Export data for PERMANOVA test.  
```{r}
test<-t(assay(gvst)) #export as matrix
test<-as.data.frame(test)

#add category columns
test$Sample<-rownames(test)
test$Timepoint<-metadata$Timepoint[match(test$Sample, metadata$AzentaSampleName)]
```

Build PERMANOVA model.  
```{r}
library(factoextra)
library(vegan)

scaled_test <-prcomp(test%>%select(where(is.numeric)), scale=TRUE, center=TRUE)
fviz_eig(scaled_test)

# scale data
vegan <- scale(test%>%select(where(is.numeric)))

# PerMANOVA 
permanova<-adonis2(vegan ~ Timepoint, data = test, method='eu')
permanova
```

```{r}
library(ggplot2)
library(ggfortify)
library(RColorBrewer)
library(ComplexHeatmap)

pca1<-ggplot2::autoplot(scaled_test, data=test, frame.colour="Timepoint", loadings=FALSE,  colour="Timepoint", loadings.label.colour="black", loadings.colour="black", loadings.label=FALSE, frame=FALSE, loadings.label.size=5, loadings.label.vjust=-1, size=5) + 
  geom_text(aes(x = PC1, y = PC2, label = Sample), vjust = -0.5)+
  theme_classic()+
   theme(legend.text = element_text(size=18), 
         legend.position="right",
        plot.background = element_blank(),
        legend.title = element_text(size=18, face="bold"), 
        axis.text = element_text(size=18), 
        axis.title = element_text(size=18,  face="bold"));pca1
```

Look at sample distances. 
```{r}
gsampleDists <- dist(t(assay(gvst))) #calculate distance matix
gsampleDistMatrix <- as.matrix(gsampleDists) #distance matrix
rownames(gsampleDistMatrix) <- colnames(gvst) #assign row names
colnames(gsampleDistMatrix) <- NULL #assign col names
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255) #assign colors

save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
   stopifnot(!missing(x))
   stopifnot(!missing(filename))
   pdf(filename, width=width, height=height)
   grid::grid.newpage()
   grid::grid.draw(x$gtable)
   dev.off()
}

pht<-pheatmap(gsampleDistMatrix, #plot matrix
         clustering_distance_rows=gsampleDists, #cluster rows
         clustering_distance_cols=gsampleDists, #cluster columns
         col=colors) #set colors

pdf(file="D-Apul/output/11.00-Apul-WGCNA/pheatmap.pdf")
print(pht)
dev.off()
```

# WGCNA

Look for outliers by examining tree of samples  
```{r}
datExpr <- as.data.frame(t(assay(gvst)))

sampleTree = hclust(dist(datExpr), method = "average");
# Plot the sample tree: Open a graphic output window of size 12 by 9 inches
# The user should change the dimensions if the window is too large or too small.
pdf("D-Apul/output/11.00-Apul-WGCNA/outliers.pdf")
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5, cex.axis = 1.5, cex.main = 2)
dev.off()
```

Soft threshold selection. The soft thresholding power (β) is the number to which the co-expression similarity is raised to calculate adjacency. The function pickSoftThreshold performs a network topology analysis. The user chooses a set of candidate powers, however the default parameters are suitable values.  
```{r, message=FALSE, warning=FALSE}
allowWGCNAThreads()
# # Choose a set of soft-thresholding powers
powers <- c(seq(from = 1, to=19, by=2), c(21:30)) #Create a string of numbers from 1 through 10, and even numbers from 10 through 20
# 
# # Call the network topology analysis function
sft <-pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
```

Plot the results.  
```{r}
sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;
# # # Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
      xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
 text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# # # this line corresponds to using an R^2 cut-off
 abline(h=0.9,col="red")
# # # Mean connectivity as a function of the soft-thresholding power
 plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
 text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```
I used a scale-free topology fit index **R^2 of 0.9**. This lowest recommended R^2 by Langfelder and Horvath is 0.8. I chose 0.9 because we want to use the smallest soft thresholding power that maximizes with model fit. It appears that our **soft thresholding power is 5** because it is the lowest power above the R^2=0.9 threshold that maximizes with model fit.  

Co-expression similarity and adjacency, using the soft thresholding power 5. I will use a **signed network**.  

Now we're ready to run WGCNA!

```{r, cache=TRUE}
picked_power = 5
temp_cor <- cor       
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
netwk_Apul <- blockwiseModules(datExpr,                # <= input here

                          # == Adjacency Function ==
                          power = picked_power,                # <= power here
                          networkType = "signed",

                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 5000,

                          # == Module Adjustments ==
                          reassignThreshold = 1e-6,
                          mergeCutHeight = 0.15,

                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = F,
                          saveTOMFileBase = "ER",

                          # == Output Options
                          numericLabels = T,
                          verbose = 3)


cor <- temp_cor     # Return cor function to original namespace

```

Take a look at dendrogram. 
```{r}
# Convert labels to colors for plotting
mergedColors = labels2colors(netwk_Apul$colors)
labels = table(netwk_Apul$colors)
labels

# Plot the dendrogram and the module colors underneath

plotDendroAndColors(
  netwk_Apul$dendrograms[[1]],
  mergedColors[netwk_Apul$blockGenes[[1]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )
```

1635 genes are unassigned.  

Show module number before and after merging. 
```{r}
length(table(netwk_Apul$unmergedColors))
length(table(netwk_Apul$colors))
```

There are 57 modules before merging to 85% similarity and 54 after. 

```{r}
MEs<-netwk_Apul$MEs
moduleLabels<-netwk_Apul$colors
```

# Correlate to traits 

Remove sample from expression data that is not present in physiological data. 1B9
```{r}
phys<-phys%>%mutate(Timepoint=if_else(timepoint=="timepoint1","TP1", 
                        if_else(timepoint=="timepoint2", "TP2", 
                                if_else(timepoint=="timepoint3", "TP3", 
                                  if_else(timepoint=="timepoint4", "TP4", NA)))))

phys$Timepoint

phys$Code <- paste0(phys$colony_id_corr, "_", phys$Timepoint)

metadata_ordered$Code<-paste0(metadata_ordered$ColonyID, "_", metadata_ordered$Timepoint)

setdiff(metadata_ordered$Code, phys$Code)

metadata_ordered%>%filter(Code=="ACR-265_TP4")

datExpr<-datExpr%>%filter(!rownames(.)=="1B9")

setdiff(rownames(datExpr), metadata_ordered$AzentaSampleName)
```

Create a physiological/factor dataset with sample in rows and phys data in columns (row name = Azenta sample name)
```{r}
phys$AzentaSampleName<-metadata_ordered$AzentaSampleName[match(phys$Code, metadata_ordered$Code)]

traits<-phys%>%
  select(AzentaSampleName, Timepoint, site, Host_AFDW.mg.cm2, Sym_AFDW.mg.cm2, Am, Rd, Ik, Ic, AQY, cells.mgAFDW, Total_Chl, Ratio_AFDW.mg.cm2, Total_Chl_cell)

#time point
#site
traits<-traits%>%
  mutate(site=if_else(site=="Mahana", "1", 
                      if_else(site=="Manava", "2", NA)))%>%
  mutate(Timepoint=gsub("TP", "", Timepoint))

traits$Timepoint<-as.numeric(traits$Timepoint)
traits$site<-as.numeric(traits$site)

traits<-as.data.frame(traits)
rownames(traits)<-traits$AzentaSampleName

datTraits<-traits%>%select(!AzentaSampleName)

datTraits<-datTraits%>%select(!Timepoint)%>%select(!site)
```

```{r}
nGenes = ncol(datExpr)
nSamples = nrow(datExpr)

nGenes
nSamples
```

Generate labels for module eigengenes as numbers.  
```{r}
MEs0 = moduleEigengenes(datExpr, moduleLabels, softPower=5)$eigengenes
MEs = orderMEs(MEs0)
names(MEs)
```

Correlations of traits with eigengenes
```{r}
moduleTraitCor = cor(MEs, datTraits, use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);
Colors=sub("ME","", names(MEs))

moduleTraitTree = hclust(dist(t(moduleTraitCor)), method = "average")

pdf(file="D-Apul/output/11.00-Apul-WGCNA/ModuleTraitClusterTree.pdf")
plot(moduleTraitTree)
dev.off()
```

Correlations of genes with eigengenes. Calculate correlations between ME's and lifestages. 
```{r}
moduleGeneCor=cor(MEs,datExpr)
moduleGenePvalue = corPvalueStudent(moduleGeneCor, nSamples);
```

Calculate kME values (module membership). 

```{r}
datKME = signedKME(datExpr, MEs, outputColumnName = "kME")
head(datKME)
```

Generate a complex heatmap of module-trait relationships.  

```{r}
#bold sig p-values
#dendrogram with WGCNA MEtree cut-off
#colored y-axis

#Create list of pvalues for eigengene correlation with specific life stages
heatmappval <- signif(moduleTraitPvalue, 1)

#Make list of heatmap row colors
htmap.colors <- names(MEs)
htmap.colors <- gsub("ME", "", htmap.colors)

library(dendsort)
row_dend = dendsort(hclust(dist(moduleTraitCor)))
col_dend = dendsort(hclust(dist(t(moduleTraitCor))))

pdf(file = "D-Apul/output/11.00-Apul-WGCNA/Module-trait-relationship-heatmap.pdf", height = 14, width = 8)
Heatmap(moduleTraitCor, name = "Eigengene", row_title = "Gene Module", column_title = "Module-Trait Eigengene Correlation", 
        col = blueWhiteRed(50), 
        row_names_side = "left", 
        #row_dend_side = "left",
        width = unit(5, "in"), 
        height = unit(8.5, "in"), 
        #column_dend_reorder = TRUE, 
        #cluster_columns = col_dend,
        row_dend_reorder = TRUE,
        #column_split = 6,
        row_split=3,
        #column_dend_height = unit(.5, "in"),
        #column_order = lifestage_order, 
        cluster_rows = row_dend, 
        row_gap = unit(2.5, "mm"), 
        border = TRUE,
        cell_fun = function(j, i, x, y, w, h, col) {
        if(heatmappval[i, j] < 0.05) {
            grid.text(sprintf("%s", heatmappval[i, j]), x, y, gp = gpar(fontsize = 10, fontface = "bold"))
        }
        else {
            grid.text(sprintf("%s", heatmappval[i, j]), x, y, gp = gpar(fontsize = 10, fontface = "plain"))
        }},
        column_names_gp =  gpar(fontsize = 12, border=FALSE),
        column_names_rot = 35,
        row_names_gp = gpar(fontsize = 12, alpha = 0.75, border = FALSE))
#draw(ht)
dev.off()

```

# Plot eigengene values 

View module eigengene data and make dataframe for Strader plots.  
```{r}
head(MEs)
names(MEs)
Eigen_MEs <- MEs
Eigen_MEs$AzentaSampleName <- rownames(Eigen_MEs)
head(Eigen_MEs)

plotTraits<-datTraits
plotTraits$AzentaSampleName <- rownames(plotTraits)

Eigen_MEs<-Eigen_MEs%>%
  droplevels() #drop unused level

dim(Eigen_MEs)
head(Eigen_MEs)
```

Plot mean module eigengene for each module.  

```{r}
#convert wide format to long format for plotting  
plot_MEs<-Eigen_MEs%>%
  gather(., key="Module", value="Mean", ME0:ME47)

#join with phys data 
plot_MEs<-left_join(plot_MEs, plotTraits)

expression_plots<-plot_MEs%>%
  group_by(Module) %>%
  
  ggplot(aes(x=Rd, y=Mean)) +
  facet_wrap(~ Module)+
  geom_point()+
  geom_smooth()+
  #ylim(-0.5,1) +
  geom_hline(yintercept = 0, linetype="dashed", color = "grey")+
  theme_bw() + 
  theme(axis.text.x=element_text(angle = 45, hjust=1, size = 12), #set x-axis label size
        axis.title.x=element_text(size = 14), #set x-axis title size
        axis.ticks.x=element_blank(), #No x-label ticks
        #axis.title.y=element_blank(), #No y-axis title
        axis.text.y=element_text(size = 14), #set y-axis label size, 
        panel.border = element_rect(color = "black", fill = NA, size = 1), #set border
        panel.grid.major = element_blank(), #Set major gridlines
        panel.grid.minor = element_blank(), #Set minor gridlines
        axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank(),
        plot.title = element_text(size=22)); expression_plots
```


