---
title: "12-Apul-miRNA-mRNA-WGCNA"
author: "Kathleen Durkin"
date: "2025-01-01"
output: 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    number_sections: true
bibliography: references.bib
---

Running Weighted Gene Correlation Network Analysis (WGCNA) to assess patterns of miRNA-mRNA coexpression in A.pulchra. 


```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Install and load packages

```{r load_libraries, inlcude = TRUE}
library(tidyverse)
library(ggplot2)
library(WGCNA)
library(magrittr)
library(genefilter)
library(DESeq2)
library(ggfortify)
library(RColorBrewer)
library(pheatmap)
library(factoextra)
library(vegan)
library(dendsort)
library(ComplexHeatmap)
```

# Load data

Load in count matrices for RNAseq. 

```{r load-data}
# raw gene counts data (will filter and variance stabilize)
Apul_genes <- read_csv("../output/02.20-D-Apul-RNAseq-alignment-HiSat2/apul-gene_count_matrix.csv")
Apul_genes <- as.data.frame(Apul_genes)

# format gene IDs as rownames (instead of a column)
rownames(Apul_genes) <- Apul_genes$gene_id
Apul_genes <- Apul_genes%>%select(!gene_id)


# load and format metadata
metadata <- read_csv("../../M-multi-species/data/rna_metadata.csv")%>%select(AzentaSampleName, ColonyID, Timepoint)%>%
  filter(grepl("ACR", ColonyID))
metadata$Sample <- paste(metadata$AzentaSampleName, metadata$ColonyID, metadata$Timepoint, sep = "_")

colonies <- unique(metadata$ColonyID)

# Load physiological data
phys<-read_csv("https://github.com/urol-e5/timeseries/raw/refs/heads/master/time_series_analysis/Output/master_timeseries.csv")%>%filter(colony_id_corr %in% colonies)%>%
  select(colony_id_corr, species, timepoint, site, Host_AFDW.mg.cm2, Sym_AFDW.mg.cm2, Am, AQY, Rd, Ik, Ic, calc.umol.cm2.hr, cells.mgAFDW, prot_mg.mgafdw, Ratio_AFDW.mg.cm2, Total_Chl, Total_Chl_cell, cre.umol.mgafdw)
# format timepoint
phys$timepoint <- gsub("timepoint", "TP", phys$timepoint)
#add column with full sample info
phys <- merge(phys, metadata, by.x = c("colony_id_corr", "timepoint"), by.y = c("ColonyID", "Timepoint")) %>%
  select(-AzentaSampleName)
  

#add site information into metadata 
metadata$Site<-phys$site[match(metadata$ColonyID, phys$colony_id_corr)]


# Rename gene column names to include full sample info (as in miRNA table)
colnames(Apul_genes) <- metadata$Sample[match(colnames(Apul_genes), metadata$AzentaSampleName)]

# raw miRNA counts (will filter and variance stabilize)
Apul_miRNA <- read.table(file = "../output/03.10-D-Apul-sRNAseq-expression-DESeq2/Apul_miRNA_ShortStack_counts_formatted.txt", header = TRUE, sep = "\t", check.names = FALSE)

```

Ensure there are no genes or miRNAs with 0 counts across all samples.

```{r}
nrow(Apul_genes)

Apul_genes<-Apul_genes %>%
     mutate(Total = rowSums(.[, 1:40]))%>%
    filter(!Total==0)%>%
    dplyr::select(!Total)

nrow(Apul_genes)

# miRNAs
nrow(Apul_miRNA)

Apul_miRNA<-Apul_miRNA %>%
     mutate(Total = rowSums(.[, 1:40]))%>%
    filter(!Total==0)%>%
    dplyr::select(!Total)

nrow(Apul_miRNA)
```

Removing genes with only 0 counts reduced number from 44371 to 35869. Retained all 51 miRNAs.

# Physiology filtering

Run PCA on physiology data to see if there are phys outliers 

Export data for PERMANOVA test.  
```{r}
test<-as.data.frame(phys)
test<-test[complete.cases(test), ]
```

Build PERMANOVA model.  
```{r}
scaled_test <-prcomp(test%>%select(where(is.numeric)), scale=TRUE, center=TRUE)
fviz_eig(scaled_test)

# scale data
vegan <- scale(test%>%select(where(is.numeric)))

# PerMANOVA 
permanova<-adonis2(vegan ~ timepoint*site, data = test, method='eu')
permanova
```

```{r}
pca1<-ggplot2::autoplot(scaled_test, data=test, frame.colour="timepoint", loadings=FALSE,  colour="timepoint", shape="site", loadings.label.colour="black", loadings.colour="black", loadings.label=FALSE, frame=FALSE, loadings.label.size=5, loadings.label.vjust=-1, size=5) + 
  geom_text(aes(x = PC1, y = PC2, label = paste(colony_id_corr, timepoint)), vjust = -0.5)+
  theme_classic()+
   theme(legend.text = element_text(size=18), 
         legend.position="right",
        plot.background = element_blank(),
        legend.title = element_text(size=18, face="bold"), 
        axis.text = element_text(size=18), 
        axis.title = element_text(size=18,  face="bold"));pca1
```

Remove ACR-173, timepoint 3 sample from analysis. This is Azenta sample 1B2. 

```{r}
Apul_genes <- Apul_genes%>%
  select(!`1B2_ACR-173_TP3`)

Apul_miRNA <- Apul_miRNA%>%
  select(!`1B2_ACR-173_TP3`)

metadata <- metadata %>%
  filter(Sample != "1B2_ACR-173_TP3")
```

We also do not have phys data for colony 1B9 ACR-265 at TP4, so I'll remove that here as well. 

```{r}
Apul_genes <- Apul_genes%>%
  select(!`1B9_ACR-265_TP4`)

Apul_miRNA <- Apul_miRNA%>%
  select(!`1B9_ACR-265_TP4`)

metadata <- metadata %>%
  filter(Sample != "1B9_ACR-265_TP4")
```



# pOverA filtering

*pOverA*: Specifying the minimum count for a proportion of samples for each gene. Here, we are using a pOverA of 0.1. This is because we have 40 samples with a minimum of n=4 samples per timepoint per site. Therefore, we will accept genes that are present in 4/40 = 0.1 of the samples because we expect different expression by life stage. We are further setting the minimum count of genes and miRNA to 10, such that 12.5% of the samples must have a gene count of >10 in order for the gene to remain in the data set.

Filter in the package "genefilter". Pre-filtering our dataset to reduce the memory size dataframe, increase the speed of the transformation and testing functions, and improve quality of statistical analysis by removing low-coverage counts. Removed counts could represent outliers in the data and removing these improves sensitivity of statistical tests.   

genes:
```{r}
filt <- filterfun(pOverA(0.1,10))

#create filter for the counts data
gfilt <- genefilter(Apul_genes, filt)

#identify genes to keep by count filter
gkeep <- Apul_genes[gfilt,]

#identify genes to keep by count filter
gkeep <- Apul_genes[gfilt,]

#identify gene lists
gn.keep <- rownames(gkeep)

#gene count data filtered in PoverA, P percent of the samples have counts over A
Apul_genes_filt <- as.data.frame(Apul_genes[which(rownames(Apul_genes) %in% gn.keep),])

#How many rows do we have before and after filtering?
nrow(Apul_genes) #Before
nrow(Apul_genes_filt) #After
```

We had 35869 genes before, and 23459 genes after filtering. 


miRNA:
```{r}
mifilt <- filterfun(pOverA(0.1,10))

#create filter for the counts data
mifilt <- genefilter(Apul_miRNA, mifilt)

#identify genes to keep by count filter
mikeep <- Apul_miRNA[mifilt,]

#identify genes to keep by count filter
mikeep <- Apul_miRNA[mifilt,]

#identify gene lists
mi.keep <- rownames(mikeep)

#gene count data filtered in PoverA, P percent of the samples have counts over A
Apul_miRNA_filt <- as.data.frame(Apul_miRNA[which(rownames(Apul_miRNA) %in% mi.keep),])

#How many rows do we have before and after filtering?
nrow(Apul_miRNA) #Before
nrow(Apul_miRNA_filt) #After
```

Of our original 51 miRNAs, 43 are retained.

# Combine counts data

```{r}
# combine gene counts and miRNA counts
# make sure column names in both are in the same order
Apul_miRNA_filt <- Apul_miRNA_filt[, colnames(Apul_genes_filt)]

# Combine by rows
Apul_g_mi_filt <- rbind(Apul_genes_filt, Apul_miRNA_filt)
```


# Assign metadata and arrange order of columns 

Display current order of metadata and gene count matrix.  
```{r}
metadata$Sample
colnames(Apul_g_mi_filt)
```

Order metadata the same as the column order in the gene matrix.  
```{r}
list<-colnames(Apul_g_mi_filt)
list<-as.factor(list)

metadata$Sample<-as.factor(metadata$Sample)

# Re-order the levels
metadata$Sample <- factor(as.character(metadata$Sample), levels=list)
# Re-order the data.frame
metadata_ordered <- metadata[order(metadata$Sample),]
metadata_ordered$Sample
```

Metadata and gene count matrix are now ordered the same. 

# Conduct variance stabilized transformation 

```{r}
#Set DESeq2 design
dds <- DESeqDataSetFromMatrix(countData = Apul_g_mi_filt,
                              colData = metadata_ordered,
                              design = ~Timepoint+ColonyID)
```

Check size factors. 
```{r}
SF.dds <- estimateSizeFactors(dds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.dds)) #View size factors

all(sizeFactors(SF.dds)) < 4
```

All size factors are less than 4, so we can use VST transformation.  

```{r}
vsd <- vst(dds, blind=FALSE) #apply a variance stabilizing transformation to minimize effects of small counts and normalize with respect to library size
head(assay(vsd), 3) #view transformed gene count data for the first three genes in the dataset.  
```

# PCA

Keep in mind this is of both the genes and miRNAs. since there are vastly more genes than miRNAs, the PCA will be dictated by genes. for miRNas alone, see `03.10`.
```{r}
plotPCA(vsd, intgroup = c("ColonyID"))
plotPCA(vsd, intgroup = c("Timepoint"))
```

# Sample clustering

```{r}
sampleDists <- dist(t(assay(vsd))) #calculate distance matrix
sampleDistMatrix <- as.matrix(sampleDists) #distance matrix
rownames(sampleDistMatrix) <- colnames(vsd) #assign row names
colnames(sampleDistMatrix) <- NULL #assign col names
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255) #assign colors

pht <- pheatmap(sampleDistMatrix, #plot matrix
         clustering_distance_rows=sampleDists, #cluster rows
         clustering_distance_cols=sampleDists, #cluster columns
         col=colors) #set colors

print(pht)
```

# Outlier checks

Look for outliers by examining tree of samples  
```{r}
datExpr <- as.data.frame(t(assay(vsd)))

sampleTree = hclust(dist(datExpr), method = "average");
plot(sampleTree, main = "Sample clustering to detect outliers: genes", sub="", xlab="", cex.lab = 1.5, cex.axis = 1.5, cex.main = 2)
```

Soft threshold selection. The soft thresholding power (β) is the number to which the co-expression similarity is raised to calculate adjacency. The function pickSoftThreshold performs a network topology analysis. The user chooses a set of candidate powers, however the default parameters are suitable values.  

```{r, message=FALSE, warning=FALSE}
allowWGCNAThreads()
# # Choose a set of soft-thresholding powers
powers <- c(seq(from = 1, to=19, by=2), c(21:30)) #Create a string of numbers from 1 through 10, and even numbers from 10 through 20
# 
# # Call the network topology analysis function
sft <- pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
```

Plot the results.  
```{r}
sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;
# # # Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
      xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
 text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# # # this line corresponds to using an R^2 cut-off
 abline(h=0.9,col="red")
# # # Mean connectivity as a function of the soft-thresholding power
 plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
 text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```
I used a scale-free topology fit index **R^2 of 0.9**. The lowest recommended R^2 by Langfelder and Horvath is 0.8. I chose 0.9 because we want to use the smallest soft thresholding power that maximizes with model fit. It appears that our **soft thresholding power is 5** because it is the lowest power above the R^2=0.9 threshold that maximizes with model fit. I will use a **signed network**.  


# WGCNA

Now we're ready to run WGCNA!

```{r, cache=TRUE}
picked_power = 5
temp_cor <- cor       
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
netwk_Apul <- blockwiseModules(datExpr,                # <= input here

                          # == Adjacency Function ==
                          power = picked_power,                # <= power here
                          networkType = "signed",

                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 5000,

                          # == Module Adjustments ==
                          reassignThreshold = 1e-6,
                          mergeCutHeight = 0.15,

                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = F,
                          saveTOMFileBase = "ER",

                          # == Output Options
                          numericLabels = T,
                          verbose = 3)


cor <- temp_cor     # Return cor function to original namespace
```

Take a look at dendrogram. 
```{r}
# Convert labels to colors for plotting
mergedColors = labels2colors(netwk_Apul$colors)
labels = table(netwk_Apul$colors)
labels

# Plot the dendrogram and the module colors underneath

plotDendroAndColors(
  netwk_Apul$dendrograms[[1]],
  mergedColors[netwk_Apul$blockGenes[[1]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )
```

Show module number before and after merging. 
```{r}
length(table(netwk_Apul$unmergedColors))
length(table(netwk_Apul$colors))
```

There are 54 modules before merging to 85% similarity and 52 after. 

```{r}
MEs<-netwk_Apul$MEs
moduleLabels<-netwk_Apul$colors
moduleColors <- labels2colors(netwk_Apul$colors)
```


```{r}
# Get Module Eigengenes per cluster
MEs0_Apul <- moduleEigengenes(datExpr, mergedColors)$eigengenes

# Add treatment names
MEs0_Apul$sample = row.names(MEs0_Apul)

# Join metadata to add timepoint to mME_Apul
mME_Apul <- mME_Apul %>%
  left_join(metadata, by = c("sample" = "Sample"))

# Order samples by timepoint
timepoint_order <- c("TP1", "TP2", "TP3", "TP4") # Specify the desired order of timepoints
mME_Apul <- mME_Apul %>%
  mutate(sample = factor(sample, levels = metadata$Sample[order(match(metadata$Timepoint, timepoint_order))]))

# Plot heatmap
mME_Apul %>% ggplot(., aes(x = sample, y = name, fill = value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1,1)
  ) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill = "corr")
```






# Correlate to traits 

Ensure we've removed sample from expression data that is not present in physiological data. 1B9
```{r}

phys<-phys%>%
  filter(!Sample=="1B2_ACR-173_TP3")%>%
  filter(!Sample=="1B9_ACR-265_TP4")

setdiff(metadata_ordered$Sample, phys$Sample)

#all have 38 observations 
```

Create a physiological/factor dataset with sample in rows and phys data in columns (row name = Azenta sample name)
```{r}
# phys$AzentaSampleName<-metadata_ordered$AzentaSampleName[match(phys$Sample, metadata_ordered$Sample)]

traits<-phys%>%
  select(Sample, timepoint, site, Host_AFDW.mg.cm2, Sym_AFDW.mg.cm2, Am, Rd, Ik, Ic, AQY, cells.mgAFDW, Total_Chl, Ratio_AFDW.mg.cm2, Total_Chl_cell)

#time point
#site
traits<-traits%>%
  mutate(site=gsub("Mahana", "1", site)) %>%
  mutate(site=gsub("Manava", "2", site))%>%
  mutate(timepoint=gsub("TP", "", timepoint))

traits$timepoint<-as.numeric(traits$timepoint)
traits$site<-as.numeric(traits$site)

traits<-as.data.frame(traits)
rownames(traits)<-traits$Sample

datTraits<-traits%>%select(!Sample)

#datTraits<-datTraits%>%select(!Timepoint)
```




From this tutorial: https://alexslemonade.github.io/refinebio-examples/04-advanced-topics/network-analysis_rnaseq_01_wgcna.html#410_Which_modules_have_biggest_differences_across_treatment_groups

```{r}
# Create the design matrix from the `time_point` variable
des_mat <- model.matrix(~ datTraits$site)

# lmFit() needs a transposed version of the matrix
fit <- limma::lmFit(t(MEs), design = des_mat)

# Apply empirical Bayes to smooth standard errors
fit <- limma::eBayes(fit)

stats_df <- limma::topTable(fit, number = ncol(MEs)) %>%
  tibble::rownames_to_column("module")

head(stats_df)
```

















Add in temperature and light information for each time point from physiology manuscript. 
```{r}
# env<-read_csv("https://github.com/urol-e5/timeseries/raw/refs/heads/master/time_series_analysis/Output/environment_characteristics_RDA.csv")%>%select(mean_Temp_mean, mean_solar_rad_kwpm2_mean, timepoint)%>%
#   mutate(timepoint=if_else(timepoint=="timepoint1", "1", 
#                            if_else(timepoint=="timepoint2", "2", 
#                                    if_else(timepoint=="timepont3", "3", "4"))))%>%
#   mutate(timepoint=as.numeric(timepoint))%>%
#   select(!timepoint)
# 
# #merge into datTraits
# 
# datTraits<-left_join(datTraits, env)
# rownames(datTraits)<-traits$Sample
# 
# str(datTraits)
# 
# datTraits<-datTraits%>%select(!timepoint)
```

```{r}
nGenes = ncol(datExpr)
nSamples = nrow(datExpr)

nGenes
nSamples
```

Generate labels for module eigengenes as numbers.  
```{r}
MEs0 = moduleEigengenes(datExpr, moduleLabels, softPower=5)$eigengenes
MEs = orderMEs(MEs0)
names(MEs)
```

Correlations of traits with eigengenes
```{r}
moduleTraitCor = cor(MEs, datTraits, use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);
Colors=sub("ME","", names(MEs))

moduleTraitTree = hclust(dist(t(moduleTraitCor)), method = "average")

# pdf(file="D-Apul/output/11.00-Apul-WGCNA/ModuleTraitClusterTree.pdf")
plot(moduleTraitTree)
# dev.off()
```

Correlations of genes with eigengenes. Calculate correlations between ME's and lifestages. 
```{r}
moduleGeneCor=cor(MEs,datExpr)
moduleGenePvalue = corPvalueStudent(moduleGeneCor, nSamples);
```

Calculate kME values (module membership). 

```{r}
datKME = signedKME(datExpr, MEs, outputColumnName = "kME")
head(datKME)
```

Generate a complex heatmap of module-trait relationships.  

```{r}
#bold sig p-values
#dendrogram with WGCNA MEtree cut-off
#colored y-axis

#Create list of pvalues for eigengene correlation with specific life stages
heatmappval <- signif(moduleTraitPvalue, 1)

#Make list of heatmap row colors
htmap.colors <- names(MEs)
htmap.colors <- gsub("ME", "", htmap.colors)

row_dend = dendsort(hclust(dist(moduleTraitCor)))
col_dend = dendsort(hclust(dist(t(moduleTraitCor))))

pdf(file = "../output/12-Apul-miRNA-mRNA-WGCNA/Module-trait-relationship-heatmap.pdf", height = 14, width = 8)
Heatmap(moduleTraitCor, name = "Eigengene", row_title = "Gene Module", column_title = "Module-Trait Eigengene Correlation", 
        col = blueWhiteRed(50), 
        row_names_side = "left", 
        #row_dend_side = "left",
        width = unit(5, "in"), 
        height = unit(8.5, "in"), 
        #column_dend_reorder = TRUE, 
        #cluster_columns = col_dend,
        row_dend_reorder = TRUE,
        #column_split = 6,
        row_split=3,
        #column_dend_height = unit(.5, "in"),
        #column_order = lifestage_order, 
        cluster_rows = row_dend, 
        row_gap = unit(2.5, "mm"), 
        border = TRUE,
        cell_fun = function(j, i, x, y, w, h, col) {
        if(heatmappval[i, j] < 0.05) {
            grid.text(sprintf("%s", heatmappval[i, j]), x, y, gp = gpar(fontsize = 10, fontface = "bold"))
        }
        else {
            grid.text(sprintf("%s", heatmappval[i, j]), x, y, gp = gpar(fontsize = 10, fontface = "plain"))
        }},
        column_names_gp =  gpar(fontsize = 12, border=FALSE),
        column_names_rot = 35,
        row_names_gp = gpar(fontsize = 12, alpha = 0.75, border = FALSE))
#draw(ht)
#dev.off()

```

This plot is cool, but I'm a little skeptical of the results. No modules show significant relationship with timepoint, despite the PCAs showing a pretty clear clustering by timepoint...

# Plot eigengene values 

View module eigengene data and make dataframe for Strader plots.  
```{r}
head(MEs)
names(MEs)
Eigen_MEs <- MEs
Eigen_MEs$AzentaSampleName <- rownames(Eigen_MEs)
head(Eigen_MEs)

plotTraits<-datTraits
plotTraits$AzentaSampleName <- rownames(plotTraits)

Eigen_MEs<-Eigen_MEs%>%
  droplevels() #drop unused level

dim(Eigen_MEs)
head(Eigen_MEs)
```

Plot mean module eigengene for each module.  

```{r}
#convert wide format to long format for plotting  
str(Eigen_MEs)

plot_MEs <- Eigen_MEs %>%
  pivot_longer(
    cols = where(is.numeric),  # Select only numeric columns
    names_to = "Module",       # Name for the new column containing the column names
    values_to = "Mean"         # Name for the new column containing the values
  )

str(plot_MEs)

#join with phys data 
plot_MEs<-left_join(plot_MEs, plotTraits)

expression_plots<-plot_MEs%>%
  group_by(Module) %>%
  
  ggplot(aes(x=mean_Temp_mean, y=Mean)) +
  facet_wrap(~ Module)+
  geom_point()+
  geom_smooth(method="lm")+
  #ylim(-0.5,1) +
  geom_hline(yintercept = 0, linetype="dashed", color = "grey")+
  theme_bw() + 
  theme(axis.text.x=element_text(angle = 45, hjust=1, size = 12), #set x-axis label size
        axis.title.x=element_text(size = 14), #set x-axis title size
        axis.ticks.x=element_blank(), #No x-label ticks
        #axis.title.y=element_blank(), #No y-axis title
        axis.text.y=element_text(size = 14), #set y-axis label size, 
        panel.border = element_rect(color = "black", fill = NA, size = 1), #set border
        panel.grid.major = element_blank(), #Set major gridlines
        panel.grid.minor = element_blank(), #Set minor gridlines
        axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank(),
        plot.title = element_text(size=22)); expression_plots
```





# Export to Cytoscape

Export modules of interest for network visualization

```{r}
mods_interest <- c("26", "9", "2", "50", "27", 
                   "10", "5", "38", "16", "14", 
                   "11", "6", "1", "49", "8", 
                   "31", "10", "39", "41", "51", 
                   "29", "11", "45", "21")

c("26", "27", "29", "31", "38", "39", "41", "45", "49", "50", "51")
```

```{r}
# Recalculate topological overlap if needed
TOM = TOMsimilarityFromExpr(datExpr, power = 5);
# Read in the annotation file
annot = datTraits;
# Select modules
modules = c("50");
# Select module probes
probes = names(datExpr)
inModule = is.finite(match(moduleLabels, modules));
modProbes = probes[inModule];
# modGenes = annot$gene_symbol[match(modProbes, annot$substanceBXH)];
# Select the corresponding Topological Overlap
modTOM = TOM[inModule, inModule];
dimnames(modTOM) = list(modProbes, modProbes)
# Export the network into edge and node list files Cytoscape can read
cyt = exportNetworkToCytoscape(modTOM,
  edgeFile = paste("CytoscapeInput-edges-", paste(modules, collapse="-"), ".txt", sep=""),
  nodeFile = paste("CytoscapeInput-nodes-", paste(modules, collapse="-"), ".txt", sep=""),
  weighted = TRUE,
  threshold = 0.02,
  nodeNames = modProbes,
  # altNodeNames = modGenes,
  nodeAttr = moduleLabels[inModule]);



```

```{r, engine='bash'}


```




```{r}
# Print session info
sessioninfo::session_info()
```




