---
title: "21-Apul-annotate-miRNA-mRNA-WGCNA"
author: "Kathleen Durkin"
date: "2025-03-19"
output: 
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
---


After running WGCNA to evaluate modules of mRNA and miRNA with correlated expression, we need to functionally annotate genes to evaluate module function(s).

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```


# Load packages

```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(topGO)
```

# Load and format annotation files

Already annotated the A.pulchra genome as part of the `deep-dive expression` project (see [`deep-dive-expression/D-Apul/code/02-Apul-reference-annotation.Rmd`](https://github.com/urol-e5/deep-dive-expression/blob/main/D-Apul/code/02-Apul-reference-annotation.Rmd)), so we'll pull the annotation file from that

```{r, engine='bash'}
# Can access file stored here if needed
# https://gannet.fish.washington.edu/kdurkin1/ravenbackups/deep-dive-expression/D-Apul/output/02-Apul-reference-annotation/Apulcra-genome-mRNA-IDmapping-2024_12_12.tab
cp ../../../deep-dive-expression/D-Apul/output/02-Apul-reference-annotation/Apulcra-genome-mRNA-IDmapping-2024_12_12.tab ../output/21-Apul-annotate-miRNA-mRNA-WGCNA
```

Read mapping file into R
```{r}
IDmapping_dirty <- read.table("../output/21-Apul-annotate-miRNA-mRNA-WGCNA/Apulcra-genome-mRNA-IDmapping-2024_12_12.tab", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# Need to remove quotations surrounding each entry
IDmapping_locations <- IDmapping_dirty
IDmapping_locations[] <- lapply(IDmapping_locations, function(x) gsub('^"(.*)"$', '\\1', x))
# Remove unneeded columns 
IDmapping_locations <- IDmapping_locations %>% dplyr::select(-X, -V13)
# Ensure there are no duplicate rows
IDmapping_locations <- IDmapping_locations %>% distinct()

head(IDmapping_locations)
# Looks good!
```

This file shows each gene as it's genomic location. We want to use gene IDs to associate genes, so add gene IDs to this annotation table

Read in file that associates each mRNA genomic location with corresponding gene ID
```{r}
mRNA_FUNids <- read.table("../output/05-Apul-annotate-UTRs/Apul-mRNA-FUNids.txt", header=FALSE, col.names=c("location", "type", "mRNA_ID", "gene_ID", "product"), sep="\t")

# Remove unwanted text from parent column
mRNA_FUNids$gene_ID <- gsub("Parent=", "", mRNA_FUNids$gene_ID)
# Only need to keep mRNA location and gene ID
mRNA_FUNids <- mRNA_FUNids %>% dplyr::select(location, gene_ID)
```

join with annotation file
```{r}
IDmapping <- left_join(IDmapping_locations, mRNA_FUNids, by = c("V1" = "location"))
```


# Annotate WGCNA modules

Load file that shows which module each gene/miRNA is in
```{r}
module_membership <- read.table("../output/12-Apul-miRNA-mRNA-WGCNA/WGCNA-module-membership.tab", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```

join with annotation file
```{r}
module_FA <- left_join(module_membership, IDmapping, by = c("gene" = "gene_ID"))
```

Many of the genes don't have available annotations, so reduce to only keep those that are annotated (note this will also remove lncRNA and miRNA).
```{r}
module_FA_available <- module_FA[!is.na(module_FA$Gene.Ontology.IDs),]
module_FA_available <- module_FA_available[module_FA_available$Gene.Ontology.IDs != "",]
```


# Functional Enrichment of modules (topGO)

I want to see whether our modules of coexpressed genes/miRNA represent specific, over-represented functionalities (i.e. does any module contain more metabolism-related genes than expected?)

To do this we'll perform functional enrichment of each module using the R package topGO

## Format Gene Ontology (GO) annotations

Want to isolate a list of GO terms per gene
```{r}
gene_to_GO <- IDmapping %>% filter(!is.na(Gene.Ontology.IDs)) %>% dplyr::select(gene_ID, Gene.Ontology.IDs)
#head(gene_to_GO)

# Needs to be formatted as a named list for use in topGO
# Convert data frame to a named list
# gene_to_GO_list <- gene_to_GO %>%
#   mutate(GO_terms = strsplit(Gene.Ontology.IDs, ";")) %>%
#   tibble::deframe()

gene_to_GO_list <- setNames(
  strsplit(as.character(gene_to_GO$Gene.Ontology.IDs), ";"), 
  gene_to_GO$gene_ID
)
```

Note: I think this means genes that had a Uniprot ID but no GO terms are excluded from this analysis

Define reference set. This should be all genes *found in our samples*, NOT all genes in the A.pulchra genome. Some genes (e.g., reproduction pathways) may not be found/expected in our samples for valid biological reasons.
```{r}
reference_genes <- module_membership$gene
length(reference_genes)
```

## Extract list of significant WGCNA modules

```{r}
WGCNA_pvals <- read.delim("../output/12-Apul-miRNA-mRNA-WGCNA/pval-cor-WGCNA_module-phys_envir.tab", header=TRUE, sep="\t")

# filter to only keep rows and columns with at least one significant pval
filtered_rows <- WGCNA_pvals[rowSums(WGCNA_pvals < 0.05, na.rm = TRUE) > 0, ]
WGCNA_pvals_significant <- filtered_rows[, colSums(filtered_rows < 0.05, na.rm = TRUE) > 0]

# Create list of modules that are significantly correlated with at least one trait
significant_modules <- row.names(WGCNA_pvals_significant)
```


## topGO function

Create a function to run topGO functional enrichment analysis for an input module (we'll want to use this for many modules of interest)
```{r}
module_topGO_FE <- function(module.name) {
  
  #Isolate genes in our input module of interest
  genes_in_module <- module_membership %>%
    filter(module == module.name) %>%
    pull(gene)
  
  # Create factor for all reference genes, where 1 represents module membership and 0 means the gene is not in module of interest
  gene_list <- factor(as.integer(reference_genes %in% genes_in_module))
  names(gene_list) <- reference_genes
  
  # Create topGO object
  GO_BP <- new("topGOdata",
              description = "Functional Enrichment Analysis", 
              ontology = "BP", # Biological Process
              allGenes = gene_list,
              annot = annFUN.gene2GO, 
              gene2GO = gene_to_GO_list)
  
  # Run GO enrichment test
  GO_BP_FE <- runTest(GO_BP, algorithm = "weight01", statistic = "fisher")
  
  # View the results
  GO_BP_results <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
  
  # Filter by significant results
  GO_BP_results$Fisher<-as.numeric(GO_BP_results$Fisher)
  GO_BP_results_sig<-GO_BP_results[GO_BP_results$Fisher<0.05,]

  # Return
  print(GO_BP_results_sig)
}
```

```{r}
results <- list()

for(module in significant_modules) {
  # Run topGo enrichent function
  module_results <- module_topGO_FE(module)
  
  # Only keep results if not empty
  if (nrow(module_results) > 0) {
    # note the source column
    module_results$module <- module
    # append to results list
    results[[module]] <- module_results
  }
}

# Combine all the resuts data frames into one
combined_GO_BP_results_sig <- do.call(rbind, results)

# View
print(combined_GO_BP_results_sig)

```


## Heatmap of significant GO terms by trait

```{r}

# Extract significant modules for each trait
significant_modules_per_trait <- apply(WGCNA_pvals_significant, 2, function(x) rownames(WGCNA_pvals_significant)[x < 0.05])

# Create empty matrix to store values for hetmap
go_term_counts <- matrix(0, nrow = ncol(WGCNA_pvals_significant), ncol = length(unique(combined_GO_BP_results_sig$Term)))
rownames(go_term_counts) <- colnames(WGCNA_pvals_significant)  # Trait names
colnames(go_term_counts) <- unique(combined_GO_BP_results_sig$Term)  # Unique GO terms

# Loop through each trait, calculate the number of significant modules enriched for each GO term
for (trait in colnames(WGCNA_pvals_significant)) {
  # Get the significant modules for this trait
  significant_modules <- significant_modules_per_trait[[trait]]
  
  # Check if a module is enriched for each GO term
  for (go_term in unique(combined_GO_BP_results_sig$Term)) {
    # Check if the module is in the list of significant modules for the trait and if it's enriched for this GO term
    enriched_modules <- combined_GO_BP_results_sig %>%
      filter(Term == go_term & module %in% significant_modules) %>%
      nrow()  # Count of modules enriched for this GO term
    
    # Count number of modules that contain each GO term
    go_term_counts[trait, go_term] <- enriched_modules
  }
}

# Convert matrix to df for plotting
go_term_counts_df <- as.data.frame(go_term_counts) %>%
  rownames_to_column("Trait") %>%
  gather(key = "GO_term", value = "Module_Count", -Trait)

# Truncate GO term names
#go_term_counts_df$GO_term <- substr(go_term_counts_df$GO_term, 1, 40)

# Plot
ggplot(go_term_counts_df, aes(x = Trait, y = GO_term, fill = Module_Count)) +
  geom_tile(color = "grey") +  # Adds grid lines
  scale_fill_gradient(low = "white", high = "blue") +  # Color gradient from white (low) to blue (high)
  theme_minimal() +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust=0.5, vjust=0.5), 
        axis.text.y = element_text(size = 8),
        ) + 
  labs(fill = "Proportion Enriched", title = "Proportion of Significant Modules Enriched for GO Terms") +
  theme(legend.position = "bottom")
ggsave("../output/21-Apul-annotate-miRNA-mRNA-WGCNA/go_term_heatmap.png", width = 12, height = 8, dpi = 300)
```


# Annotate putative binding

Outline: 

1. Have table module pair and correlation values

## Load module-module correlations

```{r}
modules_cor <- read.delim("../output/12-Apul-miRNA-mRNA-WGCNA/WGCNA-modules-correlations.tab", header=TRUE, sep="\t")
# Remove instances where a module is just correlating to itself
modules_cor <- modules_cor[modules_cor$module_A != modules_cor$module_B,]
```

## Annotate whether a given module contains miRNA

```{r}
# Separate module membership into only mRNA and only miRNA
module_membership_mRNA <- module_membership %>% filter(grepl("FUN", gene))
module_membership_miRNA <- module_membership %>% filter(grepl("Cluster", gene))

# Expand each miRNA_module to show all contained miRNAa
modules_cor_miRNA <- left_join(modules_cor, module_membership_miRNA, by = c("module_A" = "module"))
# Rename columns appropriately
colnames(modules_cor_miRNA) <- c("miRNA_module", "mRNA_module", "correlation", "miRNA")

# Remove instances where the module_A (miRNA_module) doesn't contain miRNA
modules_cor_miRNA <- modules_cor_miRNA %>% filter(!is.na(miRNA))

# Expand each mRNA module to show all contained genes
modules_cor_miRNA_mRNA <- left_join(modules_cor_miRNA, module_membership_mRNA, by = c("mRNA_module" = "module"))


```

## ID which miRNA-mRNA pairs putatively bind

Load and format miRanda tables

```{r}
## Load ##
miRanda_3UTR <- read.delim("../output/07-Apul-miRNA-mRNA-miRanda/Apul-miRanda-3UTR-strict-parsed-geneIDs.txt", header=FALSE, sep="\t")

miRanda_5UTR <- read.delim("../output/07.1-Apul-miRNA-mRNA-miRanda-additional_inputs/Apul-miRanda-5UTR_1kb-strict-parsed.txt", header=FALSE, sep="\t")
  
miRanda_mRNA <- read.delim("../output/07.1-Apul-miRNA-mRNA-miRanda-additional_inputs/Apul-miRanda-mRNA_full-strict-parsed.txt", header=FALSE, sep="\t")

## Format ##
# Remove superfluous text in miRNA name column
miRanda_3UTR <- miRanda_3UTR %>% mutate(V1 = str_extract(V1, "Cluster[^.]+"))
miRanda_5UTR <- miRanda_5UTR %>% mutate(V1 = str_extract(V1, "Cluster[^.]+"))
miRanda_mRNA <- miRanda_mRNA %>% mutate(V1 = str_extract(V1, "Cluster[^.]+"))

# For 5UTR and mRNA, need to add gene_IDs associated with each genomic location
# Load and format table of 5UTR geneIDs
FUNids_5UTR <- read.delim("../output/05-Apul-annotate-UTRs/Apul-5UTR-FUNids.txt", header=FALSE, sep="\t")
FUNids_5UTR$V4 <- gsub("Parent=", "", FUNids_5UTR$V4)
FUNids_5UTR <- FUNids_5UTR %>% dplyr::select(V1, V4)
# Already have table of mRNA geneIDs

# Add gene IDs to the miRanda tables
miRanda_5UTR <- left_join(miRanda_5UTR, FUNids_5UTR, by=c("V2" = "V1"))
miRanda_mRNA <- left_join(miRanda_mRNA, mRNA_FUNids, by=c("V2"="location"))

# Ensure no unwanted duplicate rows
miRanda_3UTR <- miRanda_3UTR %>% distinct()
miRanda_5UTR <- miRanda_5UTR %>% distinct()
miRanda_mRNA <- miRanda_mRNA %>% distinct()

# Add column to identify whether putative binding is in 3UTR, 5UTR, or CDS
miRanda_3UTR$binding_loc <- "3UTR"
miRanda_5UTR$binding_loc <- "5UTR"
miRanda_mRNA$binding_loc <- "mRNA"

# Combine into single data frame
colnames(miRanda_3UTR) <- c("V1", "V2", "V3", "V4", "V5", "V6", "V7", "V8", "V9", "V10", "binding_loc")
colnames(miRanda_5UTR) <- c("V1", "V2", "V3", "V4", "V5", "V6", "V7", "V8", "V9", "V10", "binding_loc")
colnames(miRanda_mRNA) <- c("V1", "V2", "V3", "V4", "V5", "V6", "V7", "V8", "V9", "V10", "binding_loc")
miRanda_combined <- rbind(miRanda_3UTR, miRanda_5UTR, miRanda_mRNA)

head(miRanda_combined)
```

Join with table of correlating modules and their members to annotate, for each miRNA, which of the mRNAs in its correlated module it may bind to
```{r}
# Join 
modules_cor_miRNA_mRNA_bind <- left_join(modules_cor_miRNA_mRNA, miRanda_combined, by=c("gene" = "V10"))
# Only keep rows where the full miRNA-mRNA interaction matches, not just the mRNA
modules_cor_miRNA_mRNA_bind <- modules_cor_miRNA_mRNA_bind[modules_cor_miRNA_mRNA_bind$miRNA == modules_cor_miRNA_mRNA_bind$V1,] %>% distinct()
```

This is now a table that, for each module that contians miRNA (miRNA_module), shows each module it's correlated with (mRNA_module), the genes in that module, and whether that gene putatively binds with the miRNA. It still includes, though, instances of multiple putative binding positions for the same miRNA-mRNA pair. Since we just care about the high-level miRNA-mRNA pair, let's remove this level of detail

```{r}
mod_cor_bind <- modules_cor_miRNA_mRNA_bind %>% dplyr::select(miRNA_module, mRNA_module, correlation, miRNA, gene, binding_loc) %>% distinct()

nrow(mod_cor_bind)
nrow(mod_cor_bind[mod_cor_bind$correlation < 0,])
```

So there are 41,633 instances of an miRNA putatively binding to a gene that is contained in a correlated module. 21,274 of these (roughly half) are putative binding between modules that are *negatively* correlated with each other, which is the expected behavior for canonical miRNA function.

## Join with FA table to annotate each binding mRNA

```{r}
mod_cor_bind_FA <- left_join(mod_cor_bind, IDmapping, by=c("gene" = "gene_ID")) %>% distinct()

nrow(mod_cor_bind_FA %>% filter(!is.na(Gene.Ontology.IDs)))
nrow(mod_cor_bind_FA %>% filter(!is.na(Gene.Ontology.IDs) & correlation < 0))
```
Of these 41,633 instances of an miRNA putatively binding to a gene contained in a correlated module, only 12,367 (roughly a quarter) of the putatively bound genes are annotated with GO terms. Roughly half of the annotated genes are associated with negative module correlation.


# Functional Enrichment of putative binding

In the above annotation we've functionally annotated all miRNA-mRNA pairs that are in correlated modules *and* putatively bind in the 3UTR, 5UTR and/or coding sequence (based on miRanda). The ultimate goal is to elucidate putative function of our miRNAs by evaluating the functions of the mRNAs they putatively interact with. 

To aid this we can run functional enrichment on the mRNAs that each miRNA putatively interacts with.

```{r}
# Modify topGO function for use with miRNA names

miRNA_topGO_FE <- function(miRNA.name) {
  
  #Isolate genes in our input module of interest
  interacting_genes <- mod_cor_bind_FA %>%
    filter(miRNA == miRNA.name) %>%
    pull(gene)
  if (length(interacting_genes) > 0) {
    # Create factor for all reference genes, where 1 represents module membership and 0 means the gene is not in module of interest
    gene_list <- factor(as.integer(reference_genes %in% interacting_genes))
    names(gene_list) <- reference_genes
    
    str(gene_list)
    
    # Create topGO object
    GO_BP <- new("topGOdata",
                description = "Functional Enrichment Analysis", 
                ontology = "BP", # Biological Process
                allGenes = gene_list,
                annot = annFUN.gene2GO, 
                gene2GO = gene_to_GO_list)
    
    # Run GO enrichment test
    GO_BP_FE <- runTest(GO_BP, algorithm = "weight01", statistic = "fisher")
    
    # View the results
    GO_BP_results <- GenTable(GO_BP, Fisher = GO_BP_FE, orderBy = "Fisher",  topNodes = 100, numChar = 51)
    
    # Filter by significant results
    GO_BP_results$Fisher<-as.numeric(GO_BP_results$Fisher)
    GO_BP_results_sig<-GO_BP_results[GO_BP_results$Fisher<0.05,]
  
    # Return
    print(GO_BP_results_sig)
  }
}
```

```{r}
interacting_miRNAs <- unique(mod_cor_bind_FA$miRNA) %>% na.omit
results <- list()

for(miRNA in interacting_miRNAs) {
  # Run topGo enrichment function
  miRNA_results <- miRNA_topGO_FE(miRNA)
  
  print(miRNA_results)
  
  # Only keep results if not empty
  if (nrow(miRNA_results) > 0) {
    # note the source column
    miRNA_results$miRNA <- miRNA
    # append to results list
    results[[miRNA]] <- miRNA_results
  }
}

# Combine all the resuts data frames into one
combined_GO_BP_results_miRNA <- do.call(rbind, results)

# View
print(combined_GO_BP_results_miRNA)

```


