---
title: "00.21-F-Ptua-BS-genome"
author: "Sam White"
date: "2025-01-02"
output: 
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
---

# Background

This Rmd file will create a bisulfite-converted genome by, and for, Bismark [@krueger2011] using the `Pocillopora_meandrina_HIv1.assembly.fasta` file. The genome FastA was taken from the [Cyanophora webpage](http://cyanophora.rutgers.edu/Pocillopora_meandrina/).

Due to large sizes of output files, the files cannot be sync'd to GitHub. As such, the output directories will be gzipped and available here:

- [https://gannet.fish.washington.edu/gitrepos/urol-e5/timeseries_molecular/F-Ptua/data/Bisulfite_Genome.tar.gz](https://gannet.fish.washington.edu/gitrepos/urol-e5/timeseries_molecular/F-Ptua/data/Bisulfite_Genome.tar.gz) (972MB)

- [https://gannet.fish.washington.edu/gitrepos/urol-e5/timeseries_molecular/F-Ptua/data/Bisulfite_Genome.tar.gz.md5](https://gannet.fish.washington.edu/gitrepos/urol-e5/timeseries_molecular/F-Ptua/data/Bisulfite_Genome.tar.gz.md5)

  - MD5: `654a93dacf6a5a72475ac644599f8f8e`

# Inputs

- Directory containing a FastA file with the file extension: .fa or .fasta (also ending in .gz).

# Outputs

- CT Conversion

  - Bowtie2 index files.
  - CT conversion FastA
  
- GA conversion

  - Bowtie2 index files.
  - GA conversion FastA.

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Create a Bash variables file

This allows usage of Bash variables across R Markdown chunks.

```{r save-bash-variables-to-rvars-file, engine='bash', eval=TRUE}
{
echo "#### Assign Variables ####"
echo ""

echo "# Data directories"
echo 'export timeseries_dir=/home/shared/8TB_HDD_01/sam/gitrepos/urol-e5/timeseries_molecular'
echo 'export output_dir_top=${timeseries_dir}/F-Ptua/data'
echo 'export genome_dir=${timeseries_dir}/F-Ptua/data'
echo ""

echo "# Paths to programs"
echo 'export programs_dir="/home/shared"'
echo 'export bismark_dir="${programs_dir}/Bismark-0.24.0"'
echo 'export bowtie2_dir="${programs_dir}/bowtie2-2.4.4-linux-x86_64"'
echo ""

echo "# Set number of CPUs to use"
echo 'export threads=20'
echo ""

echo "# Print formatting"
echo 'export line="--------------------------------------------------------"'
echo ""
} > .bashvars

cat .bashvars
```

# Bisfulite conversion

```{r bismark-genome-conversion, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

${bismark_dir}/bismark_genome_preparation \
${genome_dir} \
--parallel ${threads} \
--bowtie2 \
--path_to_aligner ${bowtie2_dir} \
1> ${genome_dir}/Peve-bs-genome.stderr
```

## Inpect BS output
```{r inspect-BS-output, engine='bash', eval=TRUE}
# Load bash variables into memory
source .bashvars

tree -h ${genome_dir}/Bisulfite_Genome
```

## Compress output folder
```{r compress-BS-directory, engine='bash', eval=TRUE}
source .bashvars

tar -czvf ${genome_dir}/Bisulfite_Genome.tar.gz ${genome_dir}/Bisulfite_Genome
```

## Create MD5sum
```{r md5sum, engine='bash', eval=TRUE}
source .bashvars

cd ${genome_dir}

md5sum Bisulfite_Genome.tar.gz | tee Bisulfite_Genome.tar.gz.md5
```


# REFERENCES

